### 2024-02-02 Part1/part1.qmd
Rachel added Part 1 File with the manuscript template
Rachel is having trouble getting the summary and str functions to work on the cvs file.

---
title: "Robertson and Cao Project"
subtitle: "Teenage HPV Vaccination Coverage and Socioeconomic Factors"
author: Kelly Cao and Rachel Robertson
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../assets/dataanalysis-references.bib
csl: ../assets/apa.csl
---

The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. 
You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

# Summary/Abstract
_Write a summary of your project._

{{< pagebreak >}}

# Introduction

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_

## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._

## Questions/Hypotheses to be addressed
_State the research questions you plan to answer with this analysis._

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

{{< pagebreak >}}

# Methods

_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it). @fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
We store those figures in the `assets` folder.

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
knitr::include_graphics(here("products","assets","antigen-recognition.png"))
```

## Data aquisition
_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._

## Statistical analysis
_Explain anything related to your statistical analyses._

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","height-weight-stratified.png"))
```

## Full analysis

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables-files","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.

# *Part 1*
## Data Set
  The data we chose to use is the National Immunization Survey (NIS) of 2022. It consists survey data that was collected to monitor vaccination coverage for teenagers age 13-17 in the United States of America. The surveys were conducted by the National Center for Immunization and Respiratory Diseases of the Center (NCIRD) for Disease Control and Prevention (CDC). The survey itself consists of two parts: (1) the household telephone (random digit dialing) survey answered by a parent or guardian and (2) a mailed survey for the vaccination provider, called “Immunization History Questionnaire.”
  The original data set is a fixed width file (FWF). In the data set, there are 43,738 rows and 672 columns, with each row representing an individual and the each column representing the answer to a question. The questions pertain to the teenager’s immunization history, demographics, and additional household-reported health information. 
We have truncated the data set to include 26-selected columns for the ease of observation. We have also filtered the data set to include one year, 2022, which is the most recent year of data that is published. The provided survey data will require cleaning, as there are signs of human errors present. 
## Opening the Data Set

First, we will load the readr package to open the file
```{r}
library(readr) #Loading the package to read the fwf
```

Next, we will define the column widths (beginning and end positions) and applying the column names to each position.
```{r}
column_position<- fwf_positions(
  start = c(1, 89, 287, 295, 298, 314, 330, 331, 334, 336, 343, 344, 345,
            347, 355, 359, 362, 472, 473, 495, 496, 497, 502, 504, 1296, 1297
  ),
  end = c(5, 92, 288, 295, 313, 329, 330, 332, 334, 336, 343, 344, 346,
          347, 356, 359, 362, 472, 473, 495, 496, 497, 502, 504, 1296, 1297
  ),
  col_names = c("SEQNUMT", "YEAR", "AGE", "EDUC1", "INCPORAR", "INCPORAR_I", 
                "INCPOV1", "INCQ298A", "LANGUAGE", "MOBIL_1", "RACEETHK", 
                "RACE_K", "RENT_OWN", "SEX", "STATE", "FACILITY", "WELLCHILD", 
                "P_U13HPV", "P_U13HPV3", "P_UTDHPV", "P_UTDHPV_15", 
                "P_UTDHPV_15INT", "P_UTDHPV2", "P_UTDHPV3", "INS_STAT2_I", 
                "INS_BREAK_I")
)

```

Create the data frame "data" with read_fwf function. 
```{r}
data_frame <- read_fwf( "./NISTEENPUF22.DAT", 
                  col_positions = column_position, 
                  col_types = cols(.default = "c")
)
```

Convert the data frame to a readable and downloadable .csv file
```{r}
output_file<-"NIS_Teen_Data_2022.csv"
write.csv(data_frame, file = output_file, row.names = FALSE)
```

Read the statistics in the data file.
```{r}
summary(NIS_Teen_Data_2022)
str(NIS_Teen_Data_2022)
```

## Research Question
  We want to use the 2022 National Immunization Survey for teens to identify which socioeconomic and geographic factors are associated with HPV vaccination completion for teenagers in the U.S. Specifically, we are asking the question, What is the likelihood of a teenager being up-to-date for their HPV vaccinations in the U.S., based on socioeconomic factors and geographic distribution. The outcome of this study will be measured using the columns that pertain to HPV vaccination completion. These variables are provider-collected, and they state whether or not the teen is (1) up-to-date with 1+ HPV shot, (2) up-to-date with 2+ HPV shots, or (3) up-to-date with 3+ HPV shots (excluding all vaccinations post-survey). Measuring HPV vaccine completion with “up-to-date” variables may be more robust than vaccination rate based on those who have completed the regime (2 or more shots), because not all teens included in the survey are at the age in which they would’ve completed the full regime. This means that younger teens who are up to date with 1+ shot will not be excluded from the analysis.
  We aim to measure determinants of socioeconomic status that are engrained into the survey questions. These factors include: family income, poverty status, income-to-poverty ratio, insurance status, insurance breaks, maternal education, living arrangement, and geographic mobility status. Additional demographic factors that are associated with healthcare access and may be examined include: race, ethnicity, language, facility in which the vaccine was administered, and whether the teen had completed a wellness exam between the ages of 11-12. Geographic distribution of the teens will be assessed by true state of residence. 
  We would like to examine the data for correlations between HPV vaccination completion status and the aforementioned socioeconomic determinants, demographics, and geographic location. This serves as valuable information to determine differences in healthcare access and vaccination coverage for teenagers living in the U.S.

## Proposed Analysis
	For the analysis portion, we have decided to do a correlation analysis between HPV vaccination completion and each of the socioeconomic factors. When performing this, we may choose to stratify by age or gender, if we find these to be confounders. The correlation analysis cannot be performed using a multivariate model because there may be collinearity present between many of the determinants for socioeconomic status. This will lead to spurious results. Instead, the correlation between vaccination status and each factor is examined separately. 
	Next, we may stratify this by state. Comparing these correlation values between states and regions may lead us to find patterns of healthcare accessibility in the U.S. It is generally known that those of lower income and without insurance have limited healthcare access in the U.S.(1), but it will be interesting to see if this pattern is reflected at the state level. If so, policy should be examined as it may be related to healthcare access disparities. Recent studies have examined HPV vaccination completion according to social determinants, but they have not examined geographic distribution and include data prior to 2019 (2). More recent years display differences in global HPV vaccination trends due to the COVID19 pandemic (3). Our study will fill in geographical gaps and provide updated vaccination trends based on socioeconomic factors.

{{< pagebreak >}}

## References
1. Sciences, N. A. of, Engineering, & Medicine, and. (2018, March 1). Factors that affect health-care utilization. Health-Care Utilization as a Proxy in Disability Determination. https://www.ncbi.nlm.nih.gov/books/NBK500097/ 
2. Mansfield, L. N., Chung, R. J., Silva, S. G., Merwin, E. I., & Gonzalez-Guarda, R. M. (2022). Social determinants of human papillomavirus vaccine series completion among U.S. adolescents: A mixed-methods study. SSM - population health, 18, 101082. https://doi.org/10.1016/j.ssmph.2022.101082
3. Casey, R. M., Akaba, H., Hyde, T. B, et al. (2024). Covid-19 pandemic and equity of global human papillomavirus vaccination: descriptive study of World Health Organization-Unicef vaccination coverage estimates. BMJ Medicine, 3, 000726. doi: 10.1136/bmjmed-2023-000726

### 2024-02-02 products/manuscript/Manuscript.qmd
Updated manuscript

--- 2024-02-02 Part1/part1.qmd
@@
 ---
-
-The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. 
-You don't need to have exactly these sections, but the content covering those sections should be addressed.
-
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
-
-```{r, echo=FALSE, message=FALSE}
-# load a few R packages
-library(here)
-library(knitr)
-```
-
-# Summary/Abstract
-_Write a summary of your project._
-
-{{< pagebreak >}}
-
-# Introduction
-
-## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis_
-
-## Description of data and data source
-_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
-
-## Questions/Hypotheses to be addressed
-_State the research questions you plan to answer with this analysis._
-
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key.
-Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].
-
-{{< pagebreak >}}
-
-# Methods
-
-_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._
-
-## Schematic of workflow
-
-Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
-@fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
-We store those figures in the `assets` folder.
-
-```{r}
-#| label: fig-schematic
-#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
-#| echo: FALSE
-knitr::include_graphics(here("products","assets","antigen-recognition.png"))
-```
-
-## Data aquisition
-_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._
-
-## Data import and cleaning
-_Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along._
-
-## Statistical analysis
-_Explain anything related to your statistical analyses._
-
-{{< pagebreak >}}
-
-# Results
-
-## Exploratory/Descriptive analysis
-
-_Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project._
-
-@tbl-summarytable shows a summary of the data.
-
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
-
-```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
-#| echo: FALSE
-resulttable=readRDS("../../results/summarytable.rds")
-knitr::kable(resulttable)
-```
-
-## Basic statistical analysis
-
-_To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._
-
-@fig-result shows a scatterplot figure produced by one of the R scripts.
-
-```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by gender."
-#| echo: FALSE
-knitr::include_graphics(here("results","height-weight-stratified.png"))
-```
-
-## Full analysis
-
-_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here._
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
-
-```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
-#| echo: FALSE
-resulttable2 = readRDS(here("results","tables-files","resulttable2.rds"))
-knitr::kable(resulttable2)
-```
-
-{{< pagebreak >}}
-
-# Discussion
-
-## Summary and Interpretation
-_Summarize what you did, what you found and what it means._
-
-## Strengths and Limitations
-_Discuss what you perceive as strengths and limitations of your analysis._
-
-## Conclusions
-_What are the main take-home messages?_
-
-_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_
-
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.
-
 # *Part 1*
@@
 ```{r}
-data_frame <- read_fwf( "./NISTEENPUF22.DAT", 
+data_frame <- read_fwf( "../products/NISTEENPUF22.DAT", 
                   col_positions = column_position, 
@@
 
-Convert the data frame to a readable and downloadable .csv file
+```{r}
+#Read the statistics in the data file.
+  summary(data_frame)
+  str(data_frame)
+```
+
+The data will then be cleaned. As of now, the data set requires a library to translate the code.
+
+The data will then be converted the data frame to a readable and downloadable .csv file for further analysis
 ```{r}
@@
 write.csv(data_frame, file = output_file, row.names = FALSE)
-```
-
-Read the statistics in the data file.
-```{r}
-summary(NIS_Teen_Data_2022)
-str(NIS_Teen_Data_2022)
 ```

### 2024-02-07 products/manuscript/Manuscript.qmd
Updated data set

@@
 title: "Robertson and Cao Project"
-subtitle: "Teenage HPV Vaccination Coverage and Socioeconomic Factors"
+subtitle: "Part 1:Teenage HPV Vaccination Coverage and Socioeconomic Factors"
 author: Kelly Cao and Rachel Robertson
@@
     highlight-style: github
-bibliography: ../assets/dataanalysis-references.bib
-csl: ../assets/apa.csl
 ---
-# *Part 1*
+```{r setup, include=FALSE} 
+knitr::opts_chunk$set(warning = FALSE, message = FALSE)
+```
+# Introduction
 ## Data Set
@@
   The original data set is a fixed width file (FWF).
-In the data set, there are 43,738 rows and 672 columns, with each row representing an individual and the each column representing the answer to a question.
+In the data set, there are 43,738 rows and 672 columns, with each row representing an individual and each column representing the answer to a question.
 The questions pertain to the teenager’s immunization history, demographics, and additional household-reported health information.
@@
 First, we will load the readr package to open the file
-```{r}
+```{r, echo=FALSE, message=FALSE}
 library(readr) #Loading the package to read the fwf
+library(here) #Loading the package to produce relative pathway
+library(knitr) #Loading in the package for dynamic report generation
+library(tidyverse)
+library(readxl)
+library(dplyr)
 ```
@@
 Next, we will define the column widths (beginning and end positions) and applying the column names to each position.
-```{r}
+```{r, echo=FALSE, message=FALSE}
+#Using the fwf_position() in the readr package to identify the columns widths
 column_position<- fwf_positions(
@@
 Create the data frame "data" with read_fwf function. 
-```{r}
-data_frame <- read_fwf( "../products/NISTEENPUF22.DAT", 
-                  col_positions = column_position, 
-                  col_types = cols(.default = "c")
-)
+```{r, echo=FALSE, message=FALSE}
+data_file_path <- here("products", "manuscript", "NISTEENPUF22.DAT")
+
+data_frame <- read_fwf(data_file_path, col_positions = column_position)
+
 ```
@@
 ```{r}
-#Read the statistics in the data file.
-  summary(data_frame)
-  str(data_frame)
+EDUC1levels=c(1,2,3,4)
+EDUC1labels=c("LESS THAN 12 YEARS", "12 YEARS", "MORE THAN 12 YEARS, NON-COLLEGE GRAD", "COLLEGE GRADUATE")
+
+INCPOVlevels=c(1,2,3,4)
+INCPOVlabels=c("ABOVE POVERTY > $75K", "ABOVE POVERTY <= $75K", "BELOW POVERTY", "UNKNOWN")
+
+INCQ298Alevels=c(10,11,12,13,14,3,4,5,6,7,77,8,9,99)
+INCQ298Alabels=c("$35001 - $40000", "$40001 - $50000", "$50001 - $60000", "$60001 - $75000", "$75001+", "$0 - $7500", "$7501 - $10000", "$10001 - $17500", "$17501 - $20000", "$20001 - $25000", "DON'T KNOW", "$25001 - $30000", "$30001 - $35000", "REFUSED")
+
+LANGUAGElevels=c(1,2,3)
+LANGUAGElabels=c("ENGLISH", "SPANISH", "OTHER")
+
+MOBILlevels=c(1,2,77,98,99)
+MOBILlabels=c("MOVED FROM DIFFERENT STATE", "DID NOT MOVE FROM DIFFERENT STATE", "DON'T KNOW", "MISSING IN ERROR", "REFUSED")
+
+RACEETHKlevels=c(1,2,3,4)
+RACEETHKlabels=c("HISPANIC", "NON-HISPANIC WHITE ONLY", "NON-HISPANIC BLACK ONLY", "NON-HISPANIC OTHER + MULTIPLE RACE")
+
+RACE_Klevels=c(1,2,3)
+RACE_Klabels=c("WHITE ONLY", "BLACK ONLY", "OTHER + MULTIPLE RACE")
+
+RENTOWNlevels=c(1,2,3,77,99)
+RENTOWNlabels=c("OWNED OR BEING BOUGHT", "RENTED", "OTHER ARRANGMENT", "DON'T KNOW", "REFUSED")
+
+SEXlevels=c(1,2,77,98,99)
+SEXlabels=c("MALE", "FEMALE", "DON'T KNOW", "MISSING IN ERROR", "REFUSED")
+
+STATElevels=c(1,10,11,12,13,15,16,17,18,19,2,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,4,40,41,42,44,45,46,47,48,49,5,50,51,53,54,55,56,6,66,72,78,8,9)
+STATElabels=c("ALABAMA", "DELAWARE", "DISTRICT OF COLUMBIA", "FLORIDA", "GEORGIA", "HAWAII", "IDAHO", "ILLINOIS", "INDIANA", "IOWA", "ALASKA", "KANSAS", "KENTUCKY", "LOUISIANA", "MAINE", "MARYLAND", "MASSACHUSETTS", "MICHIGAN", "MINNESOTA", "MISSISSIPPI", "MISSOURI", "MONTANA", "NEBRASKA", "NEVADA",
+              "NEW HAMPSHIRE", "NEW JERSEY", "NEW MEXICO", "NEW YORK", "NORTH CAROLINA", "NORTH DAKOTA", "OHIO", "ARIZONA", "OKLAHOMA", "OREGON", "PENNSYLVANIA", "RHODE ISLAND", "SOUTH CAROLINA", "SOUTH DAKOTA", "TENNESSEE", "TEXAS", "UTAH", "ARKANSAS", "VERMONT", "VIRGINIA", "WASHINGTON", "WEST VIRGINIA",
+              "WISCONSIN", "WYOMING", "CALIFORNIA", "GUAM", "PUERTO RICO", "U.S. VIRGIN ISLANDS", "COLORADO", "CONNECTICUT")
+
+FACILITYlevels=c(1,2,3,4,5,6, ".")
+FACILITYlabels=c("ALL PUBLIC FACILITIES", "ALL HOSPITAL FACILITIES", "ALL PRIVATE FACILITIES", "ALL STD/SCHOOL/TEEN CLINICS OR OTHER FACILITIES", "MIXED", "UNKNOWN", "Missing Data")
+
+WELLCHILDlevels=c(1,2,3)
+WELLCHILDlabels=c("YES", "NO", "DON'T KNOW")
+
+UTDlevels=c(0,1, ".")
+UTDlabels=c("NOT UTD", "UTD", "Missing Data")
+
+INS_STAT2_Ilevels=c(1,2,3,4, ".")
+INS_STAT2_Ilabels=c("PRIVATE INSURANCE ONLY", "ANY MEDICAID", "OTHER INSURANCE (CHIP, IHS, MILITARY, OR OTHER, ALONE OR IN COMB. WITH PRIVATE INSURANCE)", "UNINSURED", "MISSING Data")
+
+INS_BREAK_Ilevels=c(1,2,3,4)
+INS_BREAK_Ilabels=c("CURRENTLY INSURED BUT UNINSURED AT SOME POINT SINCE AGE 11", "CURRENTLY INSURED AND NEVER UNINSURED SINCE AGE 11", "CURRENTLY UNINSURED BUT INSURED AT SOME POINT SINCE AGE 11", "CURRENTLY UNINSURED AND NEVER INSURED SINCE AGE 11")
+
+data<-mutate(data_frame,
+             SEQNUMT = as.numeric(SEQNUMT), 
+             YEAR = as.numeric(YEAR),
+             AGE = as.numeric(AGE),
+             EDUC1 = factor(EDUC1,levels = EDUC1levels, labels = EDUC1labels),
+             INCPORAR_I = as.numeric(INCPORAR_I), 
+             INCPOV1 = factor(INCPOV1, levels = INCPOVlevels, labels = INCPOVlabels),
+             INCQ298A = factor(INCQ298A, levels = INCQ298Alevels, labels = INCQ298Alabels),             
+             LANGUAGE = factor(LANGUAGE, levels = LANGUAGElevels, labels = LANGUAGElabels),
+             MOBIL_1 = factor(MOBIL_1, levels = MOBILlevels, labels = MOBILlabels),
+             RACEETHK = factor(RACEETHK, levels = RACEETHKlevels, labels = RACEETHKlabels),
+             RACE_K = factor(RACE_K, levels = RACE_Klevels, labels = RACE_Klabels),
+             RENT_OWN = factor(RENT_OWN, levels = RENTOWNlevels, labels = RENTOWNlabels), 
+             SEX = factor(SEX, levels = SEXlevels, labels = SEXlabels),
+             STATE = factor(STATE, levels = STATElevels, labels = STATElabels), 
+             FACILITY = factor(FACILITY, levels = FACILITYlevels, labels = FACILITYlabels),
+             WELLCHILD = factor(WELLCHILD, levels = WELLCHILDlevels, labels = WELLCHILDlabels),
+             P_U13HPV = factor(P_U13HPV, levels = UTDlevels, labels = UTDlabels),
+             P_U13HPV3 = factor(P_U13HPV3, levels = UTDlevels, labels = UTDlabels),
+             P_UTDHPV = factor(P_UTDHPV, levels = UTDlevels, labels = UTDlabels),
+             P_UTDHPV_15 = factor(P_UTDHPV_15, levels = UTDlevels, labels = UTDlabels),
+             P_UTDHPV_15INT = factor(P_UTDHPV_15INT, levels = UTDlevels, labels = UTDlabels),
+             P_UTDHPV2 = factor(P_UTDHPV2, levels = UTDlevels, labels = UTDlabels),
+             P_UTDHPV3 = factor(P_UTDHPV3, levels = UTDlevels, labels = UTDlabels),
+             INS_STAT2_I = factor(INS_STAT2_I, levels = INS_STAT2_Ilevels, labels = INS_STAT2_Ilabels),
+             INS_BREAK_I = factor(INS_BREAK_I, levels = INS_BREAK_Ilevels, labels = INS_BREAK_Ilabels)
+)
+
 ```
 
-The data will then be cleaned. As of now, the data set requires a library to translate the code.
+```{r, echo=FALSE, message=FALSE}
+#Read the statistics in the data file.
+  summary(data)
+  str(data)
+```
+
+The data will then be cleaned.
 
 The data will then be converted the data frame to a readable and downloadable .csv file for further analysis
-```{r}
+```{r, echo=FALSE, message=FALSE}
 output_file<-"NIS_Teen_Data_2022.csv"
-write.csv(data_frame, file = output_file, row.names = FALSE)
+write.csv(data, file = output_file, row.names = FALSE)
 ```

### 2024-02-07 products/manuscript/supplement/Creating_Data_Frame.R

@@ -1,0 +1,109 @@
+library(readr) #Loading the package to read the fwf
+library(here) #Loading the package to produce relative pathway
+library(knitr) #Loading in the package for dynamic report generation
+library(tidyverse)
+library(readxl)
+library(dplyr)
+
+column_position<- fwf_positions(
+  start = c(1, 89, 287, 295, 314, 330, 331, 334, 336, 343, 344, 345,
+            347, 353, 359, 362, 472, 473, 495, 496, 497, 502, 504, 1296, 1297
+  ),
+  end = c(5, 92, 288, 295, 329, 330, 332, 334, 336, 343, 344, 346,
+          347, 354, 359, 362, 472, 473, 495, 496, 497, 502, 504, 1296, 1297
+  ),
+  col_names = c("SEQNUMT", "YEAR", "AGE", "EDUC1", "INCPORAR_I", 
+                "INCPOV1", "INCQ298A", "LANGUAGE", "MOBIL_1", "RACEETHK", 
+                "RACE_K", "RENT_OWN", "SEX", "STATE", "FACILITY", "WELLCHILD", 
+                "P_U13HPV", "P_U13HPV3", "P_UTDHPV", "P_UTDHPV_15", 
+                "P_UTDHPV_15INT", "P_UTDHPV2", "P_UTDHPV3", "INS_STAT2_I", 
+                "INS_BREAK_I")
+)
+file_path <- here("products", "manuscript", "NISTEENPUF22.DAT")
+
+data_frame <- read_fwf(file_path, 
+                       col_positions = column_position,
+                       col_types = cols(.default = "c")
+                       )
+                       
+EDUC1levels=c(1,2,3,4)
+EDUC1labels=c("LESS THAN 12 YEARS", "12 YEARS", "MORE THAN 12 YEARS, NON-COLLEGE GRAD", "COLLEGE GRADUATE")
+
+INCPOVlevels=c(1,2,3,4)
+INCPOVlabels=c("ABOVE POVERTY > $75K", "ABOVE POVERTY <= $75K", "BELOW POVERTY", "UNKNOWN")
+
+INCQ298Alevels=c(10,11,12,13,14,3,4,5,6,7,77,8,9,99)
[…]

### 2024-02-19 R/processing-code/processingcode-CandR.R
Rachel's Changes- Delete the data processing file and replace with the opening the dataframe file
I pushed this change, just because I made changes to the ReadMe file in processing data.

@@ -1,0 +1,197 @@
+###############################
+# processing script
+#
+#this script loads the raw data, processes and cleans it 
+#and saves it as Rds file in the processed_data folder
+#
+# Note the ## ---- name ---- notation
+# This is done so one can pull in the chunks of code into the Quarto document
+# see here: https://bookdown.org/yihui/rmarkdown-cookbook/read-chunk.html
+
+
+## ---- packages --------
+#load needed packages. make sure they are installed.
+library(readxl) #for loading Excel files
+library(dplyr) #for data processing/cleaning
+library(tidyr) #for data processing/cleaning
+library(skimr) #for nice visualization of data 
+library(here) #to set paths
+library(readr) #for reading various file types in R
+
+## ---- loaddata --------
+#The here package is used to create a relative path to the raw data before reading the csv
+#The location is called data_locaiton
+data_location <- here::here("data/raw-data/NIS-Teen_Data_2022.csv")
+
+#load data. 
+#The data is found using the read_csv function and the data location
+#The new data frame is called NIS_Teen_Data_2022
+NIS_Teen_Data_2022 <- read_csv(data_location)
+
+
+## ---- exploredata --------
+#glipmse at the structure of the data
+dplyr::glimpse(NIS_Teen_Data_2022)
+
+#produce summary statistics
+summary(NIS_Teen_Data_2022)
+
+#print the first few rows of data from the first several columns
+head(NIS_Teen_Data_2022)
+
+#examine the number of missing values and distribution of character data
+skimr::skim(NIS_Teen_Data_2022)
+
+
+
+
+## ---- cleandata1 --------
+# After inspecting the data, we see that there are several mislabeled variables
+#I will change each variable to the desired type using the mutate() function
+#from dpylr.
+#I will start by changing all of the character variables to factors
+
+NIS_Teen_Data_2022_2 <- NIS_Teen_Data_2022 %>%
+  mutate_if(is.character, as.factor)
+  dplyr::glimpse(NIS_Teen_Data_2022_2)
+  
+#Now I will convert the numeric variables that should be factors, to be factors
+#I will define their levels and labels using mutate()
+NIS_Teen_Data_2022_3 <- NIS_Teen_Data_2022_2 %>%
[…]

### 2024-02-20 products/manuscript/Manuscript.qmd
Kelly's Change- Updated some files
Deleted some unnecessary files, updated the create_data_frame file to save data set into rds file, move files to appropriate location

@@
 write.csv(data, file = output_file, row.names = FALSE)
+
+#A .rds file of it is saved in the following folder
+saveRDS(data, file = "./data/processed-data/processeddata.rds")
 ```

### 2024-02-20 products/manuscript/supplement/Creating_Data_Frame.R

@@ -19,7 +19,7 @@ column_position<- fwf_positions(
                 "P_UTDHPV_15INT", "P_UTDHPV2", "P_UTDHPV3", "INS_STAT2_I", 
                 "INS_BREAK_I")
 )
-file_path <- here("products", "manuscript", "NISTEENPUF22.DAT")
+file_path <- here("data", "raw-data", "NISTEENPUF22.DAT")
 
 data_frame <- read_fwf(file_path, 
                        col_positions = column_position,
@@ -106,4 +106,6 @@ str(data)
 
 output_file<- here("products", "manuscript", "NIS_Teen_Data_2022.csv")
 write.csv(data, file = output_file, row.names = FALSE)
+
+saveRDS(data, file = "./data/processed-data/processeddata.rds")
         

### 2024-02-20 R/processing-code/Cleaning Data/cleaning1.qmd
Rachel Added Cleaning
Hi! I need help saving the cleaned data frame as an RDS and also I am trying to create a Quarto for the R doc output, however it is having trouble pulling the chunks.

---
title: "Data Cleaning"
author: "Rachel Robertson and Kelly Cao"
date: "2/20/24"
output: html_document
---
# Data Cleaning and Exploration
## Finding NAs and Missing Values
I will start cleaning by examining if there are any missing values within the data set.

First, I load the libraries that I need for exploratory analysis.
```{r}
library(here)
library(dplyr)
library(skimr)
library(ggplot2)
library(tidyr)
library(naniar)
```
First, I load the data into a usabel dataframe.
```{r}
data_location <- here::here("data","processed-data","processeddata.rds")
NISTeenVax <- readRDS(data_location)
```

Next, I will check the structure of the data and I will check for any NA or missing values in the data frame.
```{r}
str(NISTeenVax)
dplyr::glimpse(NISTeenVax)
NA_columns <- colSums(is.na(NISTeenVax))
print(NA_columns)
```
There are no missing values, but there may be missing data that is filled with a character value within the data frame. I will first find the character value assigned to missing variables by exploring the levels of each variable using the sapply() and then lapply() function.
```{r}
factor_cols <- sapply(NISTeenVax, is.factor) #Find the columns that are factor variables
factor_levels <- lapply(NISTeenVax[, factor_cols, drop = FALSE], levels) #Find the levels of each factor variable
print(factor_levels)#Pritn factor levels for only the factor columns
```
Several columns contain factor levels that are labelled, "Data missing", "Missing data", "Unknown", "Refused", Missing in error", and "Don't know". I will replace these values for each variable with an NA in R so that they will be counted as missing values for analysis.
```{r}
missing_values <- c("Data Missing", "MISSING Data", "Missing Data", "DON'T KNOW", "UNKNOWN", "MISSING IN ERROR", "REFUSED") #Group all of the factor levels that indicate a missing value

#After back and forth with ChatGPT3.5, it suggested to use the dplyr mutate function and replace function to specify that factor variables with the levels in the group above will have those level values replaced with NA
NISTeenVax2 <- NISTeenVax %>% #Create new dataframe
  mutate(across(where(is.factor), ~ replace(., . %in% missing_values, NA))) #mutate only at factor variables in the dataframe and replace the levels specified above with NA for those factors
```
Now I will be able to check for NA values using the skim() function. I can also display the NA values graphically using gg_miss_var()
```{r}
skimr::skim(NISTeenVax2)
gg_miss_var(NISTeenVax2)
```
The most important factors for this analysis is the response variables called variations of "P_UTDHPV".
We have 27,024 missing values for each of these columns. These missing values may have come from the same survey type, however analysis cannot be performed when this crucial data is missing. For this reason, I will find the rows where these missing values take place, ensure that they are all the same rows for the multiple variables, and delete these rows specifically.
```{r}
#I used chatGPT3,5 to find the function to select for certain rows using dplyr and then filter these for NA values
##I prompted it by using the variable names a,b,c and asked for it to select these columns and filter for the rows with missing variables
rows_with_missing_in_all <- NISTeenVax2 %>%
  select("P_U13HPV", "P_U13HPV3", "P_UTDHPV", "P_UTDHPV_15", "P_UTDHPV_15INT", "P_UTDHPV2", "P_UTDHPV3") %>% #select for columns in question
  filter_all(any_vars(is.na(.))) #Find rows with NA values

##I then prompted ChatGPT3.5 to provide me with code to display the location of these rows so that I may check if they all matched.
if (nrow(rows_with_missing_in_all) > 0) {
  print("Rows with missing values in aforementioned variables")
  print(rows_with_missing_in_all)
} else {
  print("No rows with missing values in aforementioned variables")
}

```
It seems that all of the missing rows match for all of these variables. This means that we may delete the rows that contain missing values for all of the specified variables.

Because the variables listed all contain missing values in the same rows, I will select only one variable to filter.
```{r}
NISTeenVax3 <- NISTeenVax2 %>%
  filter(!is.na(P_UTDHPV)) #drop the rows with NA values in P_UTDHPV
skimr::skim(NISTeenVax3) #check again for missing values in these variables
gg_miss_var(NISTeenVax3)
```
The majority of data is missing from the WELLCHILD column, which indicates whether the teen had receieved a Well Child visit in recent years. Because this is not crucial to out analysis, I will delete this column.

The year column in unnecessary, since all values are from the year 2022. So, I will also delete the YEAR column.
```{r}
NISTeenVax4 <- NISTeenVax3 %>% #create a new data frame for the filtered data
 select(-WELLCHILD, -YEAR) #Use the select() filter to delete multiple columns from the data frame
str(NISTeenVax4) #Check the remaining variables
```
Save cleaned data as an RDS file.
```{r}
output_file2<- here("data", "processed-data", "NIS_Teen_Data_2022_clean1.csv")
write.csv(NISTeenVax4, file = output_file2, row.names = FALSE)
saveRDS(NISTeenVax4, file = "./data/processed-data/clen_data1")
```

### 2024-02-20 R/processing-code/Processing-Quarto-Output.qmd
Rachel Added Cleaning
Hi! I need help saving the cleaned data frame as an RDS and also I am trying to create a Quarto for the R doc output, however it is having trouble pulling the chunks.

---
title: "Data Processing"
author: "Kelly Cao and Rachel Robertson"
date: "2023-01-03"
output: html_document
editor: 
  markdown: 
    wrap: sentence
---

# Setup

This chunk pulls the R-script entitled Creating_Data_Frame.R.

```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('processing-code/creating-df/Creating_Data_Frame.R')
```

```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('R/processing-code/Cleaning Data/cleaning1.R')
```

Load needed packages.

```{r, packages, message = FALSE, warning = FALSE}
```

# Data loading

Note that for functions that come from specific packages (instead of base R), I often specify both package and function like so: package::function() that's not required one could just call the function specifying the package makes it clearer where the function "lives", but it adds typing.
You can do it either way.

```{r, loaddata}
```

# Specifying Factor Variables

By inspecting the data as done above, we find some problems that need addressing:

First, there is an entry for height which says "sixty" instead of a number.
Does that mean it should be a numeric 60?
It somehow doesn't make sense since the weight is 60kg, which can't happen for a 60cm person (a baby).
Since we don't know how to fix this, we might decide to remove the person.
This "sixty" entry also turned all Height entries into characters instead of numeric.
That conversion to character also means that our summary function isn't very meaningful.
So let's fix that first.

```{r, cleandata1}
```

Now we see that there is one person with a height of 6.
That could be a typo, or someone mistakenly entered their height in feet.
Since we unfortunately don't know, we might need to remove this person, which we'll do here.

# Save data

All done, data is clean now.
Time to save.
Let's assign at the end to some final variable, this makes it easier to add further cleaning steps above.

Finally, we save the clean data as RDS file.
I suggest you save your processed and cleaned data as RDS or RDA/Rdata files.
This preserves coding like factors, characters, numeric, etc.
If you save as CSV, that information would get lost.
However, CSV is better for sharing with others since it's plain text.
If you do CSV, you might want to write down somewhere what each variable is.

See here for some suggestions on how to store your processed data: http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata

```{r, savedata}
```

# Cleaning Data

```{r, loaddata2}
```

## Exploring the data

```{r, exploredata2}
```

## Eliminating NA Values and Columns

```{r, cleandata2}
```

## Saving the Data Frame

```{r, savedata2}
```

### 2024-02-20 R/eda-code/edacode.R

@@ -4,6 +4,8 @@ library(here) #for data loading/saving
 library(dplyr)
 library(skimr)
 library(ggplot2)
+library(tidyr)
+library(tidyverse)
 
 ## ---- loaddata --------
 #Path to data. Note the use of the here() package and not absolute paths
@@ -11,14 +13,14 @@ data_location <- here::here("data","processed_data","processeddata.rds")
 #load data
 mydata <- readRDS(data_location)
 
-## ---- table1 --------
+## ---- summary --------
 summary_df = skimr::skim(mydata)
 print(summary_df)
 # save to file
 summarytable_file = here("results", "summarytable.rds")
 saveRDS(summary_df, file = summarytable_file)
 
-## ---- height --------
+## ---- income-poverty --------
 p1 <- mydata %>% ggplot(aes(x=Height)) + geom_histogram() 
 plot(p1)
 figure_file = here("results","height_distribution.png")

### 2024-02-20 R/processing-code/Cleaning Data/cleaning1.R

@@ -1,0 +1,86 @@
+#I will start cleaning by examining if there are any missing values within the data set.
+##First, I load the libraries that I need for exploratory analysis.
+## 
+## ---- packages2 --------
+
+library(here)
+library(dplyr)
+library(skimr)
+library(ggplot2)
+library(tidyr)
+library(naniar)
+
+
+#First, I load the data into a usable data frame.
+## ---- loaddata2 --------
+
+data_location <- here::here("data","processed-data","processeddata.rds")
+NISTeenVax <- readRDS(data_location)
+
+
+#Next, I will check the structure of the data and I will check for any NA or missing values in the data frame.
+## ---- exploredata2 --------
+str(NISTeenVax)
+dplyr::glimpse(NISTeenVax)
+NA_columns <- colSums(is.na(NISTeenVax))
+print(NA_columns)
+
+
+#There are no missing values, but there may be missing data that is filled with a character value within the data frame. I will first find the character value assigned to missing variables by exploring the levels of each variable using the sapply() and then lapply() function.
+
+factor_cols <- sapply(NISTeenVax, is.factor) #Find the columns that are factor variables
+factor_levels <- lapply(NISTeenVax[, factor_cols, drop = FALSE], levels) #Find the levels of each factor variable
+print(factor_levels)#Pritn factor levels for only the factor columns
+
+#Several columns contain factor levels that are labelled, "Data missing", "Missing data", "Unknown", "Refused", Missing in error", and "Don't know". I will replace these values for each variable with an NA in R so that they will be counted as missing values for analysis.
+## ---- cleandata2 --------
+missing_values <- c("Data Missing", "MISSING Data", "Missing Data", "DON'T KNOW", "UNKNOWN", "MISSING IN ERROR", "REFUSED") #Group all of the factor levels that indicate a missing value
+
+#After back and forth with ChatGPT3.5, it suggested to use the dplyr mutate function and replace function to specify that factor variables with the levels in the group above will have those level values replaced with NA
+NISTeenVax2 <- NISTeenVax %>% #Create new dataframe
+  mutate(across(where(is.factor), ~ replace(., . %in% missing_values, NA))) #mutate only at factor variables in the dataframe and replace the levels specified above with NA for those factors
+
+#Now I will be able to check for NA values using the skim() function. I can also display the NA values graphically using gg_miss_var()
+
+skimr::skim(NISTeenVax2)
+gg_miss_var(NISTeenVax2)
+
[…]

### 2024-02-20 R/processing-code/creating-df/Creating_Data_Frame.R

@@ -1,0 +1,114 @@
+## ---- packages --------
+library(readr) #Loading the package to read the fwf
+library(here) #Loading the package to produce relative pathway
+library(knitr) #Loading in the package for dynamic report generation
+library(tidyverse)
+library(readxl)
+library(dplyr)
+
+## ---- loaddata --------
+column_position<- fwf_positions(
+  start = c(1, 89, 287, 295, 314, 330, 331, 334, 336, 343, 344, 345,
+            347, 353, 359, 362, 472, 473, 495, 496, 497, 502, 504, 1296, 1297
+  ),
+  end = c(5, 92, 288, 295, 329, 330, 332, 334, 336, 343, 344, 346,
+          347, 354, 359, 362, 472, 473, 495, 496, 497, 502, 504, 1296, 1297
+  ),
+  col_names = c("SEQNUMT", "YEAR", "AGE", "EDUC1", "INCPORAR_I", 
+                "INCPOV1", "INCQ298A", "LANGUAGE", "MOBIL_1", "RACEETHK", 
+                "RACE_K", "RENT_OWN", "SEX", "STATE", "FACILITY", "WELLCHILD", 
+                "P_U13HPV", "P_U13HPV3", "P_UTDHPV", "P_UTDHPV_15", 
+                "P_UTDHPV_15INT", "P_UTDHPV2", "P_UTDHPV3", "INS_STAT2_I", 
+                "INS_BREAK_I")
+)
+file_path <- here("data", "raw-data", "NISTEENPUF22.DAT")
+
+data_frame <- read_fwf(file_path, 
+                       col_positions = column_position,
+                       col_types = cols(.default = "c")
+                       )
+                       
+## ---- cleandata1 --------
+EDUC1levels=c(1,2,3,4)
+EDUC1labels=c("LESS THAN 12 YEARS", "12 YEARS", "MORE THAN 12 YEARS, NON-COLLEGE GRAD", "COLLEGE GRADUATE")
+
+INCPOVlevels=c(1,2,3,4)
+INCPOVlabels=c("ABOVE POVERTY > $75K", "ABOVE POVERTY <= $75K", "BELOW POVERTY", "UNKNOWN")
+
[…]

### 2024-02-20 R/processing-code/processingcode-CandR.R

@@ -1,197 +1,0 @@
-###############################
-# processing script
-#
-#this script loads the raw data, processes and cleans it 
-#and saves it as Rds file in the processed_data folder
-#
-# Note the ## ---- name ---- notation
-# This is done so one can pull in the chunks of code into the Quarto document
-# see here: https://bookdown.org/yihui/rmarkdown-cookbook/read-chunk.html
-
-
-## ---- packages --------
-#load needed packages. make sure they are installed.
-library(readxl) #for loading Excel files
-library(dplyr) #for data processing/cleaning
-library(tidyr) #for data processing/cleaning
-library(skimr) #for nice visualization of data 
-library(here) #to set paths
-library(readr) #for reading various file types in R
-
-## ---- loaddata --------
-#The here package is used to create a relative path to the raw data before reading the csv
-#The location is called data_locaiton
-data_location <- here::here("data/raw-data/NIS-Teen_Data_2022.csv")
-
-#load data. 
-#The data is found using the read_csv function and the data location
-#The new data frame is called NIS_Teen_Data_2022
-NIS_Teen_Data_2022 <- read_csv(data_location)
-
-
-## ---- exploredata --------
-#glipmse at the structure of the data
-dplyr::glimpse(NIS_Teen_Data_2022)
-
-#produce summary statistics
-summary(NIS_Teen_Data_2022)
-
-#print the first few rows of data from the first several columns
-head(NIS_Teen_Data_2022)
-
-#examine the number of missing values and distribution of character data
-skimr::skim(NIS_Teen_Data_2022)
-
-
-
-
-## ---- cleandata1 --------
-# After inspecting the data, we see that there are several mislabeled variables
-#I will change each variable to the desired type using the mutate() function
-#from dpylr.
-#I will start by changing all of the character variables to factors
-
-NIS_Teen_Data_2022_2 <- NIS_Teen_Data_2022 %>%
-  mutate_if(is.character, as.factor)
-  dplyr::glimpse(NIS_Teen_Data_2022_2)
-  
-#Now I will convert the numeric variables that should be factors, to be factors
-#I will define their levels and labels using mutate()
-NIS_Teen_Data_2022_3 <- NIS_Teen_Data_2022_2 %>%
[…]

### 2024-02-20 products/manuscript/supplement/Creating_Data_Frame.R

@@ -1,111 +1,0 @@
-library(readr) #Loading the package to read the fwf
-library(here) #Loading the package to produce relative pathway
-library(knitr) #Loading in the package for dynamic report generation
-library(tidyverse)
-library(readxl)
-library(dplyr)
-
-column_position<- fwf_positions(
-  start = c(1, 89, 287, 295, 314, 330, 331, 334, 336, 343, 344, 345,
-            347, 353, 359, 362, 472, 473, 495, 496, 497, 502, 504, 1296, 1297
-  ),
-  end = c(5, 92, 288, 295, 329, 330, 332, 334, 336, 343, 344, 346,
-          347, 354, 359, 362, 472, 473, 495, 496, 497, 502, 504, 1296, 1297
-  ),
-  col_names = c("SEQNUMT", "YEAR", "AGE", "EDUC1", "INCPORAR_I", 
-                "INCPOV1", "INCQ298A", "LANGUAGE", "MOBIL_1", "RACEETHK", 
-                "RACE_K", "RENT_OWN", "SEX", "STATE", "FACILITY", "WELLCHILD", 
-                "P_U13HPV", "P_U13HPV3", "P_UTDHPV", "P_UTDHPV_15", 
-                "P_UTDHPV_15INT", "P_UTDHPV2", "P_UTDHPV3", "INS_STAT2_I", 
-                "INS_BREAK_I")
-)
-file_path <- here("data", "raw-data", "NISTEENPUF22.DAT")
-
-data_frame <- read_fwf(file_path, 
-                       col_positions = column_position,
-                       col_types = cols(.default = "c")
-                       )
-                       
-EDUC1levels=c(1,2,3,4)
-EDUC1labels=c("LESS THAN 12 YEARS", "12 YEARS", "MORE THAN 12 YEARS, NON-COLLEGE GRAD", "COLLEGE GRADUATE")
-
-INCPOVlevels=c(1,2,3,4)
-INCPOVlabels=c("ABOVE POVERTY > $75K", "ABOVE POVERTY <= $75K", "BELOW POVERTY", "UNKNOWN")
-
-INCQ298Alevels=c(10,11,12,13,14,3,4,5,6,7,77,8,9,99)
[…]

### 2024-02-20 R/processing-code/Cleaning Data/cleaning1.qmd
Editting Cleaning doc 1

@@
 factor_levels <- lapply(NISTeenVax[, factor_cols, drop = FALSE], levels) #Find the levels of each factor variable
-print(factor_levels)#Pritn factor levels for only the factor columns
+print(factor_levels)#Print factor levels for only the factor columns
 ```
@@
 ```{r}
-missing_values <- c("Data Missing", "MISSING Data", "Missing Data", "DON'T KNOW", "UNKNOWN", "MISSING IN ERROR", "REFUSED") #Group all of the factor levels that indicate a missing value
+missing_values <- c("Data Missing", "MISSING Data", "Missing Data", "MISSING IN ERROR") #Group all of the factor levels that indicate a missing value
 
@@
 skimr::skim(NISTeenVax2)
-gg_miss_var(NISTeenVax2)
+naniar::gg_miss_var(NISTeenVax2)
 ```
@@
   filter(!is.na(P_UTDHPV)) #drop the rows with NA values in P_UTDHPV
+NISTeenVax3 <- NISTeenVax3 %>%
+  filter(!is.na(INS_BREAK_I))
+NISTeenVax3 <- NISTeenVax3 %>%
+  filter(!is.na(FACILITY))
 skimr::skim(NISTeenVax3) #check again for missing values in these variables
@@
 ```
+
 Save cleaned data as an RDS file.
@@
 write.csv(NISTeenVax4, file = output_file2, row.names = FALSE)
-saveRDS(NISTeenVax4, file = "./data/processed-data/clen_data1")
+saveRDS(NISTeenVax4, file = "./data/processed-data/clean_data1.rds")
 ```

### 2024-02-20 R/processing-code/Cleaning Data/cleaning1.qmd
Uploaded clean1

@@
 write.csv(NISTeenVax4, file = output_file2, row.names = FALSE)
-saveRDS(NISTeenVax4, file = "./data/processed-data/clean_data1.rds")
+saveRDS(NISTeenVax4, file = "./data/processed-data/cleandata1.rds")
 ```

### 2024-02-20 R/processing-code/Cleaning Data/cleaning1.R

@@ -81,6 +81,7 @@ str(NISTeenVax4) #Check the remaining variables
 ## ---- savedata2 --------
 output_file2<- here("data", "processed-data", "NIS_Teen_Data_2022_clean1.csv")
 write.csv(NISTeenVax4, file = output_file2, row.names = FALSE)
-saveRDS(NISTeenVax4, file = "./data/processed-data/clen_data1")
+saveRDS(NISTeenVax4, file = "./data/processed-data/cleandata1.rds")
+
 
 

### 2024-02-20 R/processing-code/Cleaning Data/cleaning1.R
dummycommit

@@ -34,7 +34,7 @@ print(factor_levels)#Pritn factor levels for only the factor columns
 
 #Several columns contain factor levels that are labelled, "Data missing", "Missing data", "Unknown", "Refused", Missing in error", and "Don't know". I will replace these values for each variable with an NA in R so that they will be counted as missing values for analysis.
 ## ---- cleandata2 --------
-missing_values <- c("Data Missing", "MISSING Data", "Missing Data", "DON'T KNOW", "UNKNOWN", "MISSING IN ERROR", "REFUSED") #Group all of the factor levels that indicate a missing value
+missing_values <- c("Data Missing", "MISSING Data", "Missing Data", "MISSING IN ERROR") #Group all of the factor levels that indicate a missing value
 
 #After back and forth with ChatGPT3.5, it suggested to use the dplyr mutate function and replace function to specify that factor variables with the levels in the group above will have those level values replaced with NA
 NISTeenVax2 <- NISTeenVax %>% #Create new dataframe
@@ -67,6 +67,13 @@ if (nrow(rows_with_missing_in_all) > 0) {
 
 NISTeenVax3 <- NISTeenVax2 %>%
   filter(!is.na(P_UTDHPV)) #drop the rows with NA values in P_UTDHPV
+NISTeenVax3 <- NISTeenVax3 %>%
+  filter(!is.na(INS_BREAK_I))
+NISTeenVax3 <- NISTeenVax3 %>%
+  filter(!is.na(FACILITY))
+
+skimr::skim(NISTeenVax3) #check again for missing values in these variables
+gg_miss_var(NISTeenVax3)
 skimr::skim(NISTeenVax3) #check again for missing values in these variables
 gg_miss_var(NISTeenVax3)
 

### 2024-02-21 R/processing-code/creating-df/Creating_Data_Frame.R
Reorganized repo, updated processing output file

@@ -27,8 +27,9 @@ data_frame <- read_fwf(file_path,
                        col_positions = column_position,
                        col_types = cols(.default = "c")
                        )
-                       
-## ---- cleandata1 --------
+                     
+  
+## ---- cleandata --------
 EDUC1levels=c(1,2,3,4)
 EDUC1labels=c("LESS THAN 12 YEARS", "12 YEARS", "MORE THAN 12 YEARS, NON-COLLEGE GRAD", "COLLEGE GRADUATE")
 
@@ -107,8 +108,9 @@ data<-mutate(data_frame,
 summary(data)
 str(data)
 
+## ---- savedata --------
 output_file<- here("data", "processed-data", "NIS_Teen_Data_2022.csv")
 write.csv(data, file = output_file, row.names = FALSE)
-
-saveRDS(data, file = "./data/processed-data/processeddata.rds")
+save_path <- here("./data/processed-data/processeddata.rds")
+saveRDS(data, file = save_path )
         

### 2024-02-21 R/processing-code/Processing-Quarto-Output.qmd

@@
 ```{r, include=FALSE, cache=FALSE}
-knitr::read_chunk('processing-code/creating-df/Creating_Data_Frame.R')
-```
-
-```{r, include=FALSE, cache=FALSE}
-knitr::read_chunk('R/processing-code/Cleaning Data/cleaning1.R')
+library(here)
+knitr::read_chunk(here("./R/processing-code/creating-df/Creating_Data_Frame.R"))
 ```
@@
 
-```{r, cleandata1}
+```{r, cleandata}
+
 ```
@@
 
+```{r, include=FALSE, cache=FALSE}
+library(here)
+knitr::read_chunk(here("./R/processing-code/cleaningdata/cleaning1.R"))
+```
+
+```{r, packages2}
+```
+
 ```{r, loaddata2}

### 2024-02-21 R/eda-code/EDA-code.qmd
Created edacode.R and eda-code.qmd. Uploaded some tables and figures in /result directory

---
title: "Exploratory Analysis on CDC Teen Vaccination Surveys"
date: "`r format(Sys.Date()'"
Author: Kelly Cao and Rachel Robertson
output: html_document
editor: 
  markdown: 
    wrap: sentence
---

The following file will load the cleaned data and perform some exploration.

The categorical and numerical variables of interest will be plotted and placed in models for analysis.

## Loading the .R script which performed the analysis

```{r, include=FALSE, cache=FALSE}
library(here)
knitr::read_chunk(here("./R/eda-code/edacode.R"))
```

## Loading the Packages

```{r, packages, message = FALSE, warning = FALSE}
```

```{r, loaddata}
```

## Showing Summary of Overall Data Set

The following data set has been cleaned. Any observations with NAs has been removed. 
```{r, table1}

```

## Addressing Numerical Data Set
The distribution of the numerical income to poverty ratio can be shown below. It displays a value from 1-3. A ratio displaying less than 1 depicts an income less than the poverty level. Any income ratio >1 indicates an income greater than the poverty level. Most of the observations were gathered from households with income ratios of 3, as shown by the skewed plot below.

```{r, INCPORAR_I}

```

## Addressing Categorical Data

The state variable shows a more uniformed distribution, suggesting that the observations were gathered fairly across the US states and territory. Although it is noted a few territories provides 0 observations.

```{r, STATE}

```

The family income variable provides a more normal distribution with the majority of the observation coming from $75000 income level.

```{r, INCQ298A}

```

The race and ethnicity variables show a normal distribution.
```{r, RACEETHK}

```

```{r, RACE_K}

```

The Insurance status variable shows that the majority of the observations came from households with private insurance only. This is mirrored in the facility variable, where a majority of its observation s show private facilities being the location.

```{r, INS_STAT2_I}

```

```{r, FACILITY}

```

The vaccination variable indicates nearly 80% of the observations had an Up-to-date status for the 1+ shot HPV vaccination.

```{r, Vaccination}

```

## Model and Plot Fitting

### Fit 1

A boxplot is created with income-poverty ratio to vaccination status. The box plot indicates little difference between the two status. A statistical model is fitted to the same variables. A p-value of 0.1358 suggests there is little significance between the income-poverty variable and the vaccination status.

```{r, fitfig1}

```

### Fit 2
A boxplot is created with income-poverty ratio to insurance status. The box plot indicates there are distinct difference between the two status. A statistical model is fitted to the same variables. The p-value of < 2.26e-16 suggests there is statistical significance between the income-poverty variable and the insurance status.

```{r, fitfig2}

```

### 2024-02-22 code/eda-code/EDA-code.qmd
Added to EDA, staged in Manuscript
I added 5 figures/ tables to the EDA and in addition I added the EDA chunk to the manuscript document. I would prefer to just add the figure images preceding the description, but I cannot get markdown to allow me to pull just the figures.

--- 2024-02-21 R/eda-code/EDA-code.qmd
@@
 library(here)
-knitr::read_chunk(here("./R/eda-code/edacode.R"))
+knitr::read_chunk(here("./code/eda-code/edacode.R"))
 ```
@@
 
-The following data set has been cleaned. Any observations with NAs has been removed. 
+The following data set has been cleaned.
+Any observations with NAs has been removed.
+
 ```{r, table1}
@@
 ## Addressing Numerical Data Set
+
 The distribution of the numerical income to poverty ratio can be shown below.
@@
 A ratio displaying less than 1 depicts an income less than the poverty level.
-Any income ratio >1 indicates an income greater than the poverty level.
+Any income ratio \>1 indicates an income greater than the poverty level.
 Most of the observations were gathered from households with income ratios of 3, as shown by the skewed plot below.
@@
 
-The family income variable provides a more normal distribution with the majority of the observation coming from $75000 income level.
+The family income variable provides a more normal distribution with the majority of the observation coming from \$75000+ income level.
 
@@
 
-The race and ethnicity variables show a normal distribution.
+The race and ethnicity variables show a distribution that is similar to the U.S.
+census data, which concludes that the racial composition of the U.S.
+is 58.9% white, 13.6% African American, and 19.1% Hispanic or Latino.
+
 ```{r, RACEETHK}
@@
 
-The Insurance status variable shows that the majority of the observations came from households with private insurance only.
-This is mirrored in the facility variable, where a majority of its observation s show private facilities being the location.
+The Insurance status variable shows that the majority of the observations came from households with private insurance only. This is followed by any medicaid and then other insurance.
+This is mirrored in the facility variable, where a majority of its observations show private facilities being the location.
 
@@
 
+We examine the percentage of up-to-date vaccination status by state.
+```{r, state-vaccine}
+
+```
+
+```{r, facility-vaccine}
+
+```
+
+```{r, race-ethnicity-vaccine}
+
+```
 ## Model and Plot Fitting
@@
 ### Fit 2
+
 A boxplot is created with income-poverty ratio to insurance status.
@@
 A statistical model is fitted to the same variables.
-The p-value of < 2.26e-16 suggests there is statistical significance between the income-poverty variable and the insurance status.
+The p-value of \< 2.26e-16 suggests there is statistical significance between the income-poverty variable and the insurance status.
 
@@
 ```
+
+### Fit 3
+```{r, fitfig3}
+
+```
+
+### Fit 4
+```{r, fitfig4}
+
+```

### 2024-02-22 code/processing-code/Processing-Quarto-Output.qmd
Added to EDA, staged in Manuscript
I added 5 figures/ tables to the EDA and in addition I added the EDA chunk to the manuscript document. I would prefer to just add the figure images preceding the description, but I cannot get markdown to allow me to pull just the figures.

--- 2024-02-21 R/processing-code/Processing-Quarto-Output.qmd
@@
 library(here)
-knitr::read_chunk(here("./R/processing-code/creating-df/Creating_Data_Frame.R"))
+knitr::read_chunk(here("./code/processing-code/creating-df/Creating_Data_Frame.R"))
 ```
@@
 
-Note that for functions that come from specific packages (instead of base R), I often specify both package and function like so: package::function() that's not required one could just call the function specifying the package makes it clearer where the function "lives", but it adds typing.
-You can do it either way.
-
 ```{r, loaddata}
@@
 
-By inspecting the data as done above, we find some problems that need addressing:
-
-First, there is an entry for height which says "sixty" instead of a number.
-Does that mean it should be a numeric 60?
-It somehow doesn't make sense since the weight is 60kg, which can't happen for a 60cm person (a baby).
-Since we don't know how to fix this, we might decide to remove the person.
-This "sixty" entry also turned all Height entries into characters instead of numeric.
-That conversion to character also means that our summary function isn't very meaningful.
-So let's fix that first.
-
 ```{r, cleandata}
@@
 
-Now we see that there is one person with a height of 6.
-That could be a typo, or someone mistakenly entered their height in feet.
-Since we unfortunately don't know, we might need to remove this person, which we'll do here.
-
 # Save data
-
-All done, data is clean now.
-Time to save.
-Let's assign at the end to some final variable, this makes it easier to add further cleaning steps above.
-
-Finally, we save the clean data as RDS file.
-I suggest you save your processed and cleaned data as RDS or RDA/Rdata files.
-This preserves coding like factors, characters, numeric, etc.
-If you save as CSV, that information would get lost.
-However, CSV is better for sharing with others since it's plain text.
-If you do CSV, you might want to write down somewhere what each variable is.
-
-See here for some suggestions on how to store your processed data: http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata
 
@@
 library(here)
-knitr::read_chunk(here("./R/processing-code/cleaningdata/cleaning1.R"))
+knitr::read_chunk(here("./code/processing-code/cleaningdata/cleaning1.R"))
 ```

### 2024-02-22 products/manuscript/Manuscript.qmd
Added to EDA, staged in Manuscript
I added 5 figures/ tables to the EDA and in addition I added the EDA chunk to the manuscript document. I would prefer to just add the figure images preceding the description, but I cannot get markdown to allow me to pull just the figures.

--- 2024-02-02 products/manuscript/Manuscript.qmd
@@
     highlight-style: github
-bibliography: ../assets/dataanalysis-references.bib
-csl: ../assets/apa.csl
+editor: 
+  markdown: 
+    wrap: sentence
 ---
-# *Part 1*
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(warning = FALSE, message = FALSE)
+```
+
+# Introduction
+
 ## Data Set
+
 The data we chose to use is the National Immunization Survey (NIS) of 2022.
@@
 The surveys were conducted by the National Center for Immunization and Respiratory Diseases of the Center (NCIRD) for Disease Control and Prevention (CDC).
-The survey itself consists of two parts: (1) the household telephone (random digit dialing) survey answered by a parent or guardian and (2) a mailed survey for the vaccination provider, called “Immunization History Questionnaire.”
-  The original data set is a fixed width file (FWF).
-In the data set, there are 43,738 rows and 672 columns, with each row representing an individual and the each column representing the answer to a question.
+The survey itself consists of two parts: (1) the household telephone (random digit dialing) survey answered by a parent or guardian and (2) a mailed survey for the vaccination provider, called “Immunization History Questionnaire.” The original data set is a fixed width file (FWF).
+In the data set, there are 43,738 rows and 672 columns, with each row representing an individual and each column representing the answer to a question.
 The questions pertain to the teenager’s immunization history, demographics, and additional household-reported health information.
@@
 The provided survey data will require cleaning, as there are signs of human errors present.
-## Opening the Data Set
-
-First, we will load the readr package to open the file
-```{r}
-library(readr) #Loading the package to read the fwf
-```
-
-Next, we will define the column widths (beginning and end positions) and applying the column names to each position.
-```{r}
-column_position<- fwf_positions(
-  start = c(1, 89, 287, 295, 298, 314, 330, 331, 334, 336, 343, 344, 345,
-            347, 355, 359, 362, 472, 473, 495, 496, 497, 502, 504, 1296, 1297
-  ),
-  end = c(5, 92, 288, 295, 313, 329, 330, 332, 334, 336, 343, 344, 346,
-          347, 356, 359, 362, 472, 473, 495, 496, 497, 502, 504, 1296, 1297
-  ),
-  col_names = c("SEQNUMT", "YEAR", "AGE", "EDUC1", "INCPORAR", "INCPORAR_I", 
-                "INCPOV1", "INCQ298A", "LANGUAGE", "MOBIL_1", "RACEETHK", 
-                "RACE_K", "RENT_OWN", "SEX", "STATE", "FACILITY", "WELLCHILD", 
-                "P_U13HPV", "P_U13HPV3", "P_UTDHPV", "P_UTDHPV_15", 
-                "P_UTDHPV_15INT", "P_UTDHPV2", "P_UTDHPV3", "INS_STAT2_I", 
-                "INS_BREAK_I")
-)
-
-```
-
-Create the data frame "data" with read_fwf function. 
-```{r}
-data_frame <- read_fwf( "../products/NISTEENPUF22.DAT", 
-                  col_positions = column_position, 
-                  col_types = cols(.default = "c")
-)
-```
-
-```{r}
-#Read the statistics in the data file.
-  summary(data_frame)
-  str(data_frame)
-```
-
-The data will then be cleaned. As of now, the data set requires a library to translate the code.
-
-The data will then be converted the data frame to a readable and downloadable .csv file for further analysis
-```{r}
-output_file<-"NIS_Teen_Data_2022.csv"
-write.csv(data_frame, file = output_file, row.names = FALSE)
-```
 
 ## Research Question
+
 We want to use the 2022 National Immunization Survey for teens to identify which socioeconomic and geographic factors are associated with HPV vaccination completion for teenagers in the U.S.
@@
 ## Proposed Analysis
+
+```         
 For the analysis portion, we have decided to do a correlation analysis between HPV vaccination completion and each of the socioeconomic factors.
@@
 Our study will fill in geographical gaps and provide updated vaccination trends based on socioeconomic factors.
+```
+
+{{< pagebreak >}}
+
+## Methods
+
+### Data Cleaning and Processing
+
+#### Loading the data
+
+The data was loaded from the original DAT file, entitled 'NISTEENPUF22.DAT' by specifying the column positions within the data file.
+The columns were then read and made into a data frame.
+Next, each of the factor variables were defined and the corresponding levels were assigned labels.
+The structure and summary statistics of all of the variables were examined.
+The loaded data is saved as a rds and a csv in the 'processed-data' folder.
+This does not alter the original raw data file.
+The definition of each factor variable in included in the README file within the 'raw-data' folder.
+
+Next, we explored the structure of the processed data and searched for any missing values.
+However, there were no NAs present.
+There are, however, missing values as labelled factor levels which must be accounted for.
+We identified the missing value labels by printing the factor levels.
+We then replaced the all of the values labelled missing with NA values.
+Next, we found a string of over 27000 NAs all of the variables for vaccination completion.
+These rows corresponded for all variables related to vaccine completion.
+Since this is the response variable of our scientific question, we eliminated these rows.
+We also eliminated rows with small numbers missing values, which were within the columns for facility, Wellchild exams, and insurance breaks.
+Lastly, we eliminatethe year column, as all of the data comes from the year 2022.
+We saved the cleaned data as an rds and csv in the 'processed-data' folder.
+
+### Exploratory Analysis
+
+```{r, include=FALSE, cache=FALSE}
+library(here)
+knitr::read_chunk(here("./code/eda-code/edacode.R"))
+```
+
+First we will load the packages needed for the exploratory analysis.
+```{r, packages, message = FALSE, warning = FALSE}
+```
+
+Next, we will load the data called 'cleandata1' from the 'processed-data' folder.
+```{r, loaddata}
+```
+
+The following data set has been cleaned.
+Any observations with NAs has been removed.
+
+```{r, table1}
+
+```
+
+The distribution of the numerical income to poverty ratio can be shown below.
+It displays a value from 1-3.
+A ratio displaying less than 1 depicts an income less than the poverty level.
+Any income ratio \>1 indicates an income greater than the poverty level.
+Most of the observations were gathered from households with income ratios of 3, as shown by the skewed plot below.
+
+```{r, INCPORAR_I}
+
+```
+
+The state variable shows a more uniformed distribution, suggesting that the observations were gathered fairly across the US states and territory.
+Although it is noted a few territories provides 0 observations.
+
+```{r, STATE}
+
+```
+
+The family income variable provides a more normal distribution with the majority of the observation coming from \$75000+ income level.
+
+```{r, INCQ298A}
+
+```
+
+The race and ethnicity variables show a distribution that is similar to the U.S.
+census data, which concludes that the racial composition of the U.S.
+is 58.9% white, 13.6% African American, and 19.1% Hispanic or Latino.
+
+```{r, RACEETHK}
+
+```
+
+```{r, RACE_K}
+
+```
+
+The Insurance status variable shows that the majority of the observations came from households with private insurance only.
+This is followed by any medicaid and then other insurance.
+This is mirrored in the facility variable, where a majority of its observations show private facilities being the location.
+
+```{r, INS_STAT2_I}
+
+```
+
+```{r, FACILITY}
+
+```
+
+The vaccination variable indicates nearly 80% of the observations had an Up-to-date status for the 1+ shot HPV vaccination.
+
+```{r, Vaccination}
+
+```
+
+We examine the percentage of up-to-date vaccination status by state.
+
+```{r, state-vaccine}
+
+```
+
+We examine the facility that submitted the survey, which is stratified by the vaccine completion status of the survey subject.
+
+```{r, facility-vaccine}
+
+```
+
+We examine the percentage of vaccination completion by race and ethnicity.
+Because there are different counts of each race/ethnicity for each factor level, the percentage of those with up-to-date vaccine status out of each race/ethnicity level was determined.
+This adjusts for the different counts in each racial category.
+The Hispanic ethnicity was found to have the highest vaccine completion while non-hispanic white has the lowest.
+
+```{r, race-ethnicity-vaccine}
+
+```
+
+### Model and Plot Fitting
+
+A boxplot is created with income-poverty ratio to vaccination status.
+The box plot indicates little difference between the two status.
+A statistical model is fitted to the same variables.
+A p-value of 0.1358 suggests there is little significance between the income-poverty variable and the vaccination status.
+
+```{r, fitfig1}
+
+```
+
+A boxplot is created with income-poverty ratio to insurance status.
+The box plot indicates there are distinct difference between the two status.
+A statistical model is fitted to the same variables.
+The p-value of \< 2.26e-16 suggests there is statistical significance between the income-poverty variable and the insurance status.
+
+```{r, fitfig2}
+
+```
+
+Income-poverty ratio was fit to vaccine status and insurance status in a multi-linear model.
+
+```{r, fitfig3}
+
+```
+
+A box-plot was created for income-poverty level versus race/ethnicity and this was stratified by vaccine completion (UTD or NOT UTD).
+A lower income-poverty ratio means the group is closer to the poverty line.
+For the non-Hispanic black and Hispanic racial and ethnic groups, vaccine completion corresponded with a lower mean income-poverty ratio.
+For mixed ethnic groups, vaccine completion corresponded with a higher mean income-poverty ratio.
+
+```{r, fitfig4}
+
+```
 
@@
 ## References
+
 1.

### 2024-02-22 R/eda-code/edacode.R

@@ -1,291 +1,0 @@
-## ---- packages --------
-#load needed packages. make sure they are installed.
-library(here) #for data loading/saving
-library(dplyr)
-library(skimr)
-library(ggplot2)
-
-## ---- loaddata --------
-#Path to data. Note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed-data","cleandata1.rds")
-#load data
-mydata <- readRDS(data_location)
-
-## ---- table1 --------
-summary_df = skimr::skim(mydata)
-print(summary_df)
-# save to file
-summarytable_file = here("results","tables", "summarytable.rds")
-saveRDS(summary_df, file = summarytable_file)
-
-## ---- INCPORAR_I --------
-p1 <- mydata %>% ggplot(aes(x=INCPORAR_I)) + geom_histogram() 
-plot(p1)
-figure_file = here("results", "figures", "income.png")
-ggsave(filename = figure_file, plot=p1) 
-
-## ---- STATE --------
-summary(mydata$STATE)
-state_distribution <- table(mydata$STATE)
-# Convert STATE to character
-mydata$STATE <- as.character(mydata$STATE)
-
-# Calculate distribution percentage
-state_distribution_percentage <- prop.table(table(mydata$STATE))*100
-
-print(state_distribution_percentage)
-
-
-p2 <- barplot(state_distribution_percentage, 
-        main = "State Distribution Percentage", 
-        xlab = "State", 
-        ylab = "Percentage",
-        col = "skyblue",   # Set bar color
-        las = 2) 
-#Recreating the graph in ggplot
-
-#adjusting the percent disitrbution data into a data frame
-data_df <- as.data.frame(as.table(state_distribution_percentage))
-
-#plot with ggplot
-p3 <- ggplot(data_df, aes(x = Var1, y = Freq)) +
-  geom_bar(stat = "identity", fill = "skyblue") +
-  labs(title = "State Distribution Percentage",
-       x = "State",
-       y = "Percentage")
-
-#Viewing the plot to confirm it is done properly
-print(p3)
-
-figure_file = here("results", "figures", "state.distribution.png")
-ggsave(filename = figure_file, plot=p3) 
-
-
-## ---- INCQ298A --------
-summary(mydata$INCQ298A)
-Family_income_distribution <- table(mydata$INCQ298A)
-
-# Calculate distribution percentage
-Family_Income_distribution <- prop.table(table(mydata$INCQ298A))*100
-
[…]

### 2024-02-22 R/processing-code/cleaningdata/cleaning1.R

@@ -1,95 +1,0 @@
-#I will start cleaning by examining if there are any missing values within the data set.
-##First, I load the libraries that I need for exploratory analysis.
-## 
-
-## ---- packages2 --------
-library(here)
-library(dplyr)
-library(skimr)
-library(ggplot2)
-library(tidyr)
-library(naniar)
-
-
-#First, I load the data into a usable data frame.
-## ---- loaddata2 --------
-
-data_location <- here::here("data","processed-data","processeddata.rds")
-NISTeenVax <- readRDS(data_location)
-
-
-#Next, I will check the structure of the data and I will check for any NA or missing values in the data frame.
-## ---- exploredata2 --------
-str(NISTeenVax)
-dplyr::glimpse(NISTeenVax)
-NA_columns <- colSums(is.na(NISTeenVax))
-print(NA_columns)
-
-
-#There are no missing values, but there may be missing data that is filled with a character value within the data frame. I will first find the character value assigned to missing variables by exploring the levels of each variable using the sapply() and then lapply() function.
-
-factor_cols <- sapply(NISTeenVax, is.factor) #Find the columns that are factor variables
-factor_levels <- lapply(NISTeenVax[, factor_cols, drop = FALSE], levels) #Find the levels of each factor variable
-print(factor_levels)#Pritn factor levels for only the factor columns
-
-#Several columns contain factor levels that are labelled, "Data missing", "Missing data", "Unknown", "Refused", Missing in error", and "Don't know". I will replace these values for each variable with an NA in R so that they will be counted as missing values for analysis.
-## ---- cleandata2 --------
-missing_values <- c("Data Missing", "MISSING Data", "Missing Data", "MISSING IN ERROR") #Group all of the factor levels that indicate a missing value
-
-#After back and forth with ChatGPT3.5, it suggested to use the dplyr mutate function and replace function to specify that factor variables with the levels in the group above will have those level values replaced with NA
-NISTeenVax2 <- NISTeenVax %>% #Create new dataframe
-  mutate(across(where(is.factor), ~ replace(., . %in% missing_values, NA))) #mutate only at factor variables in the dataframe and replace the levels specified above with NA for those factors
-
-#Now I will be able to check for NA values using the skim() function. I can also display the NA values graphically using gg_miss_var()
-
-skimr::skim(NISTeenVax2)
-gg_miss_var(NISTeenVax2)
[…]

### 2024-02-22 code/eda-code/edacode.R

@@ -1,0 +1,377 @@
+## ---- packages --------
+#load needed packages. make sure they are installed.
+library(here) #for data loading/saving
+library(dplyr)
+library(skimr)
+library(ggplot2)
+library(knitr)
+
+## ---- loaddata --------
+#Path to data. Note the use of the here() package and not absolute paths
+data_location <- here::here("data","processed-data","cleandata1.rds")
+#load data
+mydata <- readRDS(data_location)
+
+## ---- table1 --------
+summary_df = skimr::skim(mydata)
+print(summary_df)
+# save to file
+summarytable_file = here("results","tables", "summarytable.rds")
+saveRDS(summary_df, file = summarytable_file)
+
+## ---- INCPORAR_I --------
+p1 <- mydata %>% ggplot(aes(x=INCPORAR_I)) + geom_histogram() 
+plot(p1)
+figure_file = here("results", "figures", "income.png")
+ggsave(filename = figure_file, plot=p1) 
+
+## ---- STATE --------
+summary(mydata$STATE)
+state_distribution <- table(mydata$STATE)
+# Convert STATE to character
+mydata$STATE <- as.character(mydata$STATE)
+
+# Calculate distribution percentage
+state_distribution_percentage <- prop.table(table(mydata$STATE))*100
+
+print(state_distribution_percentage)
+
+
+p2 <- barplot(state_distribution_percentage, 
+        main = "State Distribution Percentage", 
+        xlab = "State", 
+        ylab = "Percentage",
+        col = "skyblue",   # Set bar color
+        las = 2) 
+#Recreating the graph in ggplot
+
+#adjusting the percent disitrbution data into a data frame
+data_df <- as.data.frame(as.table(state_distribution_percentage))
+
+#plot with ggplot
+p3 <- ggplot(data_df, aes(x = Var1, y = Freq)) +
+  geom_bar(stat = "identity", fill = "skyblue") +
+  labs(title = "State Distribution Percentage",
+       x = "State",
+       y = "Percentage") +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
+
+#Viewing the plot to confirm it is done properly
+print(p3)
+
+figure_file = here("results", "figures", "state.distribution.png")
+ggsave(filename = figure_file, plot=p3) 
+
+
+## ---- INCQ298A --------
+summary(mydata$INCQ298A)
+Family_income_distribution <- table(mydata$INCQ298A)
+
[…]

### 2024-02-22 code/processing-code/cleaningdata/cleaning1.R

@@ -1,0 +1,95 @@
+#I will start cleaning by examining if there are any missing values within the data set.
+##First, I load the libraries that I need for exploratory analysis.
+## 
+
+## ---- packages2 --------
+library(here)
+library(dplyr)
+library(skimr)
+library(ggplot2)
+library(tidyr)
+library(naniar)
+
+
+#First, I load the data into a usable data frame.
+## ---- loaddata2 --------
+
+data_location <- here::here("data","processed-data","processeddata.rds")
+NISTeenVax <- readRDS(data_location)
+
+
+#Next, I will check the structure of the data and I will check for any NA or missing values in the data frame.
+## ---- exploredata2 --------
+str(NISTeenVax)
+dplyr::glimpse(NISTeenVax)
+NA_columns <- colSums(is.na(NISTeenVax))
+print(NA_columns)
+
+
+#There are no missing values, but there may be missing data that is filled with a character value within the data frame. I will first find the character value assigned to missing variables by exploring the levels of each variable using the sapply() and then lapply() function.
+
+factor_cols <- sapply(NISTeenVax, is.factor) #Find the columns that are factor variables
+factor_levels <- lapply(NISTeenVax[, factor_cols, drop = FALSE], levels) #Find the levels of each factor variable
+print(factor_levels)#Print factor levels for only the factor columns
+
+#Several columns contain factor levels that are labelled, "Data missing", "Missing data", "Unknown", "Refused", Missing in error", and "Don't know". I will replace these values for each variable with an NA in R so that they will be counted as missing values for analysis.
+## ---- cleandata2 --------
+missing_values <- c("Data Missing", "MISSING Data", "Missing Data", "MISSING IN ERROR") #Group all of the factor levels that indicate a missing value
+
+#After back and forth with ChatGPT3.5, it suggested to use the dplyr mutate function and replace function to specify that factor variables with the levels in the group above will have those level values replaced with NA
+NISTeenVax2 <- NISTeenVax %>% #Create new dataframe
+  mutate(across(where(is.factor), ~ replace(., . %in% missing_values, NA))) #mutate only at factor variables in the dataframe and replace the levels specified above with NA for those factors
+
+#Now I will be able to check for NA values using the skim() function. I can also display the NA values graphically using gg_miss_var()
+
+skimr::skim(NISTeenVax2)
+gg_miss_var(NISTeenVax2)
+
[…]

### 2024-03-14 products/manuscript/Manuscript.qmd
Added to readme files, added images instead of chunks to manuscript

@@
 
-```         
 For the analysis portion, we have decided to do a correlation analysis between HPV vaccination completion and each of the socioeconomic factors.
@@
 Our study will fill in geographical gaps and provide updated vaccination trends based on socioeconomic factors.
-```
 
@@
 
-```{r, include=FALSE, cache=FALSE}
-library(here)
-knitr::read_chunk(here("./code/eda-code/edacode.R"))
-```
+First we loaded the packages needed for the exploratory analysis.
+And then we loaded the data called 'cleandata1' from the 'processed-data' folder.
 
-First we will load the packages needed for the exploratory analysis.
-```{r, packages, message = FALSE, warning = FALSE}
-```
-
-Next, we will load the data called 'cleandata1' from the 'processed-data' folder.
-```{r, loaddata}
-```
-
-The following data set has been cleaned.
-Any observations with NAs has been removed.
-
-```{r, table1}
-
-```
-
+We start our exploratory analysis by displaying the distribution of our predictors, including the socioeconomic variables that we chose for analysis.
 The distribution of the numerical income to poverty ratio can be shown below.
 It displays a value from 1-3.
-A ratio displaying less than 1 depicts an income less than the poverty level.
-Any income ratio \>1 indicates an income greater than the poverty level.
+A ratio displaying less than 1 depicts an income less than the poverty level.Any income ratio \>1 indicates an income greater than the poverty level.
 Most of the observations were gathered from households with income ratios of 3, as shown by the skewed plot below.
+![Figure 1. Income-poverty ratio distribution](results/figures/income.png)
 
-```{r, INCPORAR_I}
-
-```
-
+Nest, the distribution of surveys given per state was analyzed.
 The state variable shows a more uniformed distribution, suggesting that the observations were gathered fairly across the US states and territory.
 Although it is noted a few territories provides 0 observations.
-
-```{r, STATE}
-
-```
+![Figure 2. State dsitribution](results/figures/state_distribution.png)
 
 The family income variable provides a more normal distribution with the majority of the observation coming from \$75000+ income level.
-
-```{r, INCQ298A}
-
-```
+![Figure 3. The family income distribution](results/figures/family.income.distribution.png)
 
@@
 
-```{r, RACEETHK}
-
-```
-
-```{r, RACE_K}
-
-```
+![Figure 4.Racial distribution](results/figures/race.distribution.png) ![Figure 5.Ethnicity distribution](results/figures/ethnicity.distribution.png)
 
@@
 This is mirrored in the facility variable, where a majority of its observations show private facilities being the location.
-
-```{r, INS_STAT2_I}
-
-```
-
-```{r, FACILITY}
-
-```
+![Figure 6.
+Insurance status distribution](results/figures/insurance.status.distribution.png) ![Figure 7.
+Distribution of the medical facility where the survey was administerred](results/figures/facility.distribution.png)
 
 The vaccination variable indicates nearly 80% of the observations had an Up-to-date status for the 1+ shot HPV vaccination.
+We will follow by examining the percentage of up-to-date vaccination status by state.
 
-```{r, Vaccination}
-
-```
-
-We examine the percentage of up-to-date vaccination status by state.
-
-```{r, state-vaccine}
-
-```
+![Figure 8. Vaccine status distribution](results/figures/State-vaccine-completion.png)
 
@@
 
-```{r, facility-vaccine}
-
-```
+![Figure 9.Facility and vaccine completion distribution](results/figures/facility-vaccination.png)
 
@@
 
-```{r, race-ethnicity-vaccine}
-
-```
+![Figure 10. Distribution of vaccine completion stratified by race and ethnicity](results/figures/race-ethnicity-vaccination.png)
 
@@
 
-```{r, fitfig1}
-
-```
+![Figure 11.](figures/results/income-vaccination.png)
 
@@
 
-```{r, fitfig2}
-
-```
+![Figure 12. Income poverty ratio and insurance status](results/figures/insurance-income-stratified.png)
 
 Income-poverty ratio was fit to vaccine status and insurance status in a multi-linear model.
-
-```{r, fitfig3}
-
-```
 
@@
 
-```{r, fitfig4}
-
-```
+![Figure 14. Income-poverty level versus ethnic group](results/figures/raceeth-income-stratified.png)
 

### 2024-03-15 code/analysis-code/analysis1.qmd
Added to manuscript and analysis code document. There is a start in analysis1 qmd.
My computer is freaking out so I am going to try to perform some more diagnostics on the model I got with stepwise comparisons, in the morning.

---
title: "Data Analysis and Modelling of Teen Vaccination Surveys"
date: "`r format(Sys.Date()'"
Author: Kelly Cao and Rachel Robertson
output: html_document
editor: 
  markdown: 
    wrap: sentence
---
First, I will start by setting an equation for the generalized logistic regression model with all of the predictor variables intended for use.
```{r}
levels(mydata$P_UTDHPV)
mydata2 <- droplevels(mydata[!mydata$P_UTDHPV == 'Missing Data', ]) # I will do some brief cleaning and remove all rows with missing values for our response variable, 'P_UTDHPV', to perform a binomial regression
str(mydata2$P_UTDHPV) # Check for removal of missing values
fitall <- glm(P_UTDHPV ~ AGE + SEX + STATE + INS_STAT2_I + INCQ298A + INS_BREAK_I + INCPOV1 + RACEETHK + EDUC1 + LANGUAGE + MOBIL_1 + RENT_OWN + FACILITY, data = mydata2, family = binomial) # define the equation for the binomial logistic regression
```

Now, I will set the start of the step-wise comparison, using a model with an intercept only (or blank model).
```{r}
fitstart <- glm(P_UTDHPV ~ 1, data=mydata2, family = binomial)
```

Now I will start the step-wise comparison in both the forwards and backwards direction to allow the model to add or drop predictors depending on the responding AIC value.
```{r}
step(fitstart, direction = "both", scope = list(lower = ~1, upper = ~ AGE + SEX + STATE + INS_STAT2_I + INCQ298A + INS_BREAK_I + INCPOV1 + RACEETHK + EDUC1 + LANGUAGE + MOBIL_1 + RENT_OWN + FACILITY), data = mydata, family = binomial) # define the lower and upper limits of the step wise comparison 
```
The model chosen includes the predictors STATE, FACILITY, RACEETHK, AGE, EDUC1, INS_BREAK_I, INCPOV1, SEX, LANGUAGE, MOBIL_1, and INS_STAT2_I. 
Because of the extremely high residual value, I will perform some diagnostics to determine what is the issue. First, I will look at a residual plot to check for non linearity. Next, I will use a VIF analysis to check for multicollinearity. I am suspicious of co linearity as many of these socioeconomic predicts might be inherently related.

I consulted ChatGPT to ask how to perform a VIF analysis in R and was recommended the packages 'car' and 'carData' using the vif() function.
I will start by opening the libraries necessary.
```{r}
library(ggplot2)
library(car) # to perform VIF analysis
library(carData) # to look at
```

Next I will produce a residual plot
```{r}
model <- glm(glm(P_UTDHPV ~ AGE + SEX + STATE + INS_STAT2_I + INS_BREAK_I + INCPOV1 + RACEETHK + EDUC1 + LANGUAGE + MOBIL_1 + FACILITY, data = mydata2, family = binomial)
# Residual plot
residualPlot <- function(model) {
  ggplot(data = mydata2(fit = predict(model), residuals = resid(model)),
         aes(x = fit, y = residuals)) +
    geom_point() +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    xlab("Fitted values") +
    ylab("Residuals") +
    ggtitle("Residual Plot")
}

# VIF diagnostics
vifDiagnostics <- function(model) {
  vif_values <- car::vif(model)
  print(vif_values)
}

# Create residual plot
residual_plot <- residualPlot(fitall)
print(residual_plot)

# Calculate VIFs
vif_diagnostics <- vifDiagnostics(model)
print(vif_diagnostics)

```
```

### 2024-03-15 products/manuscript/Manuscript.qmd
Added to manuscript and analysis code document. There is a start in analysis1 qmd.
My computer is freaking out so I am going to try to perform some more diagnostics on the model I got with stepwise comparisons, in the morning.

@@
 
-## Methods
+# Methods
 
-### Data Cleaning and Processing
+## Data Cleaning and Processing
 
-#### Loading the data
+### Loading the data
 
@@
 
-### Exploratory Analysis
+## Exploratory Analysis
 
@@
 
-### Model and Plot Fitting
+## Model and Plot Fitting
 
@@
 {{< pagebreak >}} 
+# Analysis
 
-## References
+We begin with various socioeconomic predictors that interest us.
+Some of these predictors are a subset of the other, or may be used to describe an overall category.
+For this reason
+
+The variables used were chosen based on their interpretability and applicability to our research question “Which socioeconomic and geographic factors impact HPV vaccine completion rates among teenagers?”
+.
+There are seven potential outcome variables for HPV vaccine completion:
+
+![Table 3. Human Papillomavirus vaccine status variables and meanings.](assets/UTDHPV-variable-table.png)
+
+The first option, P_UTDHPV was the chosen outcome as it is the most general.
+It has the benefit of being able to use Age as a predictor variable and also not biasing the model towards older ages who are required to receive more than one shot to be considered up-to-date.
+Additionally, in April of 2022, the SAGE work group of the WHO officially recommended a single-dose HPV vaccine regime, because vaccine efficacy was found to be significantly high with a single dose.
+This recommendation was given with the goal of increasing vaccine coverage in areas with lower vaccine accessibility and availability (4).
+In September of 2023, PAHO followed suit in recommending a single-dose HPV vaccine regime for the Americas (5).
+
+The predictor variables we chose to examine are listed below:
+
+![Table 4. Potential predictor variable names and meanings.](assets/Predictor-variable-table.png)
+
+INCOPAR or income-poverty ratio was excluded because this is reflected by the income and poverty status variables, separately.
+Additionally, this is a numeric variable while the others are categorical.
+RACE_K or race was excluded because race is included within the RACEETHK variable.
+RACEETHK is ideal because it includes both race and ethnicity.
+
+To determine which predictor variables will be best suited for a general linear model, I will do a step-wise comparison and then measure the resulting models with AIC and WAIC.
+
+# References
 
@@
 doi: 10.1136/bmjmed-2023-000726
+4.  https://www.who.int/news/item/20-12-2022-WHO-updates-recommendations-on-HPV-vaccination-schedule
+5.  https://www.paho.org/en/news/5-9-2023-paho-technical-advisory-group-recommends-countries-americas-use-single-dose-hpv

### 2024-03-15 code/analysis-code/analysis1.qmd
Analysis for multicollinearity

@@
 library(car) # to perform VIF analysis
-library(carData) # to look at
+library(carData)
 ```
 
-Next I will produce a residual plot
+Next I will produce a residual plot. I asked ChatGPT how to use ggplot to produce a residual plot for a logistic regression function and was given the base code, which I modified to fit my analysis.
 ```{r}
-model <- glm(glm(P_UTDHPV ~ AGE + SEX + STATE + INS_STAT2_I + INS_BREAK_I + INCPOV1 + RACEETHK + EDUC1 + LANGUAGE + MOBIL_1 + FACILITY, data = mydata2, family = binomial)
-# Residual plot
-residualPlot <- function(model) {
-  ggplot(data = mydata2(fit = predict(model), residuals = resid(model)),
+# First, I will define the model given to me by the step wise comparison
+model <- glm(P_UTDHPV ~ AGE + SEX + STATE + INS_STAT2_I + INS_BREAK_I + INCPOV1 + RACEETHK + EDUC1 + LANGUAGE + MOBIL_1 + FACILITY, data = mydata2, family = binomial)
+
+# Define function for residual plot of logistic regression
+residualPlotLogit <- function(model) {
+  fitted_prob <- predict(model, type = "response") # Predicted probabilities
+  residuals <- resid(model, type = "response") # Calculate residuals
+  plot_data <- data.frame(fit = fitted_prob, residuals = residuals) # Create data frame
+
+  # Plot
+  ggplot(data = plot_data,
          aes(x = fit, y = residuals)) +
@@
     geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
-    xlab("Fitted values") +
+    xlab("Fitted probabilities") +
     ylab("Residuals") +
-    ggtitle("Residual Plot")
+    ggtitle("Residual Plot of Proposed Stepwise Model")
 }
 
-# VIF diagnostics
-vifDiagnostics <- function(model) {
-  vif_values <- car::vif(model)
-  print(vif_values)
-}
-
-# Create residual plot
-residual_plot <- residualPlot(fitall)
-print(residual_plot)
-
-# Calculate VIFs
-vif_diagnostics <- vifDiagnostics(model)
-print(vif_diagnostics)
+# Create the residual plot
+residual_plot_logit <- residualPlotLogit(fitall)
+print(residual_plot_logit)
 
 ```
+The residual plot shows clear collinearity that is not represented by the current model.
+To find the predictors with collinnearity we will use a correlation matrix.
+This is because the VIF analysis will not run when multicollinearity is too high.
+```{r}
+factor_data <- model.matrix(~ STATE + FACILITY + RACEETHK + AGE + EDUC1 + INS_BREAK_I + INCPOV1 + SEX + LANGUAGE + MOBIL_1 + INS_STAT2_I, data = mydata2)
+
+# Compute correlation matrix
+correlation_matrix <- cor(factor_data)
+
+# Find indices of correlation matrix where values are 1 or 0
+indices <- which(!is.na(correlation_matrix) & correlation_matrix == 1 | correlation_matrix == 0, arr.ind = TRUE)
+
+# Extract variable names corresponding to these indices
+variable_pairs <- rownames(correlation_matrix)[indices[,1]]
+variable_pairs <- cbind(variable_pairs, colnames(correlation_matrix)[indices[,2]])
+
+# Print variable pairs with correlation values of 1 or 0
+print(variable_pairs)
+
 ```
+This shows that many levels from each of the variables show multicollineaity. This might be addressed by using a different model for analysis.

### 2024-03-16 code/analysis-code/analysis1.qmd
Added analysis2.qmd, added more to analysis1

@@
 title: "Data Analysis and Modelling of Teen Vaccination Surveys"
-date: "`r format(Sys.Date()'"
+date: '`r format(Sys.Date())`'
 Author: Kelly Cao and Rachel Robertson
@@
 First, I will start by setting an equation for the generalized logistic regression model with all of the predictor variables intended for use.
+
+The model aims to predict the vaccination status (P_UTDHPV) based on various predictor variables like age, sex, state, insurance status, income, race/ethnicity, education, language, etc.
+The response variable P_UTDHPV represents whether a teenager is up-to-date on HPV vaccination.
 ```{r}
+#Path to data. Note the use of the here() package and not absolute paths
+data_location <- here::here("data","processed-data","cleandata1.rds")
+
+#load data
+mydata <- readRDS(data_location)
+
 levels(mydata$P_UTDHPV)
@@
 ```
+
+> A Stepwise model comparison is performed to iteratively add or remove predictor variables based on their significance in improving the model fit.
 
@@
 Now I will start the step-wise comparison in both the forwards and backwards direction to allow the model to add or drop predictors depending on the responding AIC value.
+This is performed with the `step()` function, setting up the scope for the variables to consider.
 ```{r}

### 2024-03-16 code/analysis-code/analysis2.qmd
Added analysis2.qmd, added more to analysis1

---
title: "Data Analysis and Modelling of Teen Vaccination Surveys"
date: '`r format(Sys.Date())`'
Author: Kelly Cao and Rachel Robertson
output: html_document
editor: 
  markdown: 
    wrap: sentence

---
### Introduction
The previous analysis performed shown on analysis1.qmd found in the same directory, showed collinearity among many of the predictor variables. The following script is an attempt to reduce that issue to prevent problem when performing analyses further downstream.

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

#### Loading library and data set
``` {R}
# Load required libraries
library(MASS) # for robust regression
library(caret) # for cross-validation
library(h2o) #For dimension reduction
library(dplyr)

```

I start with loading the same data set and performing the same functions to clean up the data by removing NAs. We will be looking at the variable P_UTDHPV, which reflects the up-to-date HPV vaccination status of the individual.
```{R}
# Load and preprocess data
data_location <- here::here("data","processed-data","cleandata1.rds")
mydata <- readRDS(data_location)

# Remove rows with missing values for the response variable
mydata <- mydata[complete.cases(mydata$P_UTDHPV), ]

```

### PCA Analysis
I then initialize the h2o cluster and converted my data object to a H2oFrame or an h2o object. I then performed PCA using the function `h20.prcomp` and using the newly converted h20 object as a training frame.

This is perform in attempt t oindirectly reduce the collinearity in the predictor variables.

The following lines are based on the resource shown below:

[Link](https://bradleyboehmke.github.io/HOML/pca.html)

>PCA indirectly addresses multicollinearity by transforming the original features into a new set of orthogonal variables, whereas LASSO/Ridge regression directly addresses multicollinearity by penalizing the coefficients of correlated predictors.

```{R}
# Initialize and connect to the H2O cluster
# Start an H2O cluster
h2o.init()

# Convert mydata to H2OFrame
mydata_h2o <- as.h2o(mydata)

# Perform PCA using h2o.prcomp
pca_model <- h2o.prcomp(
  training_frame = mydata_h2o, 
  x = c("AGE", "SEX", "STATE", 
        "INS_STAT2_I", "INCQ298A", 
        "INS_BREAK_I", "INCPOV1", 
        "RACEETHK", "EDUC1", 
        "LANGUAGE", "MOBIL_1", 
        "RENT_OWN", "FACILITY"))

# Viewing the result of the PCA
print(pca_model)

```

The PCA results showed one principle component with a high standard deviation and high proportion of variance, showing that the Pc1 captures a large portion (97.9%) of variability in the data.

The result of having only one principle component further supports the fact that many of the predictor variables are highly collinear. One PC shows that a PCA must reduce the data object to a single dimension. Having a single principal component could be problematic for my analysis, particularly if it leads to the loss of crucial information or structure in my data due to collinearity among the original features.

Understanding why this occurs and assessing whether it aligns with my expectations and the characteristics of my dataset is essential. I may need to potentially review my data preprocessing methods or explore different feature selection techniques.

### LASSO Regression
I first investigate alternative approaches to address collinearity, such as employing regularization methods like LASSO regression. This would ensure that I'm effectively capturing the underlying patterns in my data while mitigating the challenges posed by collinearity.

```{R}
# Use LASSO regression for variable selection and regularization
lasso_model <- h2o.glm(
  x = c("AGE", "SEX", "STATE",
        "INS_STAT2_I", "INCQ298A"
        , "INS_BREAK_I", "INCPOV1",
        "RACEETHK", "EDUC1",
        "LANGUAGE", "MOBIL_1",
        "RENT_OWN", "FACILITY"),
  y = "P_UTDHPV", training_frame = mydata_h2o, family = "binomial", alpha = 1)

# View Lasso Model
print(lasso_model)

```

The model used Lasso to pick out 97 important predictors from 105, which likely dealt with issues of collinearity among the original features. This simplification helped in making the model easier to understand and potentially improved its performance.

### Attempt at Cross-validation
I then utilized ChatGPT to write me a code to perform a cross-validation test.

```{R}
# Create a fold column for cross-validation
fold_column <- h2o.kfold_column(data = mydata_h2o, nfolds = 10, seed = 1)  # Specify the number of folds and a seed for reproducibility

# Add the fold column to the H2OFrame
mydata_h2o <- h2o.cbind(mydata_h2o, fold_column)

# Specify predictor variables
predictor_columns <- c("AGE", "SEX", "STATE", "INS_STAT2_I", "INCQ298A", "INS_BREAK_I", 
                       "INCPOV1", "RACEETHK", "EDUC1", "LANGUAGE", "MOBIL_1", "RENT_OWN", "FACILITY")

# Evaluate model performance using cross-validation
cv_metrics <- h2o.glm(x = predictor_columns, 
                      y = "P_UTDHPV", 
                      training_frame = mydata_h2o, 
                      fold_column = "C1",  # Use the name of the fold column (default name is "C1")
                      family = "binomial")

# Display results
print(cv_metrics)
```

The cross-validation results for the logistic regression model show consistent performance across folds, with an average accuracy of approximately 79.4%. The model demonstrates stability with metrics such as an area under the curve (AUC) around 0.644 and a mean per-class error of about 49.8%. Precision and recall metrics also exhibit robustness, averaging around 79.6% and 99.6%, respectively. These results indicate the model's ability to balance correctly identifying positive cases while minimizing false positives. However, areas for model refinement include residual deviance and specificity.

### Conclusion
our analysis aimed to address collinearity issues in our teen vaccination surveys dataset through dimension reduction techniques like PCA and regularization methods such as LASSO regression. LASSO regression effectively selected important predictors while simplifying the model. Subsequent cross-validation demonstrated consistent model performance, with an average accuracy of 79.4% and robust precision and recall metrics.

### 2024-03-16 products/manuscript/Manuscript.qmd
Added analysis2.qmd, added more to analysis1

@@
 knitr::opts_chunk$set(warning = FALSE, message = FALSE)
+library(here)
 ```
@@
 Most of the observations were gathered from households with income ratios of 3, as shown by the skewed plot below.
-![Figure 1. Income-poverty ratio distribution](results/figures/income.png)
+
+![Figure 1. Income-poverty ratio distribution](/results/figures/income.png)
 
@@
 
-![Figure 4.Racial distribution](results/figures/race.distribution.png) ![Figure 5.Ethnicity distribution](results/figures/ethnicity.distribution.png)
+![Figure 4.Racial distribution](results/figures/race.distribution.png) 
+![Figure 5.Ethnicity distribution](results/figures/ethnicity.distribution.png)
 
@@
 
-![Figure 8. Vaccine status distribution](results/figures/State-vaccine-completion.png)
+![Figure 8. Vaccine status distribution](../results/figures/State-vaccine-completion.png)
 

### 2024-04-02 products/manuscript/Manuscript.qmd
Added to readme and manuscript
Added background and reorganized Intro. Added more info to main repo readme and added the code order.

@@
 # Introduction
+## Background
 
-## Data Set
+Human Papillomavirus (HPV) is a double-stranded DNA virus that infects the cells of the skin and mucosal linings [1].
+HPV is the most common sexually transmitted infection (STI) in the United States, where over 90% of sexually active males and 80% of sexually active females will likely become infected within their lifetime [2, 3].
+Most cases of HPV are asymptomatic and cleared over time, but certain infections tend to become persistent and can cause genital warts.
+Specific types of HPV are known to be “high-risk” or increase the likelihood of developing cancer in persistently infected cells [4].
+The HPV group contains over 200 related viruses, but only 12 types are considered “high-risk” due to the presence of oncogenes.
+These include HPV 16, 18, 31, 33, 35, 39, 45, 51, 52, 56, 58, and 59, though HPV-16 and HPV-18 are associated with the highest number of HPV-associated cancer cases.
+In the U.S., there are 37,300 cases of HPV-associated cancer cases every year, including cases of cervical, anal, oropharyngeal (throat and mouth), vulvar, vaginal, and penile cancers [5,6].
+In an effort to prevent HPV-associated cancers worldwide, two HPV vaccines were developed and made available in 2006.
+These included a bivalent and quadrivalent vaccine that contained viral-like particles (VLPs) of the viral capsid protein (L1) from “high-risk” HPV types [7].
+In recent years, the vaccine used in the U.S., Gardasil 9, has been expanded to contain HPV types 6, 11, 16, 18, 31, 33, 45, 52, and 58 [4].
+Currently, the United States Centers for Disease Control and Prevention (CDC) recommends a two-dose Gardasil 9 regime to all teenagers, with a primary dose at age 11 to 12 and a second dose within 6 to 12 months of the first [8].
+In 2022, the CDC estimated that 76.0% of teenagers aged 13 to 17 had received at least 1 dose of the HPV vaccine while only 62.6% of teenagers were up-to-date with their vaccination regime [9].
+
+## Data
 
@@
 
-We want to use the 2022 National Immunization Survey for teens to identify which socioeconomic and geographic factors are associated with HPV vaccination completion for teenagers in the U.S.
-Specifically, we are asking the question, What is the likelihood of a teenager being up-to-date for their HPV vaccinations in the U.S., based on socioeconomic factors and geographic distribution.
-The outcome of this study will be measured using the columns that pertain to HPV vaccination completion.
-These variables are provider-collected, and they state whether or not the teen is (1) up-to-date with 1+ HPV shot, (2) up-to-date with 2+ HPV shots, or (3) up-to-date with 3+ HPV shots (excluding all vaccinations post-survey).
+Data from the 2022 National Immunization Survey for teens will be used to identify which socioeconomic and geographic factors are associated with HPV vaccination completion for teenagers aged 13 to 17 in the U.S.
+Specifically, we want to find the most important socioeconomic and geographic factors for a teenager being up-to-date for their HPV vaccinations in the U.S.
+The outcome of this study will be measured using the overall up-to-date HPV vaccination completion indicator, which includes all age groups in any step of the vaccination process.
+The variable includes teenagers aged 13 to 17 who are up-to-date and have received at least one HPV shot.
+
+## Purpose
+
 Measuring HPV vaccine completion with “up-to-date” variables may be more robust than vaccination rate based on those who have completed the regime (2 or more shots), because not all teens included in the survey are at the age in which they would’ve completed the full regime.
@@
 This serves as valuable information to determine differences in healthcare access and vaccination coverage for teenagers living in the U.S.
-
-## Proposed Analysis
-
-For the analysis portion, we have decided to do a correlation analysis between HPV vaccination completion and each of the socioeconomic factors.
-When performing this, we may choose to stratify by age or gender, if we find these to be confounders.
-The correlation analysis cannot be performed using a multivariate model because there may be collinearity present between many of the determinants for socioeconomic status.
-This will lead to spurious results.
-Instead, the correlation between vaccination status and each factor is examined separately.
-Next, we may stratify this by state.
-Comparing these correlation values between states and regions may lead us to find patterns of healthcare accessibility in the U.S.
-It is generally known that those of lower income and without insurance have limited healthcare access in the U.S.(1), but it will be interesting to see if this pattern is reflected at the state level.
+It is generally known that those of lower income and without insurance have limited healthcare access in the U.S.
+[10], but it will be interesting to see if this pattern is reflected at the state level.
 If so, policy should be examined as it may be related to healthcare access disparities.
-Recent studies have examined HPV vaccination completion according to social determinants, but they have not examined geographic distribution and include data prior to 2019 (2).
-More recent years display differences in global HPV vaccination trends due to the COVID19 pandemic (3).
+Recent studies have examined HPV vaccination completion according to social determinants, but they have not examined geographic distribution and include data prior to 2019 [11].
+More recent years display differences in global HPV vaccination trends due to the COVID19 pandemic [12].
 Our study will fill in geographical gaps and provide updated vaccination trends based on socioeconomic factors.
@@
 
-First we loaded the packages needed for the exploratory analysis.
-And then we loaded the data called 'cleandata1' from the 'processed-data' folder.
-
-We start our exploratory analysis by displaying the distribution of our predictors, including the socioeconomic variables that we chose for analysis.
+We use our exploratory analysis to determine the distribution of our predictors, including the socioeconomic variables that we chose for analysis.
 The distribution of the numerical income to poverty ratio can be shown below.
@@
 Additionally, in April of 2022, the SAGE work group of the WHO officially recommended a single-dose HPV vaccine regime, because vaccine efficacy was found to be significantly high with a single dose.
-This recommendation was given with the goal of increasing vaccine coverage in areas with lower vaccine accessibility and availability (4).
-In September of 2023, PAHO followed suit in recommending a single-dose HPV vaccine regime for the Americas (5).
+This recommendation was given with the goal of increasing vaccine coverage in areas with lower vaccine accessibility and availability (13).
+In September of 2023, PAHO followed suit in recommending a single-dose HPV vaccine regime for the Americas (14).
 
@@
 
-1.
+# References
+1.	https://www.who.int/teams/health-product-policy-and-standards/standards-and-specifications/vaccine-standardization/human-papillomavirus#:~:text=Human%20papillomavirus%20(HPV)%20is%20a,approximately%208%2Dkb%20in%20length.
+2.	https://www.womenshealth.gov/a-z-topics/human-papillomavirus#:~:text=Genital%20HPV%20is%20the%20most,79%20million%20Americans%20have%20HPV.&text=It%20is%20so%20common%20that,some%20point%20in%20their%20lifetime.
+3.	https://www.health.state.mn.us/data/mcrs/data/qfhpv.html#:~:text=Human%20Papillomavirus%20(HPV)%20is%20a,with%20HPV%20in%20their%20lifetime.&text=Around%2050%20percent%20of%20HPV,HPV%2C%20which%20can%20cause%20cancer.
+4.	HPV and Cancer. In: National Cancer Institute [Internet]. 1 Mar 2019 [cited 12 Mar 2024]. Available: https://www.cancer.gov/about-cancer/causes-prevention/risk/infectious-agents/hpv-and-cancer
+5.	https://www.cancer.gov/about-cancer/causes-prevention/risk/infectious-agents/hpv-and-cancer
+6.	https://www.cdc.gov/vaccines/vpd/hpv/hcp/vaccines.html#:~:text=Vaccine%20Composition,45%2C%2052%2C%20and%2058. 
+7.	https://www.who.int/teams/health-product-policy-and-standards/standards-and-specifications/vaccine-standardization/human-papillomavirus#:~:text=Human%20papillomavirus%20(HPV)%20is%20a,approximately%208%2Dkb%20in%20length.
+8.	https://www.cdc.gov/vaccines/vpd/hpv/hcp/recommendations.html#:~:text=Two%20doses%20of%20HPV%20vaccine,third%20dose%20of%20HPV%20vaccine.
+9.	https://www.cdc.gov/mmwr/volumes/72/wr/mm7234a3.htm#:~:text=In%202022%2C%2089.9%25%20of%20adolescents,HPV%20vaccination%20(HPV%20UTD).
+
+10.
 Sciences, N.
@@
 https://www.ncbi.nlm.nih.gov/books/NBK500097/
-2.
+11.
 Mansfield, L.
@@
 https://doi.org/10.1016/j.ssmph.2022.101082
-3.
+12.
 Casey, R.
@@
 doi: 10.1136/bmjmed-2023-000726
-4.  https://www.who.int/news/item/20-12-2022-WHO-updates-recommendations-on-HPV-vaccination-schedule
-5.  https://www.paho.org/en/news/5-9-2023-paho-technical-advisory-group-recommends-countries-americas-use-single-dose-hpv
+13.  https://www.who.int/news/item/20-12-2022-WHO-updates-recommendations-on-HPV-vaccination-schedule
+14.  https://www.paho.org/en/news/5-9-2023-paho-technical-advisory-group-recommends-countries-americas-use-single-dose-hpv

### 2024-04-02 code/analysis-code/analysis2.qmd
Transferred code from .R to .qmd, updated README
Updated README to reflect changes and transfers

@@
 #### Loading library and data set
-``` {R}
+``` {R, warnings=FALSE, messages=FALSE}
 # Load required libraries

### 2024-04-02 code/eda-code/EDA-code.qmd
Transferred code from .R to .qmd, updated README
Updated README to reflect changes and transfers

@@
 title: "Exploratory Analysis on CDC Teen Vaccination Surveys"
-date: "`r format(Sys.Date()'"
+date: 02/21/2024
 Author: Kelly Cao and Rachel Robertson
@@
 
-## Loading the .R script which performed the analysis
+## Loading the Packages and Data Set
 
-```{r, include=FALSE, cache=FALSE}
-library(here)
-knitr::read_chunk(here("./code/eda-code/edacode.R"))
+```{r, warnings=FALSE, message=FALSE}
+#load needed packages. make sure they are installed.
+library(here) #for data loading/saving
+library(dplyr)
+library(skimr)
+library(ggplot2)
+library(knitr)
 ```
 
-## Loading the Packages
-
-```{r, packages, message = FALSE, warning = FALSE}
-```
-
-```{r, loaddata}
+```{r}
+#Path to data. Note the use of the here() package and not absolute paths
+data_location <- here::here("data","processed-data","cleandata1.rds")
+#load data
+mydata <- readRDS(data_location)
 ```
@@
 
-```{r, table1}
-
+```{r}
+summary_df = skimr::skim(mydata)
+print(summary_df)
+# save to file
+summarytable_file = here("results","tables", "summarytable.rds")
+saveRDS(summary_df, file = summarytable_file)
 ```
@@
 
-```{r, INCPORAR_I}
-
+```{r}
+p1 <- mydata %>% ggplot(aes(x=INCPORAR_I)) + geom_histogram() 
+plot(p1)
+figure_file = here("results", "figures", "income.png")
+ggsave(filename = figure_file, plot=p1) 
 ```
@@
 
-```{r, STATE}
+```{r}
+summary(mydata$STATE)
+state_distribution <- table(mydata$STATE)
+# Convert STATE to character
+mydata$STATE <- as.character(mydata$STATE)
+
+# Calculate distribution percentage
+state_distribution_percentage <- prop.table(table(mydata$STATE))*100
+
+print(state_distribution_percentage)
+
+p2 <- barplot(state_distribution_percentage, 
+        main = "State Distribution Percentage", 
+        xlab = "State", 
+        ylab = "Percentage",
+        col = "skyblue",   # Set bar color
+        las = 2) 
+#Recreating the graph in ggplot
+
+#adjusting the percent disitrbution data into a data frame
+data_df <- as.data.frame(as.table(state_distribution_percentage))
+
+#plot with ggplot
+p3 <- ggplot(data_df, aes(x = Var1, y = Freq)) +
+  geom_bar(stat = "identity", fill = "skyblue") +
+  labs(title = "State Distribution Percentage",
+       x = "State",
+       y = "Percentage") +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
+
+#Viewing the plot to confirm it is done properly
+print(p3)
+
+figure_file = here("results", "figures", "state.distribution.png")
+ggsave(filename = figure_file, plot=p3)
 
@@
 
-```{r, INCQ298A}
+```{r}
+summary(mydata$INCQ298A)
+Family_income_distribution <- table(mydata$INCQ298A)
+
+# Calculate distribution percentage
+Family_Income_distribution <- prop.table(table(mydata$INCQ298A))*100
+
+print(Family_Income_distribution)
+
+p4 <- barplot(Family_Income_distribution, 
+              main = "Family Income Distribution Percentage", 
+              xlab = "Income levels", 
+              ylab = "Percentage",
+              col = "skyblue",   # Set bar color
+              las = 2) +
+      theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
+
+#Recreating the graph in ggplot
+
+#adjusting the percent disitrbution data into a data frame
+family_income_df <- as.data.frame(as.table(Family_Income_distribution))
+
+#plot with ggplot
+p5 <- ggplot(family_income_df, aes(x = Var1, y = Freq)) +
+  geom_bar(stat = "identity", fill = "skyblue") +
+  labs(title = "Family Income Distribution Percentage",
+       x = "Income",
+       y = "Percentage")
+
+#Viewing the plot to confirm it is done properly
+print(p5)
+
+figure_file = here("results", "figures", "family.income.distribution.png")
+ggsave(filename = figure_file, plot=p5)
 
@@
 
-```{r, RACEETHK}
+```{r}
+summary(mydata$RACEETHK)
+ethnicity_distribution <- table(mydata$RACEETHK)
+
+# Calculate distribution percentage
+Ethnicity_distribution <- prop.table(table(mydata$RACEETHK))*100
+
+print(Ethnicity_distribution)
+
+p6 <- barplot(Ethnicity_distribution, 
+              main = "Ethnicity Distribution Percentage", 
+              xlab = "Ethnicity", 
+              ylab = "Percentage",
+              col = "skyblue",   # Set bar color
+              las = 2) +
+theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
+#Recreating the graph in ggplot
+
+#adjusting the percent disitrbution data into a data frame
+ethnicity_df <- as.data.frame(as.table(Ethnicity_distribution))
+
+#plot with ggplot
+p7 <- ggplot(ethnicity_df, aes(x = Var1, y = Freq)) +
+  geom_bar(stat = "identity", fill = "skyblue") +
+  labs(title = "Ethnicity Distribution Percentage",
+       x = "Ethnicity",
+       y = "Percentage")
+
+#Viewing the plot to confirm it is done properly
+print(p7)
+
+figure_file = here("results", "figures", "ethnicity.distribution.png")
+ggsave(filename = figure_file, plot=p7)
 
@@
 
-```{r, RACE_K}
+```{r}
+summary(mydata$RACE_K)
+race_distribution <- table(mydata$RACE_K)
 
+# Calculate distribution percentage
+Race_distribution <- prop.table(table(mydata$RACE_K))*100
+
+print(Race_distribution)
+
+p8 <- barplot(Race_distribution, 
+              main = "Race Distribution Percentage", 
+              xlab = "Race", 
+              ylab = "Percentage",
+              col = "skyblue",   # Set bar color
+              las = 2) 
+#Recreating the graph in ggplot
+
+#adjusting the percent disitrbution data into a data frame
+race_df <- as.data.frame(as.table(Race_distribution))
+
+#plot with ggplot
+p9 <- ggplot(race_df, aes(x = Var1, y = Freq)) +
+  geom_bar(stat = "identity", fill = "skyblue") +
+  labs(title = "Race Distribution Percentage",
+       x = "Race",
+       y = "Percentage")
+
+#Viewing the plot to confirm it is done properly
+print(p9)
+
+figure_file = here("results", "figures", "race.distribution.png")
+ggsave(filename = figure_file, plot=p9)
 ```
@@
 
-```{r, INS_STAT2_I}
+```{r}
+summary(mydata$INS_STAT2_I)
+insurance_status_distribution <- table(mydata$INS_STAT2_I)
 
+# Calculate distribution percentage
+Insurance_distribution <- prop.table(table(mydata$INS_STAT2_I))*100
+
+print(Insurance_distribution)
+
+p10 <- barplot(Insurance_distribution, 
+              main = "Insurance Status Distribution Percentage", 
+              xlab = "Insurance Status", 
+              ylab = "Percentage",
+              col = "skyblue",   # Set bar color
+              las = 2) 
+#Recreating the graph in ggplot
+
+#adjusting the percent disitrbution data into a data frame
+insurance_df <- as.data.frame(as.table(Insurance_distribution))
+
+#plot with ggplot
+p11 <- ggplot(insurance_df, aes(x = Var1, y = Freq)) +
+  geom_bar(stat = "identity", fill = "skyblue") +
+  labs(title = "Insurance Status Distribution Percentage",
+       x = "Insurance Status",
+       y = "Percentage")
+
+#Viewing the plot to confirm it is done properly
+print(p11)
+
+figure_file = here("results", "figures", "insurance.status.distribution.png")
+ggsave(filename = figure_file, plot=p11) 
 ```
 
-```{r, FACILITY}
+```{r}
+summary(mydata$FACILITY)
+Facility_distribution <- table(mydata$FACILITY)
+
+# Calculate distribution percentage
+Facility_distribution <- prop.table(table(mydata$FACILITY))*100
+
+print(Facility_distribution)
+
+p12 <- barplot(Facility_distribution, 
+              main = "Facility Distribution Percentage", 
+              xlab = "Facility", 
+              ylab = "Percentage",
+              col = "skyblue",   # Set bar color
+              las = 2) +
+    theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
+#Recreating the graph in ggplot
+
+#adjusting the percent disitrbution data into a data frame
+facility_df <- as.data.frame(as.table(Facility_distribution))
+
+#plot with ggplot
+p13 <- ggplot(facility_df, aes(x = Var1, y = Freq)) +
+  geom_bar(stat = "identity", fill = "skyblue") +
+  labs(title = "Facility Distribution Percentage",
+       x = "Facility",
+       y = "Percentage") +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
+
+#Viewing the plot to confirm it is done properly
+print(p13)
+
+figure_file = here("results", "figures", "facility.distribution.png")
+ggsave(filename = figure_file, plot=p13)
 
@@
 
-```{r, Vaccination}
+```{r}
+summary(mydata$P_UTDHPV)
+Vaccination_status_distribution <- table(mydata$P_UTDHPV)
+
+# Calculate distribution percentage
+Vaccination_status_distribution <- prop.table(table(mydata$P_UTDHPV))*100
+
+print(Vaccination_status_distribution)
+
+p14 <- barplot(Vaccination_status_distribution, 
+              main = "Vaccination Status Distribution Percentage", 
+              xlab = "Vaccination Status", 
+              ylab = "Percentage",
+              col = "skyblue",   # Set bar color
+              las = 2) 
+#Recreating the graph in ggplot
+
+#adjusting the percent disitrbution data into a data frame
+Vaccine_df <- as.data.frame(as.table(Vaccination_status_distribution))
+
+#plot with ggplot
+p15 <- ggplot(Vaccine_df, aes(x = Var1, y = Freq)) +
+  geom_bar(stat = "identity", fill = "skyblue") +
+  labs(title = "Vaccination Status Distribution Percentage",
+       x = "Vaccination Status",
+       y = "Percentage")
+
+#Viewing the plot to confirm it is done properly
+print(p15)
+
+figure_file = here("results", "figures", "vaccination.status.distribution.png")
+ggsave(filename = figure_file, plot=p15) 
+```
+
+We examine the percentage of up-to-date vaccination status by state.
+```{r}
+#Aggregate vaccine status stratified by state
+vaccine_state_percent <- mydata %>%
+  group_by(STATE) %>%
+  summarise(percentage_UTD = sum(P_UTDHPV == "UTD") / n() * 100)
+
+#Create table to display only up-to-date status in percents for each state
+table1 <- mydata %>%
+  filter(P_UTDHPV == "UTD") %>%
+  count(STATE) %>%
+  rename(UTD_count = n) %>%
+  left_join(vaccine_state_percent, by = "STATE") %>%
+  select(STATE, percentage_UTD) %>%
+  arrange(STATE)
+
+#Print the table
+kable(table1, caption = "Vaccine Distribution Percentage per State (Only 'UTD' Status)")
+
+#Write the table to save to the results
+write.table(table1, file = here("results","tables", "state-vaccine-completion.txt"), sep = "\t", row.names = FALSE)
 
@@
 
-We examine the percentage of up-to-date vaccination status by state.
-```{r, state-vaccine}
+```{r}
+#Stratifying the HPV vaccine completion by facility
+##Make a stacked bar plot to compare which facilities reported UTD patients
+p18 <- ggplot(mydata, aes(x = P_UTDHPV, fill = FACILITY)) +
+  geom_bar(stat = "count") +  # Use geom_bar() with the count of each facility
+  labs(x = "Vaccination Status", y = "Count", title = "Facility Vaccination Completion") +
+  theme_minimal()
+  print(p18)
+
+#Save the bar plot
+figure_file <- here("results", "figures", "facility-vaccination.png")
+ggsave(filename = figure_file, plot = p18, width = 8, height = 6, dpi = 300)
 
@@
 
-```{r, facility-vaccine}
+```{r}
+#Aggregate total count per each race-ethnicity factor level
+total_count_by_race <- mydata %>%
+  group_by(RACEETHK) %>%
+  summarise(total_count = n())
 
-```
+#Aggregate count of those with UTD status for each race-enthnicity factor level
+UTD_by_race <- mydata %>%
+  filter(P_UTDHPV == "UTD") %>%
+  group_by(RACEETHK) %>%
+  summarise(UTD_count = n())
 
-```{r, race-ethnicity-vaccine}
+#Merge the total count and UTD count data frames
+vaccine_percent_by_race <- total_count_by_race %>%
+  left_join(UTD_by_race, by = "RACEETHK") %>%
+  mutate(percentage_UTD = UTD_count / total_count * 100)
+
+# Print the table
+kable(vaccine_percent_by_race, caption = "Percentage of UTD Vaccination Status by RACEETHK")
+
+# Save the table as a text file
+write.table(vaccine_percent_by_race, file = here("results", "tables", "vaccine_percent_by_race.txt"), sep = "\t", row.names = FALSE)
 
@@
 
-```{r, fitfig1}
+```{r}
+#Stratifying the Income-poverty levels to the vaccination status in a boxplot figure
+p16 <- mydata %>% ggplot(aes(x=P_UTDHPV, y=INCPORAR_I)) + geom_boxplot()
+plot(p16)
+figure_file = here("results","figures", "income-vaccination.png")
+ggsave(filename = figure_file, plot=p16)
+
+# Fitting Income-poverty levels and Vaccination status to a statistical model
+model <- lm(INCPORAR_I ~ P_UTDHPV, data = mydata)
+summary(model)
 
@@
 
-```{r, fitfig2}
+```{r}
+#Stratifying the Income-poverty levels to the insurance status in a boxplot figure
+original_labels <- levels(mydata$INS_STAT2_I)
+print(original_labels)
 
+p17 <- mydata %>% 
+  ggplot(aes(x=INS_STAT2_I, y=INCPORAR_I, color = P_UTDHPV)) +  
+  geom_boxplot() +
+  labs(x = "Insurance status", y = "Income-poverty ratio", title = "Income-poverty ratio stratified by insurance and vaccination") +
+  scale_x_discrete(labels = c("PRIVATE INSURANCE ONLY" = "Private", "ANY MEDICAID" = "Medicaid", "OTHER INSURANCE (CHIP, IHS, MILITARY, OR OTHER, ALONE OR IN COMB.
+WITH PRIVATE INSURANCE)" = "Other", "UNINSURED" = "Uninsured")) +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))  # Rotate x-axis text by 90 degrees
+plot(p17)
+figure_file = here("results","figures", "insurance-income-stratified.png")
+ggsave(filename = figure_file, plot=p17)
+
+# Fitting Income-poverty levels and insurance status to a statistical model
+model <- lm(INCPORAR_I ~ INS_STAT2_I, data = mydata)
+summary(model)
 ```
@@
 ### Fit 3
-```{r, fitfig3}
-
+```{r}
+# Fitting Income-poverty levels and insurance status to a statistical model
+model <- lm(INCPORAR_I ~ P_UTDHPV, INS_STAT2_I, data = mydata)
+summary(model)
 ```
@@
 ### Fit 4
-```{r, fitfig4}
+```{r}
+#Stratifying the HPV vaccine completion by both Race-Ethnicity and (choose either insurance or income-poverty status)
+p19 <- mydata %>% 
+  ggplot(aes(x=INCPORAR_I, y=RACEETHK, color = P_UTDHPV)) +  geom_boxplot()
+plot(p19)
+figure_file = here("results","figures", "raceeth-income-stratified.png")
+ggsave(filename = figure_file, plot=p19)
+
+# Fitting Income-poverty levels and insurance status to a statistical model
+model <- lm(INCPORAR_I ~ P_UTDHPV, RACEETHK, data = mydata)
+summary(model)
 

### 2024-04-02 code/processing-code/Processing-Quarto-Output.qmd
Transferred code from .R to .qmd, updated README
Updated README to reflect changes and transfers

@@
 # Setup
-
-This chunk pulls the R-script entitled Creating_Data_Frame.R.
-
-```{r, include=FALSE, cache=FALSE}
-library(here)
-knitr::read_chunk(here("./code/processing-code/creating-df/Creating_Data_Frame.R"))
-```
-
-Load needed packages.
-
-```{r, packages, message = FALSE, warning = FALSE}
+I first load the needed packages.
+```{r, message = FALSE, warning = FALSE}
+library(readr) #Loading the package to read the fwf
+library(here) #Loading the package to produce relative pathway
+library(knitr) #Loading in the package for dynamic report generation
+library(tidyverse)
+library(readxl)
+library(dplyr)
 ```
@@
 # Data loading
+I load the raw .DAT file into the environment and using the `readr` package, I declare the column position and labels of the desired columns of interest.
+I then apply the declared object with `read_fwf` function to the raw data (.DAT) file into a data frame named `data_frame`.
+```{r}
+column_position<- fwf_positions(
+  start = c(1, 89, 287, 295, 314, 330, 331, 334, 336, 343, 344, 345,
+            347, 353, 359, 362, 472, 473, 495, 496, 497, 502, 504, 1296, 1297
+  ),
+  end = c(5, 92, 288, 295, 329, 330, 332, 334, 336, 343, 344, 346,
+          347, 354, 359, 362, 472, 473, 495, 496, 497, 502, 504, 1296, 1297
+  ),
+  col_names = c("SEQNUMT", "YEAR", "AGE", "EDUC1", "INCPORAR_I", 
+                "INCPOV1", "INCQ298A", "LANGUAGE", "MOBIL_1", "RACEETHK", 
+                "RACE_K", "RENT_OWN", "SEX", "STATE", "FACILITY", "WELLCHILD", 
+                "P_U13HPV", "P_U13HPV3", "P_UTDHPV", "P_UTDHPV_15", 
+                "P_UTDHPV_15INT", "P_UTDHPV2", "P_UTDHPV3", "INS_STAT2_I", 
+                "INS_BREAK_I")
+)
+file_path <- here("data", "raw-data", "NISTEENPUF22.DAT")
 
-```{r, loaddata}
+data_frame <- read_fwf(file_path, 
+                       col_positions = column_position,
+                       col_types = cols(.default = "c")
+                       )
 ```
@@
 # Specifying Factor Variables
+With the data_frame in the environment, I nthen apply the labels for each factor levels using the provided codebook (this codebook can be found in ./data/raw_data/NIS-Teen_Puff22-CODEBOOK.pdf).
+This will make further downstream analysis much easier to work with.
+```{r}
+EDUC1levels=c(1,2,3,4)
+EDUC1labels=c("LESS THAN 12 YEARS", "12 YEARS", "MORE THAN 12 YEARS, NON-COLLEGE GRAD", "COLLEGE GRADUATE")
 
-```{r, cleandata}
+INCPOVlevels=c(1,2,3,4)
+INCPOVlabels=c("ABOVE POVERTY > $75K", "ABOVE POVERTY <= $75K", "BELOW POVERTY", "UNKNOWN")
+
+INCQ298Alevels=c(10,11,12,13,14,3,4,5,6,7,77,8,9,99)
+INCQ298Alabels=c("$35001 - $40000", "$40001 - $50000", "$50001 - $60000", "$60001 - $75000", "$75001+", "$0 - $7500", "$7501 - $10000", "$10001 - $17500", "$17501 - $20000", "$20001 - $25000", "DON'T KNOW", "$25001 - $30000", "$30001 - $35000", "REFUSED")
+
+LANGUAGElevels=c(1,2,3)
+LANGUAGElabels=c("ENGLISH", "SPANISH", "OTHER")
+
+MOBILlevels=c(1,2,77,98,99)
+MOBILlabels=c("MOVED FROM DIFFERENT STATE", "DID NOT MOVE FROM DIFFERENT STATE", "DON'T KNOW", "MISSING IN ERROR", "REFUSED")
+
+RACEETHKlevels=c(1,2,3,4)
+RACEETHKlabels=c("HISPANIC", "NON-HISPANIC WHITE ONLY", "NON-HISPANIC BLACK ONLY", "NON-HISPANIC OTHER + MULTIPLE RACE")
+
+RACE_Klevels=c(1,2,3)
+RACE_Klabels=c("WHITE ONLY", "BLACK ONLY", "OTHER + MULTIPLE RACE")
+
+RENTOWNlevels=c(1,2,3,77,99)
+RENTOWNlabels=c("OWNED OR BEING BOUGHT", "RENTED", "OTHER ARRANGMENT", "DON'T KNOW", "REFUSED")
+
+SEXlevels=c(1,2,77,98,99)
+SEXlabels=c("MALE", "FEMALE", "DON'T KNOW", "MISSING IN ERROR", "REFUSED")
+
+STATElevels=c(1,10,11,12,13,15,16,17,18,19,2,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,4,40,41,42,44,45,46,47,48,49,5,50,51,53,54,55,56,6,66,72,78,8,9)
+STATElabels=c("ALABAMA", "DELAWARE", "DISTRICT OF COLUMBIA", "FLORIDA", "GEORGIA", "HAWAII", "IDAHO", "ILLINOIS", "INDIANA", "IOWA", "ALASKA", "KANSAS", "KENTUCKY", "LOUISIANA", "MAINE", "MARYLAND", "MASSACHUSETTS", "MICHIGAN", "MINNESOTA", "MISSISSIPPI", "MISSOURI", "MONTANA", "NEBRASKA", "NEVADA",
+              "NEW HAMPSHIRE", "NEW JERSEY", "NEW MEXICO", "NEW YORK", "NORTH CAROLINA", "NORTH DAKOTA", "OHIO", "ARIZONA", "OKLAHOMA", "OREGON", "PENNSYLVANIA", "RHODE ISLAND", "SOUTH CAROLINA", "SOUTH DAKOTA", "TENNESSEE", "TEXAS", "UTAH", "ARKANSAS", "VERMONT", "VIRGINIA", "WASHINGTON", "WEST VIRGINIA",
+              "WISCONSIN", "WYOMING", "CALIFORNIA", "GUAM", "PUERTO RICO", "U.S. VIRGIN ISLANDS", "COLORADO", "CONNECTICUT")
+
+FACILITYlevels=c(1,2,3,4,5,6, ".")
+FACILITYlabels=c("ALL PUBLIC FACILITIES", "ALL HOSPITAL FACILITIES", "ALL PRIVATE FACILITIES", "ALL STD/SCHOOL/TEEN CLINICS OR OTHER FACILITIES", "MIXED", "UNKNOWN", "Missing Data")
+
+WELLCHILDlevels=c(1,2,3)
+WELLCHILDlabels=c("YES", "NO", "DON'T KNOW")
+
+UTDlevels=c(0,1, ".")
+UTDlabels=c("NOT UTD", "UTD", "Missing Data")
+
+INS_STAT2_Ilevels=c(1,2,3,4, ".")
+INS_STAT2_Ilabels=c("PRIVATE INSURANCE ONLY", "ANY MEDICAID", "OTHER INSURANCE (CHIP, IHS, MILITARY, OR OTHER, ALONE OR IN COMB. WITH PRIVATE INSURANCE)", "UNINSURED", "MISSING Data")
+
+INS_BREAK_Ilevels=c(1,2,3,4,".")
+INS_BREAK_Ilabels=c("CURRENTLY INSURED BUT UNINSURED AT SOME POINT SINCE AGE 11", "CURRENTLY INSURED AND NEVER UNINSURED SINCE AGE 11", "CURRENTLY UNINSURED BUT INSURED AT SOME POINT SINCE AGE 11", "CURRENTLY UNINSURED AND NEVER INSURED SINCE AGE 11", "Data Missing")
+
+data<-mutate(data_frame,
+             SEQNUMT = as.numeric(SEQNUMT), 
+             YEAR = as.numeric(YEAR),
+             AGE = as.numeric(AGE),
+             EDUC1 = factor(EDUC1,levels = EDUC1levels, labels = EDUC1labels),
+             INCPORAR_I = as.numeric(INCPORAR_I), 
+             INCPOV1 = factor(INCPOV1, levels = INCPOVlevels, labels = INCPOVlabels),
+             INCQ298A = factor(INCQ298A, levels = INCQ298Alevels, labels = INCQ298Alabels),             
+             LANGUAGE = factor(LANGUAGE, levels = LANGUAGElevels, labels = LANGUAGElabels),
+             MOBIL_1 = factor(MOBIL_1, levels = MOBILlevels, labels = MOBILlabels),
+             RACEETHK = factor(RACEETHK, levels = RACEETHKlevels, labels = RACEETHKlabels),
+             RACE_K = factor(RACE_K, levels = RACE_Klevels, labels = RACE_Klabels),
+             RENT_OWN = factor(RENT_OWN, levels = RENTOWNlevels, labels = RENTOWNlabels), 
+             SEX = factor(SEX, levels = SEXlevels, labels = SEXlabels),
+             STATE = factor(STATE, levels = STATElevels, labels = STATElabels), 
+             FACILITY = factor(FACILITY, levels = FACILITYlevels, labels = FACILITYlabels),
+             WELLCHILD = factor(WELLCHILD, levels = WELLCHILDlevels, labels = WELLCHILDlabels),
+             P_U13HPV = factor(P_U13HPV, levels = UTDlevels, labels = UTDlabels),
+             P_U13HPV3 = factor(P_U13HPV3, levels = UTDlevels, labels = UTDlabels),
+             P_UTDHPV = factor(P_UTDHPV, levels = UTDlevels, labels = UTDlabels),
+             P_UTDHPV_15 = factor(P_UTDHPV_15, levels = UTDlevels, labels = UTDlabels),
+             P_UTDHPV_15INT = factor(P_UTDHPV_15INT, levels = UTDlevels, labels = UTDlabels),
+             P_UTDHPV2 = factor(P_UTDHPV2, levels = UTDlevels, labels = UTDlabels),
+             P_UTDHPV3 = factor(P_UTDHPV3, levels = UTDlevels, labels = UTDlabels),
+             INS_STAT2_I = factor(INS_STAT2_I, levels = INS_STAT2_Ilevels, labels = INS_STAT2_Ilabels),
+             INS_BREAK_I = factor(INS_BREAK_I, levels = INS_BREAK_Ilevels, labels = INS_BREAK_Ilabels)
+)
+
+summary(data)
+str(data)
 
@@
 # Save data
-
-```{r, savedata}
+We then save the data frame into a .rds and a csv file named "NIS_Teen_Data_2022.csv" and "processeddata.rds"
+```{r}
+output_file<- here("data", "processed-data", "NIS_Teen_Data_2022.csv")
+write.csv(data, file = output_file, row.names = FALSE)
+save_path <- here("./data/processed-data/processeddata.rds")
+saveRDS(data, file = save_path )
 ```
@@
 # Cleaning Data
-
-```{r, include=FALSE, cache=FALSE}
+```{r, warning=FALSE, message=FALSE}
 library(here)
-knitr::read_chunk(here("./code/processing-code/cleaningdata/cleaning1.R"))
+library(dplyr)
+library(skimr)
+library(ggplot2)
+library(tidyr)
+library(naniar)
 ```
 
-```{r, packages2}
-```
-
-```{r, loaddata2}
+First, I load the data into a usable data frame.
+```{r}
+data_location <- here::here("data","processed-data","processeddata.rds")
+NISTeenVax <- readRDS(data_location)
 ```
@@
 ## Exploring the data
+Next, I will check the structure of the data and I will check for any NA or missing values in the data frame.
+```{r}
+str(NISTeenVax)
+dplyr::glimpse(NISTeenVax)
+NA_columns <- colSums(is.na(NISTeenVax))
+print(NA_columns)
 
-```{r, exploredata2}
+#There are no missing values, but there may be missing data that is filled with a character value within the data frame.
+I will first find the character value assigned to missing variables by exploring the levels of each variable using the sapply() and then lapply() function.
+
+factor_cols <- sapply(NISTeenVax, is.factor) #Find the columns that are factor variables
+factor_levels <- lapply(NISTeenVax[, factor_cols, drop = FALSE], levels) #Find the levels of each factor variable
+print(factor_levels)#Print factor levels for only the factor columns
+
+#Several columns contain factor levels that are labelled, "Data missing", "Missing data", "Unknown", "Refused", Missing in error", and "Don't know".
+I will replace these values for each variable with an NA in R so that they will be counted as missing values for analysis.
 ```
@@
 ## Eliminating NA Values and Columns
+Several columns contain factor levels that are labelled, "Data missing", "Missing data", "Unknown", "Refused", Missing in error", and "Don't know".
+I will replace these values for each variable with an NA in R so that they will be counted as missing values for analysis.
+```{r}
+missing_values <- c("Data Missing", "MISSING Data", "Missing Data", "MISSING IN ERROR") #Group all of the factor levels that indicate a missing value
 
-```{r, cleandata2}
+#After back and forth with ChatGPT3.5, it suggested to use the dplyr mutate function and replace function to specify that factor variables with the levels in the group above will have those level values replaced with NA
+NISTeenVax2 <- NISTeenVax %>% #Create new dataframe
+  mutate(across(where(is.factor), ~ replace(., . %in% missing_values, NA))) #mutate only at factor variables in the dataframe and replace the levels specified above with NA for those factors
+
+#Now I will be able to check for NA values using the skim() function. I can also display the NA values graphically using gg_miss_var()
+
+skimr::skim(NISTeenVax2)
+gg_miss_var(NISTeenVax2)
+
+#The most important factors for this analysis is the response variables called variations of "P_UTDHPV".
+#We have 27,024 missing values for each of these columns.
+These missing values may have come from the same survey type, however analysis cannot be performed when this crucial data is missing.
+For this reason, I will find the rows where these missing values take place, ensure that they are all the same rows for the multiple variables, and delete these rows specifically.
+#I used chatGPT3,5 to find the function to select for certain rows using dplyr and then filter these for NA values
+##I prompted it by using the variable names a,b,c and asked for it to select these columns and filter for the rows with missing variables
+rows_with_missing_in_all <- NISTeenVax2 %>%
+  select("P_U13HPV", "P_U13HPV3", "P_UTDHPV", "P_UTDHPV_15", "P_UTDHPV_15INT", "P_UTDHPV2", "P_UTDHPV3") %>% #select for columns in question
+  filter_all(any_vars(is.na(.))) #Find rows with NA values
+
+##I then prompted ChatGPT3.5 to provide me with code to display the location of these rows so that I may check if they all matched.
+if (nrow(rows_with_missing_in_all) > 0) {
+  print("Rows with missing values in aforementioned variables")
+  print(rows_with_missing_in_all)
+} else {
+  print("No rows with missing values in aforementioned variables")
+}
+
+#It seems that all of the missing rows match for all of these variables. This means that we may delete the rows that contain missing values for all of the specified variables.
+#Because the variables listed all contain missing values in the same rows, I will select only one variable to filter.
+
+NISTeenVax3 <- NISTeenVax2 %>%
+  filter(!is.na(P_UTDHPV)) #drop the rows with NA values in P_UTDHPV
+NISTeenVax3 <- NISTeenVax3 %>%
+  filter(!is.na(INS_BREAK_I))
+NISTeenVax3 <- NISTeenVax3 %>%
+  filter(!is.na(FACILITY))
+
+skimr::skim(NISTeenVax3) #check again for missing values in these variables
+gg_miss_var(NISTeenVax3)
+skimr::skim(NISTeenVax3) #check again for missing values in these variables
+gg_miss_var(NISTeenVax3)
+
+#The majority of data is missing from the WELLCHILD column, which indicates whether the teen had receieved a Well Child visit in recent years.
+Because this is not crucial to out analysis, I will delete this column.
+#The year column in unnecessary, since all values are from the year 2022. So, I will also delete the YEAR column.
+
+NISTeenVax4 <- NISTeenVax3 %>% #create a new data frame for the filtered data
+ select(-WELLCHILD, -YEAR) #Use the select() filter to delete multiple columns from the data frame
+str(NISTeenVax4) #Check the remaining variables
+
 ```
@@
 ## Saving the Data Frame
+Save cleaned data as an RDS file.
+```{r}
+output_file2<- here("data", "processed-data", "NIS_Teen_Data_2022_clean1.csv")
+write.csv(NISTeenVax4, file = output_file2, row.names = FALSE)
+saved_pathway <- here("./data/processed-data/cleandata1.rds")
+saveRDS(NISTeenVax4, file = saved_pathway )
 
-```{r, savedata2}
 ```

### 2024-04-02 products/manuscript/Manuscript.qmd
Transferred code from .R to .qmd, updated README
Updated README to reflect changes and transfers

@@
 # Introduction
+
 ## Background
 
-Human Papillomavirus (HPV) is a double-stranded DNA virus that infects the cells of the skin and mucosal linings [1].
-HPV is the most common sexually transmitted infection (STI) in the United States, where over 90% of sexually active males and 80% of sexually active females will likely become infected within their lifetime [2, 3].
+Human Papillomavirus (HPV) is a double-stranded DNA virus that infects the cells of the skin and mucosal linings \[1\].
+HPV is the most common sexually transmitted infection (STI) in the United States, where over 90% of sexually active males and 80% of sexually active females will likely become infected within their lifetime \[2, 3\].
 Most cases of HPV are asymptomatic and cleared over time, but certain infections tend to become persistent and can cause genital warts.
-Specific types of HPV are known to be “high-risk” or increase the likelihood of developing cancer in persistently infected cells [4].
+Specific types of HPV are known to be “high-risk” or increase the likelihood of developing cancer in persistently infected cells \[4\].
 The HPV group contains over 200 related viruses, but only 12 types are considered “high-risk” due to the presence of oncogenes.
 These include HPV 16, 18, 31, 33, 35, 39, 45, 51, 52, 56, 58, and 59, though HPV-16 and HPV-18 are associated with the highest number of HPV-associated cancer cases.
-In the U.S., there are 37,300 cases of HPV-associated cancer cases every year, including cases of cervical, anal, oropharyngeal (throat and mouth), vulvar, vaginal, and penile cancers [5,6].
+In the U.S., there are 37,300 cases of HPV-associated cancer cases every year, including cases of cervical, anal, oropharyngeal (throat and mouth), vulvar, vaginal, and penile cancers \[5,6\].
 In an effort to prevent HPV-associated cancers worldwide, two HPV vaccines were developed and made available in 2006.
-These included a bivalent and quadrivalent vaccine that contained viral-like particles (VLPs) of the viral capsid protein (L1) from “high-risk” HPV types [7].
-In recent years, the vaccine used in the U.S., Gardasil 9, has been expanded to contain HPV types 6, 11, 16, 18, 31, 33, 45, 52, and 58 [4].
-Currently, the United States Centers for Disease Control and Prevention (CDC) recommends a two-dose Gardasil 9 regime to all teenagers, with a primary dose at age 11 to 12 and a second dose within 6 to 12 months of the first [8].
-In 2022, the CDC estimated that 76.0% of teenagers aged 13 to 17 had received at least 1 dose of the HPV vaccine while only 62.6% of teenagers were up-to-date with their vaccination regime [9].
+These included a bivalent and quadrivalent vaccine that contained viral-like particles (VLPs) of the viral capsid protein (L1) from “high-risk” HPV types \[7\].
+In recent years, the vaccine used in the U.S., Gardasil 9, has been expanded to contain HPV types 6, 11, 16, 18, 31, 33, 45, 52, and 58 \[4\].
+Currently, the United States Centers for Disease Control and Prevention (CDC) recommends a two-dose Gardasil 9 regime to all teenagers, with a primary dose at age 11 to 12 and a second dose within 6 to 12 months of the first \[8\].
+In 2022, the CDC estimated that 76.0% of teenagers aged 13 to 17 had received at least 1 dose of the HPV vaccine while only 62.6% of teenagers were up-to-date with their vaccination regime \[9\].
 
@@
 It is generally known that those of lower income and without insurance have limited healthcare access in the U.S.
-[10], but it will be interesting to see if this pattern is reflected at the state level.
+\[10\], but it will be interesting to see if this pattern is reflected at the state level.
 If so, policy should be examined as it may be related to healthcare access disparities.
-Recent studies have examined HPV vaccination completion according to social determinants, but they have not examined geographic distribution and include data prior to 2019 [11].
-More recent years display differences in global HPV vaccination trends due to the COVID19 pandemic [12].
+Recent studies have examined HPV vaccination completion according to social determinants, but they have not examined geographic distribution and include data prior to 2019 \[11\].
+More recent years display differences in global HPV vaccination trends due to the COVID19 pandemic \[12\].
 Our study will fill in geographical gaps and provide updated vaccination trends based on socioeconomic factors.
@@
 
-![Figure 1. Income-poverty ratio distribution](/results/figures/income.png)
+![Figure 1. Income-poverty ratio distribution](../results/figures/income.png)
 
@@
 
-![Figure 4.Racial distribution](results/figures/race.distribution.png) 
-![Figure 5.Ethnicity distribution](results/figures/ethnicity.distribution.png)
+![Figure 4.Racial distribution](results/figures/race.distribution.png) ![Figure 5.Ethnicity distribution](results/figures/ethnicity.distribution.png)
 
@@
 
-{{< pagebreak >}} 
-# Analysis
+{{< pagebreak >}} \# Analysis
 
@@
 # References
-1.	https://www.who.int/teams/health-product-policy-and-standards/standards-and-specifications/vaccine-standardization/human-papillomavirus#:~:text=Human%20papillomavirus%20(HPV)%20is%20a,approximately%208%2Dkb%20in%20length.
-2.	https://www.womenshealth.gov/a-z-topics/human-papillomavirus#:~:text=Genital%20HPV%20is%20the%20most,79%20million%20Americans%20have%20HPV.&text=It%20is%20so%20common%20that,some%20point%20in%20their%20lifetime.
-3.	https://www.health.state.mn.us/data/mcrs/data/qfhpv.html#:~:text=Human%20Papillomavirus%20(HPV)%20is%20a,with%20HPV%20in%20their%20lifetime.&text=Around%2050%20percent%20of%20HPV,HPV%2C%20which%20can%20cause%20cancer.
-4.	HPV and Cancer. In: National Cancer Institute [Internet]. 1 Mar 2019 [cited 12 Mar 2024]. Available: https://www.cancer.gov/about-cancer/causes-prevention/risk/infectious-agents/hpv-and-cancer
+
+1.
+https://www.who.int/teams/health-product-policy-and-standards/standards-and-specifications/vaccine-standardization/human-papillomavirus#:\~:text=Human%20papillomavirus%20(HPV)%20is%20a,approximately%208%2Dkb%20in%20length.
+
+2.
+https://www.womenshealth.gov/a-z-topics/human-papillomavirus#:\~:text=Genital%20HPV%20is%20the%20most,79%20million%20Americans%20have%20HPV.&text=It%20is%20so%20common%20that,some%20point%20in%20their%20lifetime.
+
+3.
+https://www.health.state.mn.us/data/mcrs/data/qfhpv.html#:\~:text=Human%20Papillomavirus%20(HPV)%20is%20a,with%20HPV%20in%20their%20lifetime.&text=Around%2050%20percent%20of%20HPV,HPV%2C%20which%20can%20cause%20cancer.
+
+4.  HPV and Cancer.
+    In: National Cancer Institute \[Internet\].
+    1 Mar 2019 \[cited 12 Mar 2024\].
+    Available: https://www.cancer.gov/about-cancer/causes-prevention/risk/infectious-agents/hpv-and-cancer
+
 5.  https://www.cancer.gov/about-cancer/causes-prevention/risk/infectious-agents/hpv-and-cancer
-6.	https://www.cdc.gov/vaccines/vpd/hpv/hcp/vaccines.html#:~:text=Vaccine%20Composition,45%2C%2052%2C%20and%2058. 
-7.	https://www.who.int/teams/health-product-policy-and-standards/standards-and-specifications/vaccine-standardization/human-papillomavirus#:~:text=Human%20papillomavirus%20(HPV)%20is%20a,approximately%208%2Dkb%20in%20length.
-8.	https://www.cdc.gov/vaccines/vpd/hpv/hcp/recommendations.html#:~:text=Two%20doses%20of%20HPV%20vaccine,third%20dose%20of%20HPV%20vaccine.
-9.	https://www.cdc.gov/mmwr/volumes/72/wr/mm7234a3.htm#:~:text=In%202022%2C%2089.9%25%20of%20adolescents,HPV%20vaccination%20(HPV%20UTD).
 
-10.
-Sciences, N.
-A.
+6.  https://www.cdc.gov/vaccines/vpd/hpv/hcp/vaccines.html#:\~:text=Vaccine%20Composition,45%2C%2052%2C%20and%2058.
+
+7.
+https://www.who.int/teams/health-product-policy-and-standards/standards-and-specifications/vaccine-standardization/human-papillomavirus#:\~:text=Human%20papillomavirus%20(HPV)%20is%20a,approximately%208%2Dkb%20in%20length.
+
+8.  https://www.cdc.gov/vaccines/vpd/hpv/hcp/recommendations.html#:\~:text=Two%20doses%20of%20HPV%20vaccine,third%20dose%20of%20HPV%20vaccine.
+
+9.  https://www.cdc.gov/mmwr/volumes/72/wr/mm7234a3.htm#:\~:text=In%202022%2C%2089.9%25%20of%20adolescents,HPV%20vaccination%20(HPV%20UTD).
+
+10. Sciences, N. A.
     of, Engineering, & Medicine, and.
@@
     https://www.ncbi.nlm.nih.gov/books/NBK500097/
-11.
-Mansfield, L.
-N., Chung, R.
-J., Silva, S.
-G., Merwin, E.
-I., & Gonzalez-Guarda, R.
-M.
+
+11. Mansfield, L. N., Chung, R. J., Silva, S. G., Merwin, E. I., & Gonzalez-Guarda, R. M.
     (2022).
-Social determinants of human papillomavirus vaccine series completion among U.S.
-adolescents: A mixed-methods study.
+    Social determinants of human papillomavirus vaccine series completion among U.S. adolescents: A mixed-methods study.
     SSM - population health, 18, 101082.
     https://doi.org/10.1016/j.ssmph.2022.101082
-12.
-Casey, R.
-M., Akaba, H., Hyde, T.
-B, et al.
-(2024).
+
+12. Casey, R. M., Akaba, H., Hyde, T. B, et al. (2024).
     Covid-19 pandemic and equity of global human papillomavirus vaccination: descriptive study of World Health Organization-Unicef vaccination coverage estimates.
@@
     doi: 10.1136/bmjmed-2023-000726
+
 13. https://www.who.int/news/item/20-12-2022-WHO-updates-recommendations-on-HPV-vaccination-schedule
+
 14. https://www.paho.org/en/news/5-9-2023-paho-technical-advisory-group-recommends-countries-americas-use-single-dose-hpv

### 2024-04-04 products/manuscript/Manuscript.qmd
Images and Summary Table
Fixed images in manuscript and added summary table in EDA (work in progress).

@@
 We use our exploratory analysis to determine the distribution of our predictors, including the socioeconomic variables that we chose for analysis.
+The overall distribution of each factor variable is shown below.
+![Table 1.]
 The distribution of the numerical income to poverty ratio can be shown below.
@@
 
-![Figure 1. Income-poverty ratio distribution](../results/figures/income.png)
+![Figure 1. Income-poverty ratio distribution](./images/income.png)
 
@@
 Although it is noted a few territories provides 0 observations.
-![Figure 2. State dsitribution](results/figures/state_distribution.png)
+![Figure 2. State distribution](./images/state.distribution.png)
 
 The family income variable provides a more normal distribution with the majority of the observation coming from \$75000+ income level.
-![Figure 3. The family income distribution](results/figures/family.income.distribution.png)
+![Figure 3. The family income distribution](./images/family.income.distribution.png)
 
@@
 
-![Figure 4.Racial distribution](results/figures/race.distribution.png) ![Figure 5.Ethnicity distribution](results/figures/ethnicity.distribution.png)
+![Figure 4.Racial distribution](./images/race.distribution.png)
+
+![Figure 5.Ethnicity distribution](./images/ethnicity.distribution.png)
 
@@
 ![Figure 6.
-Insurance status distribution](results/figures/insurance.status.distribution.png) ![Figure 7.
-Distribution of the medical facility where the survey was administerred](results/figures/facility.distribution.png)
+Insurance status distribution](./images/insurance.status.distribution.png) ![Figure 7.
+Distribution of the medical facility where the survey was administerred](./images/facility.distribution.png)
 
@@
 
-![Figure 8. Vaccine status distribution](../results/figures/State-vaccine-completion.png)
+![Table 2. Vaccine status distribution](./images/state-vaccine-completion.png)
 
@@
 
-![Figure 9.Facility and vaccine completion distribution](results/figures/facility-vaccination.png)
+![Figure 8.Facility and vaccine completion distribution](./images/facility-vaccination.png)
 
@@
 
-![Figure 10. Distribution of vaccine completion stratified by race and ethnicity](results/figures/race-ethnicity-vaccination.png)
+![Figure 9. Distribution of vaccine completion stratified by race and ethnicity](./images/raceeth-income-stratified.png)
 
@@
 
-![Figure 11.](figures/results/income-vaccination.png)
+![Figure 10.](./images/income-vaccination.png)
 
@@
 
-![Figure 12. Income poverty ratio and insurance status](results/figures/insurance-income-stratified.png)
+![Figure 11. Income poverty ratio and insurance status](./images/insurance-income-stratified.png)
 
@@
 
-![Figure 14. Income-poverty level versus ethnic group](results/figures/raceeth-income-stratified.png)
+![Figure 12. Income-poverty level versus ethnic group](./images/raceeth-income-stratified.png)
 
-{{< pagebreak >}} \# Analysis
+{{< pagebreak >}}
+
+# Analysis
 

### 2024-04-04 code/eda-code/edacode.R

@@ -5,6 +5,10 @@ library(dplyr)
 library(skimr)
 library(ggplot2)
 library(knitr)
+library(kableExtra)
+library(webshot2)
+library(tidyverse)
+library(gtsummary)
 
 ## ---- loaddata --------
 #Path to data. Note the use of the here() package and not absolute paths
@@ -19,11 +23,40 @@ print(summary_df)
 summarytable_file = here("results","tables", "summarytable.rds")
 saveRDS(summary_df, file = summarytable_file)
 
+# Select columns for a summary table
+table_vars <- mydata %>%
+  select(P_UTDHPV, AGE, SEX, INS_STAT2_I, INCPOV1, RACEETHK, EDUC1, LANGUAGE, RENT_OWN)%>%
+  filter(SEX %in% c("MALE", "FEMALE"))%>%
+  filter(INS_STAT2_I %in% c("PRIVATE INSURANCE ONLY", "ANY MEDICAID", "OTHER INSURANCE (CHIP, IHS, MILITARY, OR OTHER, ALONE OR IN COMB. WITH PRIVATE INSURANCE)", "UNINSURED"))
+
+# Create Table 1 (study sample characteristics) from the data
+table1 <- table_vars %>% tbl_summary(by = P_UTDHPV, 
+                                 statistic = list(all_categorical() ~ "{n} ({p}%)"),
+                                 label = c(AGE ~ "Age (years)", 
+                                           SEX ~ "Sex",   
+                                           INS_STAT2_I ~ "Insurance Status", 
+                                           INCPOV1 ~ "Poverty Status",      
+                                           RACEETHK ~ "Race and Ethnicity", 
+                                           EDUC1 ~ "Maternal Education-level",
+                                           LANGUAGE ~ "Survey Language",
+                                           RENT_OWN ~ "Housing Status"))%>%
+  modify_caption("Table 1. Socioeconomic characteristics of U.S. teenagers who have completed or are up-to-date with HPV vaccination series.")
+
+table1
+
+# Save the summary table
+saveRDS(table1, file = here("results/tables/table1.rds"))
+
+
+
 ## ---- INCPORAR_I --------
 p1 <- mydata %>% ggplot(aes(x=INCPORAR_I)) + geom_histogram() 
 plot(p1)
 figure_file = here("results", "figures", "income.png")
+figure_image = here("products", "manuscript", "images", "income.png")
[…]

### 2024-04-04 code/analysis-code/analysis3.qmd
Demographics table and lasso
Demographics table was added to the manuscript and the first round of lasso was fitted. However, for some reason the estimate for all of the predictors chosen by LASSO were equal to zero. I am not sure if this is because the penalty is too high or why else. I will try and research this.

---
title: "Data Analysis and Modelling of Teen Vaccination Surveys"
date: '`r format(Sys.Date())`'
Author: Kelly Cao and Rachel Robertson
output: html_document
editor: 
  markdown: 
    wrap: sentence
---
# LASSO using Tidymodels
We will start by openning the libraries that we will need for creating a LASSO regression model for the data
```{r}
library(tidymodels) # use tidymodels framework
library(ggplot2) # producing visual displays of data
library(dplyr) # manipulating and cleaning data
library(here) # making relative pathways
library(glmnet) # for LASSO regression engine
library(doParallel) # for parallel processing
library(rsample) # for cross validation
```

Now, we will load the preprocessed data.
```{r}
# Load and preprocess data
data_location <- here::here("data","processed-data","cleandata1.rds")
mydata <- readRDS(data_location)

# Remove rows with missing values
mydata$SEX <- droplevels(mydata$SEX, exclude = c("DON'T KNOW", "MISSING IN ERROR", "REFUSED"))
mydata$INS_STAT2_I <- droplevels(mydata$INS_STAT2_I, exclude = "MISSING Data")
mydata$STATE <- droplevels(mydata$STATE, exclude = "Missing Data")
mydata$MOBIL_1 <- droplevels(mydata$MOBIL_1, exclude = c("DON'T KNOW", "MISSING IN ERROR", "REFUSED"))
mydata$FACILTY <- droplevels(mydata$FACILITY, exclude = "Missing Data")
mydata$P_UTDHPV <- droplevels(mydata$P_UTDHPV, exclude = "Missing Data")
```
Next, we will split the data into training and testing groups so that the LASSO regression has a comparison group.
```{r}
# Split data into training and testing datasets
set.seed(123) # seed for reproducibility
split_data <- initial_split(mydata, prop = 0.8) # 80% split for training/testing data
train_data <- training(split_data)
test_data <- testing(split_data)
```
Next, we will specify the LASSO recipe, or the regression formula. This will contain the predictors of interest used for previous analysis.
```{r}
# Create recipe containing predictors of interest
lasso_rec <- recipe( P_UTDHPV ~ AGE + SEX + STATE + INS_STAT2_I + INCQ298A + INS_BREAK_I + INCPOV1 + RACEETHK + EDUC1 + LANGUAGE + MOBIL_1 + RENT_OWN + FACILITY, data = train_data) %>%
  step_dummy(all_nominal(), -all_outcomes())  %>%
  step_rm(all_outcomes(), -all_outcomes())

# Prepare the recipe
lasso_prep <- lasso_rec %>%
  prep()
```
Now I will impliment the function. This involves setting the specific LASSO regression engine, creating a workflow, and then extracting the fit of the model.
```{r}
# Specify logistic regression model
lasso_spec <- logistic_reg(penalty = 0.05, mixture = 1) %>%
  set_engine("glmnet")

# Create workflow
lasso_wf <- workflow() %>%
  add_recipe(lasso_rec)

# Fit the model using training data
lasso_fit <- lasso_wf %>%
  add_model(lasso_spec) %>%
  fit(data = train_data)

# Extract and tidy model parameters
lasso_fit %>%
  extract_fit_parsnip() %>%
  tidy()
```

Hi Kelly, I am confused because only the intercept has an estimate, while every other factor is zero. I wonder if maybe the penalty is too high? I am not sure how to fix this.

# Tuning LASSO model

### 2024-04-04 products/manuscript/Manuscript.qmd
Demographics table and lasso
Demographics table was added to the manuscript and the first round of lasso was fitted. However, for some reason the estimate for all of the predictors chosen by LASSO were equal to zero. I am not sure if this is because the penalty is too high or why else. I will try and research this.

@@
 ```{r setup, include=FALSE}
+library(here)
+library(knitr)
 knitr::opts_chunk$set(warning = FALSE, message = FALSE)
-library(here)
 ```
@@
 The overall distribution of each factor variable is shown below.
-![Table 1.]
+
+```{r}
+#| label: Table1
+#| tbl-cap: "Table 1.Socioeconomic characteristics of U.S. teenagers who have completed or are up-to-date with HPV vaccination series."
+#| echo: FALSE
+Table1 <- readRDS(here("results/tables/table1.rds"))
+kable(Table1)
+```
+```
+
 The distribution of the numerical income to poverty ratio can be shown below.
@@
 
-Nest, the distribution of surveys given per state was analyzed.
+Next, the distribution of surveys given per state was analyzed.
 The state variable shows a more uniformed distribution, suggesting that the observations were gathered fairly across the US states and territory.

### 2024-04-04 code/eda-code/edacode.R

@@ -25,23 +25,25 @@ saveRDS(summary_df, file = summarytable_file)
 
 # Select columns for a summary table
 table_vars <- mydata %>%
-  select(P_UTDHPV, AGE, SEX, INS_STAT2_I, INCPOV1, RACEETHK, EDUC1, LANGUAGE, RENT_OWN)%>%
-  filter(SEX %in% c("MALE", "FEMALE"))%>%
-  filter(INS_STAT2_I %in% c("PRIVATE INSURANCE ONLY", "ANY MEDICAID", "OTHER INSURANCE (CHIP, IHS, MILITARY, OR OTHER, ALONE OR IN COMB. WITH PRIVATE INSURANCE)", "UNINSURED"))
+  select(P_UTDHPV, AGE, SEX, INS_STAT2_I, INCPOV1, RACEETHK, EDUC1, LANGUAGE, RENT_OWN)
+table_vars$SEX <- droplevels(table_vars$SEX, exclude = c("DON'T KNOW", "MISSING IN ERROR", "REFUSED"))
+table_vars$INS_STAT2_I <- droplevels(table_vars$INS_STAT2_I, exclude = "MISSING Data")
 
 # Create Table 1 (study sample characteristics) from the data
-table1 <- table_vars %>% tbl_summary(by = P_UTDHPV, 
-                                 statistic = list(all_categorical() ~ "{n} ({p}%)"),
-                                 label = c(AGE ~ "Age (years)", 
-                                           SEX ~ "Sex",   
-                                           INS_STAT2_I ~ "Insurance Status", 
-                                           INCPOV1 ~ "Poverty Status",      
-                                           RACEETHK ~ "Race and Ethnicity", 
-                                           EDUC1 ~ "Maternal Education-level",
-                                           LANGUAGE ~ "Survey Language",
-                                           RENT_OWN ~ "Housing Status"))%>%
+table1 <- table_vars %>% 
+  tbl_summary(by = P_UTDHPV,  
+              statistic = list(all_categorical() ~ "{n} ({p}%)"),
+              label = c(AGE ~ "Age (years)", 
+                        SEX ~ "Sex",   
+                        INS_STAT2_I ~ "Insurance Status", 
+                        INCPOV1 ~ "Poverty Status",      
+                        RACEETHK ~ "Race and Ethnicity", 
+                        EDUC1 ~ "Maternal Education-level",
+                        LANGUAGE ~ "Survey Language",
[…]

### 2024-04-05 code/analysis-code/analysis3.qmd
Finished analysis 3

@@
 ---
-title: "Data Analysis and Modelling of Teen Vaccination Surveys"
+title: "Data Analysis and Modeling of Teen Vaccination Surveys"
 date: '`r format(Sys.Date())`'
@@
 # LASSO using Tidymodels
-We will start by openning the libraries that we will need for creating a LASSO regression model for the data
+```{r setup, include=FALSE} 
+knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
+```
+We will start by opening the libraries that we will need for creating a LASSO regression model for the data
 ```{r}
@@
 library(rsample) # for cross validation
+library(yardstick) # for metrics
+library(mlbench)
+
 ```
@@
 ```
-Next, we will specify the LASSO recipe, or the regression formula. This will contain the predictors of interest used for previous analysis.
+Next, we will specify the LASSO recipe, or the regression formula. This will contain the predictors of interest used for previous analysis and `P_UTDHPV` as the outcome. 
 ```{r}
@@
 lasso_rec <- recipe( P_UTDHPV ~ AGE + SEX + STATE + INS_STAT2_I + INCQ298A + INS_BREAK_I + INCPOV1 + RACEETHK + EDUC1 + LANGUAGE + MOBIL_1 + RENT_OWN + FACILITY, data = train_data) %>%
-  step_dummy(all_nominal(), -all_outcomes())  %>%
-  step_rm(all_outcomes(), -all_outcomes())
+  step_dummy(all_nominal(), -all_outcomes())  %>% # convert categorical variables (nominal) into dummy variables
+  step_rm(all_outcomes(), -all_outcomes()) #remove the outcome variables from the dataset
 
@@
 ```
-Now I will impliment the function. This involves setting the specific LASSO regression engine, creating a workflow, and then extracting the fit of the model.
+Now I will implement the function. This involves setting the specific LASSO regression engine, creating a workflow, and then extracting the fit of the model.
 ```{r}
@@
   tidy()
+
+# Extract and tidy model parameters
+lasso_coef <- lasso_fit %>%
+  extract_fit_parsnip() %>%
+  tidy()
+
+# Filter out predictors with non-zero coefficients
+lasso_nonzero_coef <- lasso_coef %>%
+  filter(estimate != 0)
+
+# View the remaining predictors with non-zero coefficients
+num_observations <- nrow(lasso_nonzero_coef)
+print(num_observations)
+
 ```
 
-Hi Kelly, I am confused because only the intercept has an estimate, while every other factor is zero. I wonder if maybe the penalty is too high? I am not sure how to fix this.
+It seems like the LASSO model is penalizing the coefficients too heavily, resulting in most of them being shrunk to zero.
+We are going to try lowering the penalty parameter (penalty) to see if it is effective in solving our problem.
+
+```{r}
+# Specify logistic regression model
+lasso_spec <- logistic_reg(penalty = 0.005, mixture = 1) %>%
+  set_engine("glmnet")
+
+# Create workflow
+lasso_wf <- workflow() %>%
+  add_recipe(lasso_rec)
+
+# Fit the model using training data
+lasso_fit <- lasso_wf %>%
+  add_model(lasso_spec) %>%
+  fit(data = train_data)
+
+# Extract and tidy model parameters
+lasso_fit %>%
+  extract_fit_parsnip() %>%
+  tidy()
+
+# Extract and tidy model parameters
+lasso_coef <- lasso_fit %>%
+  extract_fit_parsnip() %>%
+  tidy()
+
+# Filter out predictors with non-zero coefficients
+lasso_nonzero_coef <- lasso_coef %>%
+  filter(estimate != 0)
+
+# View the remaining predictors with non-zero coefficients
+num_observations <- nrow(lasso_nonzero_coef)
+print(num_observations)
+
+```
+
+```{r}
+# Generate predictions
+predictions <- lasso_fit %>%
+  predict(new_data = mydata) %>%
+  bind_cols(mydata)
+
+# Calculate ROC AUC
+predictions %>%
+  mutate(predicted_numeric = as.numeric(.pred_class)) %>%
+  roc_auc(P_UTDHPV, predicted_numeric)
+
+# Generate predictions
+predictions2 <- lasso_fit %>%
+  predict(new_data = train_data) %>%
+  bind_cols(train_data)
+
+# Calculate ROC AUC
+predictions2 %>%
+  mutate(predicted_numeric = as.numeric(.pred_class)) %>%
+  roc_auc(P_UTDHPV, predicted_numeric)
+
+```
+When attempting to determine how lowering the penalty effected the ability of the model, we produced some predictions with both the original data set and training data, both performed fairly poorly, with a ROC_AUC score close to 0.5, suggesting the the model created predictions that were comparable to random.
+
+To determine the ideal penalty, we are going to tune the model.
 
 # Tuning LASSO model
+Any penalty above 0.05 resulted in all of the predictor coefficient reduced to zero. When performing the following tuning, I set the penalty grid from 0.001 to 0.05. 
+```{r}
+# Define a grid of penalty values to tune over
+penalty_grid <- expand.grid(penalty = seq(0.001, 0.05, length.out = 20))
+
+# Set up parallel processing for faster tuning
+registerDoParallel()
+
+# LASSO model specification
+lasso_model_tuned <- 
+  logistic_reg(penalty = tune(), mixture = 1) %>%
+  set_engine("glmnet") %>%
+  set_mode("classification")  # Ensure correct mode for classification problem
+
+# LASSO Workflow
+lasso_workflow_tuned <- 
+  workflow() %>%
+  add_model(lasso_model_tuned) %>%
+  add_recipe(lasso_rec)
+
+# Set up the grid for tuning
+lasso_tuned <- 
+  lasso_workflow_tuned %>%
+  tune_grid(
+    resamples = vfold_cv(train_data, v = 5, repeats = 5), 
+    grid = penalty_grid,
+    control = control_grid(verbose = TRUE)
+  )
+
+# View tuning results
+tuned_plot <- autoplot(lasso_tuned)
+tuned_plot + ggtitle( "Tuned LASSO Model")
+
+# Collect tuning results
+tuning_results <- lasso_tuned %>%
+  collect_metrics()
+
+# Filter for ROC AUC metric
+roc_auc <- tuning_results %>%
+  filter(.metric == "roc_auc")
+
+# View the ROC AUC values
+print(roc_auc)
+
+```
+
+When viewing the metrics of the tuning, we can see that the regularization is directly correlated with the reduction in roc_auc and accuracy and the increase in brier_Class.
+The accuracy is later shown to start increasing suggesting underfitting starts occurring around 0.01.
+Just reviewing the roc_auc alone, the best performance occurs when the penalty is at its lowest.

### 2024-04-05 code/analysis-code/analysis2.5.qmd
Random decision tree and improving LASSO
RF tree is in analysis 2.5 document and LASSO improved at bottom of analysis3 document. Maybe cross validation tuning and refitting, like we did in class, would make this model the best. RF tree did not turn out well, but I did not use it to make predictions yet.

---
title: "Data Analysis and Modeling of Teen Vaccination Surveys"
date: '`r format(Sys.Date())`'
Author: Kelly Cao and Rachel Robertson
output: html_document
editor: 
  markdown: 
    wrap: sentence
---
# Random Forest Model
```{r}
library(tidymodels) # use tidymodels framework.
library(ggplot2) # producing visual displays of data
library(dplyr) # manipulating and cleaning data
library(here) # making relative pathways
library(randomForest) # making random forest model
library(doParallel) # for parallel processing
library(rsample) # for cross validation
```

```{r}
# Load and preprocess data
data_location <- here::here("data","processed-data","cleandata1.rds")
mydata <- readRDS(data_location)

# Remove rows with missing values
mydata$SEX <- droplevels(mydata$SEX, exclude = c("DON'T KNOW", "MISSING IN ERROR", "REFUSED"))
mydata$INS_STAT2_I <- droplevels(mydata$INS_STAT2_I, exclude = "MISSING Data")
mydata$STATE <- droplevels(mydata$STATE, exclude = "Missing Data")
mydata$MOBIL_1 <- droplevels(mydata$MOBIL_1, exclude = c("DON'T KNOW", "MISSING IN ERROR", "REFUSED"))
mydata$FACILTY <- droplevels(mydata$FACILITY, exclude = "Missing Data")
mydata$P_UTDHPV <- droplevels(mydata$P_UTDHPV, exclude = "Missing Data")
```

```{r}
# Split data into training and testing datasets
set.seed(123) # seed for reproducibility
split_data <- initial_split(mydata, prop = 0.8) # 80% split for training/testing data
train_data <- training(split_data)
test_data <- testing(split_data)
```

```{r}
rf_rec <- recipe(P_UTDHPV ~ AGE + SEX + STATE + INS_STAT2_I + INCQ298A + INS_BREAK_I + INCPOV1 + RACEETHK + EDUC1 + LANGUAGE + MOBIL_1 + RENT_OWN + FACILITY, data = train_data) # use full model as recipe for the random forest model

rf_model <- rand_forest()%>% # use rand_forest() to make a random forest model
  set_engine("randomForest", seed = 123)%>%
  set_mode("classification")

rf_workflow <- workflow() %>% # create workflow for rf model
  add_recipe(rf_rec)%>% # apply recipe
  add_model(rf_model) # apply model

rf_fit <- rf_workflow%>%
  fit(data = test_data)%>% # use the workflow to fit the rf model to the data
print(rf_fit)
```

```{r}

tune_spec <- 
  rand_forest(
    mtry = tune(), # parameters of random forest model to tune are the mtry, trees, and min_n
    trees = 50,
    min_n = tune()) %>% 
  set_engine("randomForest", seed = 123) %>% # make sure to set seed again for the tree
  set_mode("classification")

## I used ChatGPT to help me define the tree grid for the parameters mtry and min_n
tree_grid <- grid_regular(mtry(range = c(1, 4)), 
                          min_n(range = c(1, 5)),
                          levels = 6)

rf_tune_wf <- workflow() %>% # create workflow
  add_recipe(rf_rec) %>% # add recipe from earlier
  add_model(tune_spec) # add spec that is defined above

folds_data <- vfold_cv(test_data, v=5 )
# Set the number of cores to use
num_cores <- detectCores() - 1

# Initialize parallel backend
doParallel::registerDoParallel(cores = num_cores)

set.seed(123) # set seed for reproducibility

rf_res <- rf_tune_wf %>%
  tune_grid(
    resamples = folds_data, # use data from the CV folds
    grid = tree_grid)
rf_res %>%
  autoplot()

# Stop parallel processing
stopImplicitCluster()
```

### 2024-04-05 code/analysis-code/analysis3.qmd
Random decision tree and improving LASSO
RF tree is in analysis 2.5 document and LASSO improved at bottom of analysis3 document. Maybe cross validation tuning and refitting, like we did in class, would make this model the best. RF tree did not turn out well, but I did not use it to make predictions yet.

@@
 # Create recipe containing predictors of interest
-lasso_rec <- recipe( P_UTDHPV ~ AGE + SEX + STATE + INS_STAT2_I + INCQ298A + INS_BREAK_I + INCPOV1 + RACEETHK + EDUC1 + LANGUAGE + MOBIL_1 + RENT_OWN + FACILITY, data = train_data) %>%
+lasso_rec <- recipe(P_UTDHPV ~ AGE + SEX + STATE + INS_STAT2_I + INCQ298A + INS_BREAK_I + INCPOV1 + RACEETHK + EDUC1 + LANGUAGE + MOBIL_1 + RENT_OWN + FACILITY, data = train_data) %>%
   step_dummy(all_nominal(), -all_outcomes())  %>% # convert categorical variables (nominal) into dummy variables
@@
 # Specify logistic regression model
-lasso_spec <- logistic_reg(penalty = 0.005, mixture = 1) %>%
+lasso_spec2 <- logistic_reg(penalty = 0.005, mixture = 1) %>%
   set_engine("glmnet")
 
-# Create workflow
-lasso_wf <- workflow() %>%
-  add_recipe(lasso_rec)
-
 # Fit the model using training data
-lasso_fit <- lasso_wf %>%
-  add_model(lasso_spec) %>%
+lasso_fit2 <- lasso_wf %>%
+  add_model(lasso_spec2) %>%
   fit(data = train_data)
@@
 # Extract and tidy model parameters
-lasso_fit %>%
+lasso_fit2 %>%
   extract_fit_parsnip() %>%
@@
 # Extract and tidy model parameters
-lasso_coef <- lasso_fit %>%
+lasso_coef <- lasso_fit2 %>%
   extract_fit_parsnip() %>%
@@
 ```{r}
-# Generate predictions
+# Make predictions for the first LASSO model
 predictions <- lasso_fit %>%
-  predict(new_data = mydata) %>%
-  bind_cols(mydata)
+  augment(new_data = test_data)
 
 # Calculate ROC AUC
-predictions %>%
-  mutate(predicted_numeric = as.numeric(.pred_class)) %>%
-  roc_auc(P_UTDHPV, predicted_numeric)
+roc_auc <- predictions %>%
+  roc_auc(truth = P_UTDHPV, .pred_UTD)
+print(roc_auc)
 
-# Generate predictions
-predictions2 <- lasso_fit %>%
-  predict(new_data = train_data) %>%
-  bind_cols(train_data)
+# Make predictions for the LASSO model with lower penalty
+predictions2 <- lasso_fit2 %>%
+  augment(new_data = test_data)
 
 # Calculate ROC AUC
-predictions2 %>%
-  mutate(predicted_numeric = as.numeric(.pred_class)) %>%
-  roc_auc(P_UTDHPV, predicted_numeric)
+roc_auc <- predictions2 %>%
+  roc_auc(truth = P_UTDHPV, .pred_UTD)
+print(roc_auc)
 
 ```
-When attempting to determine how lowering the penalty effected the ability of the model, we produced some predictions with both the original data set and training data, both performed fairly poorly, with a ROC_AUC score close to 0.5, suggesting the the model created predictions that were comparable to random.
+When attempting to determine how lowering the penalty effected the ability of the model, we produced some predictions for the model with a higher penalty and the model with a very low penalty.
+Lowering the penalty from 0.05 to 0.005 improved the ROC AUC to from 0.5 to 0.37.
+This suggests that the second model is more informative.
+However, to improve the model further, we should find the ideal penalty parameter.
 
@@
 Just reviewing the roc_auc alone, the best performance occurs when the penalty is at its lowest.
+
+(**Rachel redundantly added code that does the same thing. We can choose our favorite of the two next time we meet.**)
+I will now tune the penalty parameter using cross validation to find the penalty with the best roc_auc.
+```{r}
+lasso_penalty <- logistic_reg(penalty = tune(), mixture = 1) %>%
+  set_engine("glmnet")
+
+folds <- rsample::vfold_cv(train_data, v = 5)
+tune_grid <- tibble(penalty = 10^seq(-2, -1, length.out = 10))
+
+wf <- workflow() %>%
+  add_recipe(lasso_rec)%>%
+  add_model(lasso_penalty)
+
+tune_wf <- wf %>% 
+  tune_grid(resamples = folds,
+            grid = tune_grid,
+            control = control_grid(verbose = FALSE, save_pred = TRUE),
+            metrics = metric_set(roc_auc))
+
+# Select the best model based on ROC AUC
+best_mod <- tune_wf %>% 
+  select_best(metric = "roc_auc")
+
+final_wf <- finalize_workflow(wf, best_mod)
+print(final_wf)
+```
+
+# LASSO Re-fit
+Now, we will run the LASSO regression again, but with the tuned penalty parameter of 0.01. 
+```{r}
+# Specify logistic regression model
+lasso_spec3 <- logistic_reg(penalty = 0.01, mixture = 1) %>%
+  set_engine("glmnet")
+
+# Fit the model using training data
+lasso_fit3 <- lasso_wf %>%
+  add_model(lasso_spec3) %>%
+  fit(data = train_data)
+
+# Extract and tidy model parameters
+lasso_fit3 %>%
+  extract_fit_parsnip() %>%
+  tidy()
+
+# Extract and tidy model parameters
+lasso_coef <- lasso_fit3 %>%
+  extract_fit_parsnip() %>%
+  tidy()
+
+# Filter out predictors with non-zero coefficients
+lasso_nonzero_coef <- lasso_coef %>%
+  filter(estimate != 0)
+
+# View the remaining predictors with non-zero coefficients
+num_observations <- nrow(lasso_nonzero_coef)
+print(num_observations)
+
+# Make predictions for ROC_AUC with the tuned model
+predictions3 <- lasso_fit3 %>%
+  augment(new_data = test_data)
+
+# Calculate ROC AUC
+roc_auc <- predictions3 %>%
+  roc_auc(truth = P_UTDHPV, .pred_UTD)
+print(roc_auc)
+```
+The ROC_AUC for this model with a penalty parameter of 0.01 is now 0.39.
+This is slightly worse than the model with a penalty of 0.05.
+However, the model with a lower penalty is more "unstable", so doing cross validation for the ROC_AUC may reveal which model is truly ideal.
+
+# Cross Validation
+Now we will compare the two lasso models with the penalty parameters of 0.005 and 0.01 to determine which model performs best when presented with "new data".

### 2024-04-05 code/analysis-code/analysis4.qmd
Random decision tree and improving LASSO
RF tree is in analysis 2.5 document and LASSO improved at bottom of analysis3 document. Maybe cross validation tuning and refitting, like we did in class, would make this model the best. RF tree did not turn out well, but I did not use it to make predictions yet.

---
title: "Data Analysis and Modeling of Teen Vaccination Surveys"
date: '`r format(Sys.Date())`'
Author: Kelly Cao and Rachel Robertson
output: html_document
editor: 
  markdown: 
    wrap: sentence
---
# Ridge Regression

### 2024-04-13 code/analysis-code/analysis2.5.qmd
Added to RF model
RF model added and tuned.

@@
 library(rsample) # for cross validation
+library(yardstick)
 ```
@@
 rf_fit <- rf_workflow%>%
-  fit(data = test_data)%>% # use the workflow to fit the rf model to the data
+  fit(data = train_data)%>% # use the workflow to fit the rf model to the data
 print(rf_fit)
 ```
+Now, I will examine the performance of the base random forest model.
+```{r}
+# Performance metrics of the first fit
+rf_aug <- augment(rf_fit, train_data) # augment to make predictions for rf_fit
 
+metrics <- metric_set(accuracy, f_meas) # create a set of metrics to test for classification
+first_fit_metrics <- metrics(truth = P_UTDHPV, estimate = .pred_class, data = rf_aug) # calculate metrics for first fit
+print(first_fit_metrics)
+```
+The accuracy of the model is fairly high, meaning that there it performs fairly when classifying the predictors in the model.
+However, the F1 score is relatively low which reflects low precision and recall of the model.
+To improve this we will tune the model.
+
+Now, I will continue by tuning the RF model using cross validation.
+I start with a smaller numer of trees and lower number of folds because my computer cannot computationally handle many trees or folds.
 ```{r}
-
 tune_spec <- 
@@
     mtry = tune(), # parameters of random forest model to tune are the mtry, trees, and min_n
-    trees = 50,
+    trees = 100,
     min_n = tune()) %>% 
@@
 ## I used ChatGPT to help me define the tree grid for the parameters mtry and min_n
-tree_grid <- grid_regular(mtry(range = c(1, 4)), 
-                          min_n(range = c(1, 5)),
+tree_grid <- grid_regular(mtry(range = c(1, 13)), 
+                          min_n(range = c(1, 200)),
                           levels = 6)
@@
 
-folds_data <- vfold_cv(test_data, v=5 )
+folds_data <- vfold_cv(train_data, v=5 )
 # Set the number of cores to use
@@
 ```
+```{r}
+# Find the best values for the tuned metrics, mtry and min_n
+# find the ideal ROC_AUC metric for the rf tree
+rf_res %>%
+  collect_metrics() %>% # get metrics from the rf model created
+  filter(.metric == "roc_auc") %>% # filter for the roc auc metric
+  select(mean, min_n, mtry) %>% # select the columsn for mean, min_n and mtry
+  pivot_longer(min_n:mtry,
+    values_to = "value",
+    names_to = "parameter"
+  ) %>%
+# Plot the ROC_AUC to find the ideal min_n and mtry
+  ggplot(aes(value, mean, color = parameter)) +
+  geom_point(show.legend = FALSE) +
+  facet_wrap(~parameter, scales = "free_x") +
+  labs(x = NULL, y = "AUC")
+```
+It seems that mtry is maximized between 100 and 150 and min_n is maximized between 1 and 5.
+We can use a grid search to tune within a range to determine which is the ideal parameter value.
+
+```{r}
+# grid search tuning
+rf_tune_grid <- grid_regular( #define a grid with a range for mtry and min_n
+  mtry(range = c(100, 150)),
+  min_n(range = c(1, 5)),
+  levels = 6
+)
+
+rf_tune_grid # display the grid
+
+# Tune using the defined tuning grid
+doParallel::registerDoParallel(cores = num_cores)
+
+set.seed(123) # set seed for reproducibility
+
+rf_res2 <- rf_tune_wf %>%
+  tune_grid(
+    resamples = folds_data, # use data from the CV folds
+    grid = rf_tune_grid) # use new tune grid
+rf_res2 %>%
+  autoplot()
+
+# Stop parallel processing
+stopImplicitCluster()
+```
+From the plot, it seems that the ideal minimum node size (min_n) is 5 when the number of randomly selected predictors (mtry) is  140, when accuracy and ROC_AUC are both accounted for.
+We can check this using the select_best function.
+```{r}
+# Specify the metric and optimization criteria
+best_auc <- select_best(rf_res2, metric = "roc_auc")
+
+final_rf <- finalize_model( # select the best model based on roc_auc
+  tune_spec,
+  best_auc
+)
+# fit the tuned model
+rf_rec <- recipe(P_UTDHPV ~ AGE + SEX + STATE + INS_STAT2_I + INCQ298A + INS_BREAK_I + INCPOV1 + RACEETHK + EDUC1 + LANGUAGE + MOBIL_1 + RENT_OWN + FACILITY, data = train_data) # use full model as recipe for the random forest model
+
+rf_model <- rand_forest()%>% # use rand_forest() to make a random forest model
+  set_engine("randomForest", seed = 123)%>%
+  set_mode("classification")
+
+rf_workflow <- workflow() %>% # create workflow for rf model
+  add_recipe(rf_rec)%>% # apply recipe
+  add_model(rf_model) # apply model
+
+rf_fit <- rf_workflow%>%
+  fit(data = train_data)%>% # use the workflow to fit the rf model to the data
+print(rf_fit)
+```
+The final random forest model was selected to optimize roc_auc, but it also has a fairly high accuracy. We can use
+```{r}
+final_rf_fit <- fit(final_rf, data = train_data)
+
+```
+
+```{r}
+# RF model compared to single predictor model
+## Create a single model
+rf_rec_null <- recipe(P_UTDHPV ~ SEX, data = train_data) # specify the model with a recipe; I could not get a null model to work so I used one predictor as a minimal model for comparison
+
+rf_spec_null <- rand_forest() %>%
+  set_engine("randomForest", importance = TRUE, seed = 123) %>% # set the seed and specify that importance is true
+  set_mode("classification")
+
+rf_workflow_mull <- workflow() %>% # create workflow with the recipe and spec
+  add_recipe(rf_rec_null) %>% 
+  add_model(rf_spec_null)
+
+rf_fit_null <- rf_workflow_null %>% # fit the null model
+  fit(data = train_data)
+
+# Get predictions for tuned model
+rf_tuned <- augment(rf_res, train_data)
+
+# Get predictions for first fit model
+
+# Create combined data frame
+
+# Plot tuned model, first fit, and single predictor model together
+
+```
+```{r}
+# RF model performance with cross validation
+
+```

### 2024-04-13 code/analysis-code/analysis4.qmd
Added to RF model
RF model added and tuned.

@@
 ---
-# Ridge Regression
+# Finding Best Model

### 2024-04-14 code/analysis-code/analysis4.qmd
Updated analysis

--- 2024-04-05 code/analysis-code/analysis4.qmd
@@
 ---
-# Ridge Regression
+# Finding Best Model
+Comparing models

### 2024-04-14 code/analysis-code/analysis2.5.R
Fixed RF
I fixed the issue in the RF code and made an R doc with the code in case we need to switch to pulling the chunks. This could speed up the rendering.

@@ -1,14 +1,4 @@
----
-title: "Data Analysis and Modeling of Teen Vaccination Surveys"
-date: '`r format(Sys.Date())`'
-Author: Kelly Cao and Rachel Robertson
-output: html_document
-editor: 
-  markdown: 
-    wrap: sentence
----
-# Random Forest Model
-```{r}
+## ---- packages --------
 library(tidymodels) # use tidymodels framework.
 library(ggplot2) # producing visual displays of data
 library(dplyr) # manipulating and cleaning data
@@ -17,9 +7,8 @@ library(randomForest) # making random forest model
 library(doParallel) # for parallel processing
 library(rsample) # for cross validation
 library(yardstick)
-```
 
-```{r}
+## ---- load-data --------
 # Load and preprocess data
 data_location <- here::here("data","processed-data","cleandata1.rds")
 mydata <- readRDS(data_location)
@@ -31,17 +20,15 @@ mydata$STATE <- droplevels(mydata$STATE, exclude = "Missing Data")
 mydata$MOBIL_1 <- droplevels(mydata$MOBIL_1, exclude = c("DON'T KNOW", "MISSING IN ERROR", "REFUSED"))
 mydata$FACILTY <- droplevels(mydata$FACILITY, exclude = "Missing Data")
 mydata$P_UTDHPV <- droplevels(mydata$P_UTDHPV, exclude = "Missing Data")
-```
 
-```{r}
+## ---- split-data --------
 # Split data into training and testing datasets
 set.seed(123) # seed for reproducibility
 split_data <- initial_split(mydata, prop = 0.8) # 80% split for training/testing data
 train_data <- training(split_data)
 test_data <- testing(split_data)
-```
 
-```{r}
+## ---- first-fit --------
 rf_rec <- recipe(P_UTDHPV ~ AGE + SEX + STATE + INS_STAT2_I + INCQ298A + INS_BREAK_I + INCPOV1 + RACEETHK + EDUC1 + LANGUAGE + MOBIL_1 + RENT_OWN + FACILITY, data = train_data) # use full model as recipe for the random forest model
 
 rf_model <- rand_forest()%>% # use rand_forest() to make a random forest model
@@ -54,21 +41,17 @@ rf_workflow <- workflow() %>% # create workflow for rf model
 
 rf_fit <- rf_workflow%>%
   fit(data = train_data)%>% # use the workflow to fit the rf model to the data
-print(rf_fit)
-```
[…]

### 2024-04-17 code/analysis-code/analysis2.5.qmd
updated analysis

@@
 )
+
 # fit the tuned model
@@
 
-rf_model <- rand_forest()%>% # use rand_forest() to make a random forest model
-  set_engine("randomForest", seed = 123)%>%
-  set_mode("classification")
+rf_tune_workflow <- workflow() %>% # create workflow for rf model
+  add_recipe(rf_rec)%>% # apply recipe
+  add_model(final_rf) # apply model
 
-rf_workflow <- workflow() %>% # create workflow for rf model
-  add_recipe(rf_rec)%>% # apply recipe
-  add_model(rf_model) # apply model
-
-rf_fit <- rf_workflow%>%
+rf_tune_fit <- rf_tune_workflow%>%
   fit(data = train_data)%>% # use the workflow to fit the rf model to the data
@@
 ```{r}
-final_rf_fit <- fit(final_rf, data = train_data)
+# make predictions
+rf_tuned_aug <- augment(rf_tune_fit, train_data)
 
+# find metrics for tuned fit
+tuned_fit_metrics <- metrics(truth = P_UTDHPV, estimate = .pred_class, data = rf_tuned_aug) # calculate metrics for first fit
+print(tuned_fit_metrics)
 ```
+The F1 measure has improved significantly, to 0.86 and the accuracy increased slightly, to 0.95.
 
+Now I will compare the model will the a single predictor model and the original fit to verify that there is improvement in the predicitons made.
 ```{r}
@@
 
-rf_workflow_mull <- workflow() %>% # create workflow with the recipe and spec
+rf_workflow_null <- workflow() %>% # create workflow with the recipe and spec
   add_recipe(rf_rec_null) %>% 
@@
 
-# Get predictions for tuned model
-rf_tuned <- augment(rf_res, train_data)
-
-# Get predictions for first fit model
-
-# Create combined data frame
-
-# Plot tuned model, first fit, and single predictor model together
-
+# Get predictions from single predictor model
+rf_null_aug <- augment(rf_fit_null, train_data)
 ```
-```{r}
-# RF model performance with cross validation
-
-```
+Note to future self: You cannot plot the models on coordinates because the predictions are classifications (not coninuous). Find another visual representation to compare the models.

### 2024-04-17 code/analysis-code/analysis3.qmd
updated analysis

@@
 
-(**Rachel redundantly added code that does the same thing. We can choose our favorite of the two next time we meet.**)
-I will now tune the penalty parameter using cross validation to find the penalty with the best roc_auc.
-```{r}
-lasso_penalty <- logistic_reg(penalty = tune(), mixture = 1) %>%
-  set_engine("glmnet")
-
-folds <- rsample::vfold_cv(train_data, v = 5)
-tune_grid <- tibble(penalty = 10^seq(-2, -1, length.out = 10))
-
-wf <- workflow() %>%
-  add_recipe(lasso_rec)%>%
-  add_model(lasso_penalty)
-
-tune_wf <- wf %>% 
-  tune_grid(resamples = folds,
-            grid = tune_grid,
-            control = control_grid(verbose = FALSE, save_pred = TRUE),
-            metrics = metric_set(roc_auc))
-
-# Select the best model based on ROC AUC
-best_mod <- tune_wf %>% 
-  select_best(metric = "roc_auc")
-
-final_wf <- finalize_workflow(wf, best_mod)
-print(final_wf)
-```
-
 # LASSO Re-fit

### 2024-04-17 code/analysis-code/analysis4.qmd
updated analysis

--- 2024-04-05 code/analysis-code/analysis4.qmd
@@
 ---
-# Ridge Regression
+# Finding Best Model
+Comparing the previous three models used in analysis1.qmd, analysis2.5.qmd, and analysis4.qmd.
+These models include : general logistic regression model, Random Forest regression model, and LASSO regression model.
+
+# Summary of the predictions and Metrics
+
+# Test Data

### 2024-04-17 code/analysis-code/analysis2.5.qmd
Analysis of rf and lasso completed
LASSO is complete, kinda, but we could also tune the penalty to F1 instead of roc_auc and see if this improves it. Also, we could try an elastic net model if we have time. Overall rf seems to have better performance. note: changes have not been rendered

@@
 ---
+```{r, include=FALSE, cache=FALSE}
+library(here)
+knitr::read_chunk(here("./code/analysis-code/rf-tree.R"))
+```
+
 # Random Forest Model
-```{r}
+```{r, message = FALSE, warning = FALSE}
 library(tidymodels) # use tidymodels framework.
@@
 library(rsample) # for cross validation
-library(yardstick)
+library(yardstick) # performance metrics
+library(vip) # visualize most important variables
 ```
 
-```{r}
+```{r, message = FALSE, warning = FALSE}
 # Load and preprocess data
@@
 mydata$P_UTDHPV <- droplevels(mydata$P_UTDHPV, exclude = "Missing Data")
+
 ```
 
-```{r}
+```{r, message = FALSE, warning = FALSE}
 # Split data into training and testing datasets
@@
 
-```{r}
+```{r, message = FALSE, warning = FALSE}
 rf_rec <- recipe(P_UTDHPV ~ AGE + SEX + STATE + INS_STAT2_I + INCQ298A + INS_BREAK_I + INCPOV1 + RACEETHK + EDUC1 + LANGUAGE + MOBIL_1 + RENT_OWN + FACILITY, data = train_data) # use full model as recipe for the random forest model
@@
 Now, I will examine the performance of the base random forest model.
-```{r}
+```{r, message = FALSE, warning = FALSE}
 # Performance metrics of the first fit
@@
 I start with a smaller numer of trees and lower number of folds because my computer cannot computationally handle many trees or folds.
-```{r}
+```{r, message = FALSE, warning = FALSE}
 tune_spec <- 
@@
 ```
-```{r}
-# Find the best values for the tuned metrics, mtry and min_n
-# find the ideal ROC_AUC metric for the rf tree
+
+```{r, message = FALSE, warning = FALSE}
 rf_res %>%
@@
 
-```{r}
+```{r, message = FALSE, warning = FALSE}
 # grid search tuning
@@
 We can check this using the select_best function.
-```{r}
+```{r, message = FALSE, warning = FALSE}
 # Specify the metric and optimization criteria
@@
 # fit the tuned model
+set.seed(123)
 rf_rec <- recipe(P_UTDHPV ~ AGE + SEX + STATE + INS_STAT2_I + INCQ298A + INS_BREAK_I + INCPOV1 + RACEETHK + EDUC1 + LANGUAGE + MOBIL_1 + RENT_OWN + FACILITY, data = train_data) # use full model as recipe for the random forest model
 
-rf_tune_workflow <- workflow() %>% # create workflow for rf model
+rf_workflow <- workflow() %>% # create workflow for rf model
   add_recipe(rf_rec)%>% # apply recipe
@@
 
-rf_tune_fit <- rf_tune_workflow%>%
+rf_tune_fit <- rf_workflow%>%
   fit(data = train_data)%>% # use the workflow to fit the rf model to the data
-print(rf_fit)
+  print(rf_tune_fit)
 ```
 The final random forest model was selected to optimize roc_auc, but it also has a fairly high accuracy. We can use
-```{r}
+```{r, message = FALSE, warning = FALSE}
 # make predictions
@@
 
-Now I will compare the model will the a single predictor model and the original fit to verify that there is improvement in the predicitons made.
-```{r}
+Now I will compare the model will the a single predictor model and the original fit to verify that there is improvement in the predictions made.
+```{r, message = FALSE, warning = FALSE}
 # RF model compared to single predictor model
@@
 ```
-Note to future self: You cannot plot the models on coordinates because the predictions are classifications (not coninuous). Find another visual representation to compare the models.
+Model metrics could not be computed because there were missing classifications in the single predictor model.
+Because of this, I will compare the single predictor and full model using confusion matrices.
+```{r}
+# confusion matrix for tuned model
+conf_mat(
+  data = rf_tuned_aug,
+  truth = P_UTDHPV,
+  estimate = .pred_class,
+  dnn = c("Prediction", "Truth"),
+  case_weights = NULL)
+
+#Confusion matrix for null model
+conf_mat(
+  data = rf_null_aug,
+  truth = P_UTDHPV,
+  estimate = .pred_class,
+  dnn = c("Prediction", "Truth"),
+  case_weights = NULL)
+```
+The confusion matrix for the tuned model (top) demonstrates that there are more false positives than false negatives.
+After calculating sensitivity (76.6%) and specificity(99.4%), a clear trade off between the two metrics is observed.
+The confusion matrix for the single predictor model (bottom) indicates that the model cannot predict an outcome of not up to date on HPV vaccination, where the NOT UTD prediction row is zero.
+Finally, I will test the tuned model using the test data from the split. To make pre
+```{r}
+# Making predictions using test data
+test_rf_aug <- augment(rf_tune_fit, test_data)
+
+# find metrics for tuned fit
+test_tuned_metrics <- metrics(truth = P_UTDHPV, estimate = .pred_class, data = test_rf_aug) # calculate metrics for first fit
+print(test_tuned_metrics)
+```
+The tibble reveals that the accuracy and F1 measure are greatly reduced when fitting the tuned model to the test data compared to the train data. This may indicate overfitting to the train data.
+Although the model is overfit, I will attempt to visualize the predictors chosen by the tuned model.
+```{r}
+# Finding importance of variables chosen
+rf_vars <- rf_tune_fit %>%
+  extract_fit_parsnip() %>% # use parsnip to extract to fit
+  vip::vip() #use vip for visualizing the most important variables
+
+print(rf_vars)
+```
+The most important variable, according to this model, is state, followed by income poverty ratio, then facility, then facility.
+Despite the importance above, it should be noted that the multicollinearity may affect the importance of the variables selected.
+This means that a model that regularizes the parameters could be ideal for finding the most important predictors.
+For this, we will try using the LASSO model

### 2024-04-17 code/analysis-code/analysis3.qmd
Analysis of rf and lasso completed
LASSO is complete, kinda, but we could also tune the penalty to F1 instead of roc_auc and see if this improves it. Also, we could try an elastic net model if we have time. Overall rf seems to have better performance. note: changes have not been rendered

@@
 library(mlbench)
-
+library(vip)
 ```
@@
 predictions <- lasso_fit %>%
-  augment(new_data = test_data)
+  augment(new_data = train_data)
 
@@
 predictions2 <- lasso_fit2 %>%
-  augment(new_data = test_data)
+  augment(new_data = train_data)
 
@@
 predictions3 <- lasso_fit3 %>%
-  augment(new_data = test_data)
+  augment(new_data = train_data)
 
@@
 
-# Cross Validation
+# Comparing Predictions
 Now we will compare the two lasso models with the penalty parameters of 0.005 and 0.01 to determine which model performs best when presented with "new data".
+The new data is that split from the original data (the test data).
+```{r}
+pen1_aug <- augment(lasso_fit2, test_data)
+pen2_aug <- augment(lasso_fit3, test_data)
+# Make find performance metrics with the tuned model
+metrics <- metric_set(accuracy, f_meas) # create a set of metrics to test for classification
+# Model tuned to parameter penalty = 0.005
+penaltytune1 <- metrics(truth = P_UTDHPV, estimate = .pred_class, data = pen1_aug)
+print(penaltytune1)
+# Model tuned to parameter penalty = 0.01
+penaltytune2 <- metrics(truth = P_UTDHPV, estimate = .pred_class, data = pen2_aug)
+print(penaltytune2)
+```
+The performance metrics reveal a similar accuracy between both models, but a higher, thus better, F1 measure, from the LASSO model with the penalty of 0.005.
+This might indicate that, although the model tuned to the penalty 0.01 maximizes roc_auc, it may be at the cost of other performance metrics such as F1 (or precision and/or recall).
+We will examine this further with confusion matrices.
+```{r}
+# Compare confusion matrices
+# penalty 0.005
+conf_mat(
+  data = pen1_aug,
+  truth = P_UTDHPV,
+  estimate = .pred_class,
+  dnn = c("Prediction", "Truth"),
+  case_weights = NULL)
+# penalty 0.01
+conf_mat(
+  data = pen2_aug,
+  truth = P_UTDHPV,
+  estimate = .pred_class,
+  dnn = c("Prediction", "Truth"),
+  case_weights = NULL)
+```
+The confusion matrix for neither of these models is ideal, as they both have a fairly high number of false positives, leading to a poor sensitivity.
+Because the model with a lower penalty parameter seems to perform better when presented with new data, I will look at the coefficients chosen by this model and determine which variables it deems most important to predicting HPV vaccine completion.
+```{r}
+# Finding importance of variables chosen
+lasso_vars <- lasso_fit2 %>%
+  extract_fit_parsnip() %>% # use parsnip to extract to fit
+  vip::vip() #use vip for visualizing the most important variables
+
+print(lasso_vars)
+```
+It seems that the most important factors for HPV vaccine completion are the state, facility and housing status.
+The states that have the highest predictive power of HPV vaccine completion are Rhode Island, Vermont, Mississippi, and North Dakota.
+
+Note to Kelly: Perhaps we should try tuning the penalty to a different metric. May we can try tuning to F1 and see if that produces a different penalty value?
+Note to self/ Discussion idea: Maybe vaccine policy or vaccine requirements to start school differs by state, so this might affect completion rates.

### 2024-04-17 products/manuscript/Manuscript.qmd
Analysis of rf and lasso completed
LASSO is complete, kinda, but we could also tune the penalty to F1 instead of roc_auc and see if this improves it. Also, we could try an elastic net model if we have time. Overall rf seems to have better performance. note: changes have not been rendered

@@
 # Methods
-
-## Data Cleaning and Processing
-
-### Loading the data
-
+## Loading Data
 The data was loaded from the original DAT file, entitled 'NISTEENPUF22.DAT' by specifying the column positions within the data file.
 The columns were then read and made into a data frame.
-Next, each of the factor variables were defined and the corresponding levels were assigned labels.
+Each of the factor variables was defined and the corresponding levels were assigned labels.
 The structure and summary statistics of all of the variables were examined.
-The loaded data is saved as a rds and a csv in the 'processed-data' folder.
-This does not alter the original raw data file.
+The loaded data is saved as a rds and a csv in the 'processed-data' folder, as to not alter the original raw data file.
 The definition of each factor variable in included in the README file within the 'raw-data' folder.
 
-Next, we explored the structure of the processed data and searched for any missing values.
-However, there were no NAs present.
-There are, however, missing values as labelled factor levels which must be accounted for.
-We identified the missing value labels by printing the factor levels.
-We then replaced the all of the values labelled missing with NA values.
-Next, we found a string of over 27000 NAs all of the variables for vaccination completion.
-These rows corresponded for all variables related to vaccine completion.
-Since this is the response variable of our scientific question, we eliminated these rows.
-We also eliminated rows with small numbers missing values, which were within the columns for facility, Wellchild exams, and insurance breaks.
-Lastly, we eliminatethe year column, as all of the data comes from the year 2022.
+## Cleaning Data
+The structure of the processed data was examined.
+The data was examined for missing values.
+No NAs were present, however missing values were present as labeled factor levels.
+The missing factor levels were printed, identified, and replaced with NA values.
+A string of 27,000 NAs were present in the outcome variable, indicating vaccination completion.
+All rows corresponding to the NAs for the missing outcome value were removed.
+Approximately 160 additional rows with NA values for facility or insurance status variables were removed.
+Over 16,000 rows of data remained.
+The year column was eliminated, as all of the data is from 2022.
 We saved the cleaned data as an rds and csv in the 'processed-data' folder.
+## Exploratory Analysis
+Exploratory analysis was used to determine the distribution of the predictors, including the socioeconomic variables that were chosen for analysis.
+The distribution of single predictors including income-to-poverty ratio, income, state, race, ethnicity, vaccination status, and facility were examined using histograms and scatter plots.
+The following predictors were compared visually and fitted to a linear regression to examine preexisting associations: vaccination status and ethnicity, vaccination status and poverty status, income poverty ratio and insurance status, income poverty ratio and ethnicity.
+## Analysis
+### Generalized Logistic Regression
+A generalized logistic regression was performed with bi-directional step-wise selection for predictor reduction.
+Variables that were redundant of a major variable were excluded from analysis.
+The variables initially included were age, sex, state, race/ethnicity, insurance status, income, insurance breaks, poverty status, education, language, geographic mobility, facility, and housing status.
+The model was evaluated with a residual plot and correlation matrix to reveal strong multicollinearity.
+This is the basis for the subsequent analysis methods, as they are more resistant to collinearity.
+### Principal Component Analysis
+Principal component analysis (PCA) was performed to do dimensionality reduction.
+PCA would ideally transform the highly correlated predictors into uncorrelated components.
+The h2o package was used to provide a machine-learning platform in RStudio, to allow machine-learning methods ot be implemented.
+Due to the data being mainly categorical variables, the PCA proved to be unsuccessful.
+A LASSO Ridge Regression was performed with the h2o package, to mitigate the multicollinearity issue.
+### Random Forest Model
+A random forest model was used, as it is resistant to multicollinearity.
+The tidymodels package was used to produce a random forest model with the classification method.
+The model was trained and fit from training data, which was an 80% split of the original data.
+After evaluation, the model parameters for minimal node size and number of selected predictors were tuned for using a grid search with cross-validation resamples.
+After tuning, the best model was extracted based on the ROC AUC metric and was fit to the training data.
+The tuned model was used to make predictions and evaluated with performance metrics, a confusion matrix, and a comparison to a single-predictor random forest model.
+Finally, the tuned random forest model was re-evaluated by being fit to the testing data.
+### LASSO Regression
 
-## Exploratory Analysis
+{{< pagebreak >}}
 
-We use our exploratory analysis to determine the distribution of our predictors, including the socioeconomic variables that we chose for analysis.
-The overall distribution of each factor variable is shown below.
+# Results
+## Exploratory Data Analysis
+Exploratory analysis is used to determine the distribution of our predictors, including the socioeconomic variables that we chose for analysis.
+Nearly all of the demographic factors have a similar distribution between those who have completed their HPV vaccination regime versus those who have not completed it.
 
@@
 ```
-```
-
 The distribution of the numerical income to poverty ratio can be shown below.
@@
 
-{{< pagebreak >}}
-
-# Analysis
+## Analysis
 
@@
 
-# References
+# Discussion
+
+# Conclusion
 

### 2024-04-17 code/analysis-code/rf-tree.R

@@ -1,4 +1,6 @@
+
 ## ---- packages --------
+
 library(tidymodels) # use tidymodels framework.
 library(ggplot2) # producing visual displays of data
 library(dplyr) # manipulating and cleaning data
@@ -9,6 +11,7 @@ library(rsample) # for cross validation
 library(yardstick)
 
 ## ---- load-data --------
+
 # Load and preprocess data
 data_location <- here::here("data","processed-data","cleandata1.rds")
 mydata <- readRDS(data_location)

### 2024-04-17 code/analysis-code/analysis2.qmd
Updated manuscript analysis

--- 2024-03-16 code/analysis-code/analysis2.qmd
@@
 library(dplyr) 
+library(rsample)
 
@@
 mydata <- mydata[complete.cases(mydata$P_UTDHPV), ]
+
+# Set Seed for reproducibility
+rngseed = 1234
+set.seed(rngseed)
+
+# Create training data by alloting 3/4 of the data for training
+data_split <- initial_split(mydata, prop = 3/4)
+
+# Create data frames for the two sets:
+train_data <- training(data_split)
+test_data  <- testing(data_split)
 
@@
 
-# Convert mydata to H2OFrame
-mydata_h2o <- as.h2o(mydata)
+# Convert train_data to H2OFrame
+mydata_h2o <- as.h2o(train_data)
 
@@
 
+```{r}
+# Replicating the previous analysis with test data
+
+# Convert train_data to H2OFrame
+mydata_h2o <- as.h2o(test_data)
+
+# Create a fold column for cross-validation
+fold_column <- h2o.kfold_column(data = mydata_h2o, nfolds = 10, seed = 1)  # Specify the number of folds and a seed for reproducibility
+
+# Add the fold column to the H2OFrame
+mydata_h2o <- h2o.cbind(mydata_h2o, fold_column)
+
+# Specify predictor variables
+predictor_columns <- c("AGE", "SEX", "STATE", "INS_STAT2_I", "INCQ298A", "INS_BREAK_I", 
+                       "INCPOV1", "RACEETHK", "EDUC1", "LANGUAGE", "MOBIL_1", "RENT_OWN", "FACILITY")
+
+# Evaluate model performance using cross-validation
+cv_metrics <- h2o.glm(x = predictor_columns, 
+                      y = "P_UTDHPV", 
+                      training_frame = mydata_h2o, 
+                      fold_column = "C1",  # Use the name of the fold column (default name is "C1")
+                      family = "binomial")
+
+# Display results
+print(cv_metrics)
+
+```
+
 ### Conclusion

### 2024-04-17 products/manuscript/Manuscript.qmd
Updated manuscript analysis

@@
 # Methods
+
 ## Loading Data
+
 The data was loaded from the original DAT file, entitled 'NISTEENPUF22.DAT' by specifying the column positions within the data file.
@@
 ## Cleaning Data
+
 The structure of the processed data was examined.
@@
 We saved the cleaned data as an rds and csv in the 'processed-data' folder.
-## Exploratory Analysis
-Exploratory analysis was used to determine the distribution of the predictors, including the socioeconomic variables that were chosen for analysis.
+## Exploratory Analysis Exploratory analysis was used to determine the distribution of the predictors, including the socioeconomic variables that were chosen for analysis.
 The distribution of single predictors including income-to-poverty ratio, income, state, race, ethnicity, vaccination status, and facility were examined using histograms and scatter plots.
 The following predictors were compared visually and fitted to a linear regression to examine preexisting associations: vaccination status and ethnicity, vaccination status and poverty status, income poverty ratio and insurance status, income poverty ratio and ethnicity.
-## Analysis
-### Generalized Logistic Regression
-A generalized logistic regression was performed with bi-directional step-wise selection for predictor reduction.
+## Analysis \### Generalized Logistic Regression A generalized logistic regression was performed with bi-directional step-wise selection for predictor reduction.
 Variables that were redundant of a major variable were excluded from analysis.
@@
 This is the basis for the subsequent analysis methods, as they are more resistant to collinearity.
-### Principal Component Analysis
-Principal component analysis (PCA) was performed to do dimensionality reduction.
+### Principal Component Analysis Principal component analysis (PCA) was performed to do dimensionality reduction.
 PCA would ideally transform the highly correlated predictors into uncorrelated components.
@@
 A LASSO Ridge Regression was performed with the h2o package, to mitigate the multicollinearity issue.
-### Random Forest Model
-A random forest model was used, as it is resistant to multicollinearity.
+### Random Forest Model A random forest model was used, as it is resistant to multicollinearity.
 The tidymodels package was used to produce a random forest model with the classification method.
@@
 # Results
+
 ## Exploratory Data Analysis
+
 Exploratory analysis is used to determine the distribution of our predictors, including the socioeconomic variables that we chose for analysis.
@@
 ```
+
 The distribution of the numerical income to poverty ratio can be shown below.
@@
 Some of these predictors are a subset of the other, or may be used to describe an overall category.
-For this reason
 
@@
 
-![Table 3. Human Papillomavirus vaccine status variables and meanings.](assets/UTDHPV-variable-table.png)
+![Table 3. Human Papillomavirus vaccine status variables and meanings.](./images/UTDHPV-variable-table.png)
 
@@
 
-![Table 4. Potential predictor variable names and meanings.](assets/Predictor-variable-table.png)
+![Table 4. Potential predictor variable names and meanings.](./images/Predictor-variable-table.png)
 
@@
 
-To determine which predictor variables will be best suited for a general linear model, I will do a step-wise comparison and then measure the resulting models with AIC and WAIC.
+To determine which predictor variables will be best suited for a general linear model, we employed a stepwise comparison is performed and then assessed with AIC and WAIC.
+The stepwise model is set up with `P_UTDPHPV` as the outcome and the remaining chosen variables as the predictors.
+Each predictors were added and dropped in different variations, and the resulting AIC were recorded for each of the different models.
+From this, we were able to determine that the model with the lowest AIC (and thus the greatest fit) was the model with all of the predictors that we previously picked out.
 
+Moving from that, we plotted the residuals from the logistic regression model to check for non-linearity and discovered that the predictors potentially had a high indication of collinearity.
+to further confirm this, a VIF analysis was attempted but was unable to proceed as the multicollinearity was too great for too many variables.
+As a result, we decided to perform a correlation matrix, which performed as expected and verified our concerns.
+
+INCLUDE EITHER REGRESSION PLOT OR STEPWISE FINAL RESULT IMAGE
+
+In order to reduce the multicollinearity issue, we initially attempted to use PCA.
+To employ the unsupervised learning model, we use the `h2o` package.
+The PCA resulted in one principle component with a high standard deviation and high proportion of variance, showing that the Pc1 captures a large portion (97.9%) of variability in the data.
+
+INCLUDE IMAGE OF PCA RESULTS
+
+The result of having only one principle component further supports the fact that many of the predictor variables are highly collinear.
+One PC shows that a PCA must reduce the data object to a single dimension.
+Having a single principal component could be problematic for my analysis, particularly if it leads to the loss of crucial information or structure in my data due to collinearity among the original features.
+
+Using the same package, I used the LASSO logistical regression model and performed a co-validation test on it.
+
+```
+MSE:  0.1558815
+RMSE:  0.3948183
+LogLoss:  0.4850906
+Mean Per-Class Error:  0.4906615
+AUC:  0.6596287
+AUCPR:  0.8704217
+Gini:  0.3192575
+R^2:  0.05771195
+Residual Deviance:  16070.08
+AIC:  16266.08
+```
+
+Random Forest
+
+LASSO Regression
 # Discussion

### 2024-04-18 code/analysis-code/analysis1.qmd
Updated manuscript analysis
Examined problems with analysis in Rf and with LASSO. Added to manuscript.
I think that neither might be more suitable and we should try elastic net.

@@
 The residual plot shows clear collinearity that is not represented by the current model.
-To find the predictors with collinnearity we will use a correlation matrix.
+To find the predictors with multicollinearity, we will use a correlation matrix.
 This is because the VIF analysis will not run when multicollinearity is too high.
@@
 print(variable_pairs)
-
 ```

### 2024-04-18 code/analysis-code/analysis2.5.qmd
Examined problems with analysis in Rf and with LASSO. Added to manuscript.
I think that neither might be more suitable and we should try elastic net.

@@
 tree_grid <- grid_regular(mtry(range = c(1, 13)), 
-                          min_n(range = c(1, 200)),
+                          min_n(range = c(1, 500)),
                           levels = 6)
@@
 This means that a model that regularizes the parameters could be ideal for finding the most important predictors.
-For this, we will try using the LASSO model
+For this, we will try using the LASSO model.

### 2024-04-18 code/analysis-code/analysis3.qmd
Examined problems with analysis in Rf and with LASSO. Added to manuscript.
I think that neither might be more suitable and we should try elastic net.

@@
   prep()
+
 ```
@@
 
+set.seed(123) # ste seed for reproducability
+
 # LASSO model specification
@@
 
+# find the best penalty for ROC AUC
+best_roc <- lasso_tuned %>%
+  select_best(metric = "roc_auc")
+best_roc
+
 ```
@@
 # LASSO Re-fit
-Now, we will run the LASSO regression again, but with the tuned penalty parameter of 0.01. 
+Now, we will run the LASSO regression again, but with the tuned penalty parameter of 0.001. 
 ```{r}
-# Specify logistic regression model
-lasso_spec3 <- logistic_reg(penalty = 0.01, mixture = 1) %>%
+lasso_tune_spec <- logistic_reg(penalty = 0.001, mixture = 1) %>%
   set_engine("glmnet")
 
+# Create workflow
+lasso_tune_wf <- workflow() %>%
+  add_recipe(lasso_rec)%>%
+  add_model(lasso_tune_spec)
+
 # Fit the model using training data
-lasso_fit3 <- lasso_wf %>%
-  add_model(lasso_spec3) %>%
+lasso_tune_fit <- lasso_tune_wf %>%
   fit(data = train_data)
@@
 # Extract and tidy model parameters
-lasso_fit3 %>%
-  extract_fit_parsnip() %>%
-  tidy()
-
-# Extract and tidy model parameters
-lasso_coef <- lasso_fit3 %>%
+lasso_tune_coef <- lasso_tune_fit %>%
   extract_fit_parsnip() %>%
@@
 # Filter out predictors with non-zero coefficients
-lasso_nonzero_coef <- lasso_coef %>%
+lasso_nonzero_coef <- lasso_tune_coef %>%
   filter(estimate != 0)
@@
 # Make predictions for ROC_AUC with the tuned model
-predictions3 <- lasso_fit3 %>%
+predictions3 <- lasso_tune_fit %>%
   augment(new_data = train_data)
@@
 # Comparing Predictions
-Now we will compare the two lasso models with the penalty parameters of 0.005 and 0.01 to determine which model performs best when presented with "new data".
+Now we will compare the two lasso models with the penalty parameters of 0.005 and 0.001 to determine which model performs best when presented with "new data".
 The new data is that split from the original data (the test data).
@@
 pen1_aug <- augment(lasso_fit2, test_data)
-pen2_aug <- augment(lasso_fit3, test_data)
+pen2_aug <- augment(lasso_tune_fit, test_data)
 # Make find performance metrics with the tuned model
@@
 ```
-The performance metrics reveal a similar accuracy between both models, but a higher, thus better, F1 measure, from the LASSO model with the penalty of 0.005.
-This might indicate that, although the model tuned to the penalty 0.01 maximizes roc_auc, it may be at the cost of other performance metrics such as F1 (or precision and/or recall).
-We will examine this further with confusion matrices.
+The performance metrics reveal a similar accuracy between both models, but a higher, thus better, F1 measure, from the LASSO model with the penalty of 0.001. 
 ```{r}
@@
   case_weights = NULL)
-# penalty 0.01
+# penalty 0.001
 conf_mat(
@@
 # Finding importance of variables chosen
-lasso_vars <- lasso_fit2 %>%
+lasso_vars <- lasso_tune_fit %>%
   extract_fit_parsnip() %>% # use parsnip to extract to fit
@@
 The states that have the highest predictive power of HPV vaccine completion are Rhode Island, Vermont, Mississippi, and North Dakota.
+```{r}
+# Extract and tidy model parameters
+lasso_tune_coef <- lasso_tune_fit %>%
+  extract_fit_parsnip() %>%
+  tidy()
+
+# Filter out predictors with non-zero coefficients
+lasso_tune_coef <- lasso_tune_coef %>%
+  filter(estimate != 0)
+lasso_tune_coef
+```
 

### 2024-04-18 products/manuscript/Manuscript.qmd
Examined problems with analysis in Rf and with LASSO. Added to manuscript.
I think that neither might be more suitable and we should try elastic net.

@@
 # Results
-
 ## Exploratory Data Analysis
-
 Exploratory analysis is used to determine the distribution of our predictors, including the socioeconomic variables that we chose for analysis.
@@
 ```
-
-The distribution of the numerical income to poverty ratio can be shown below.
-It displays a value from 1-3.
-A ratio displaying less than 1 depicts an income less than the poverty level.Any income ratio \>1 indicates an income greater than the poverty level.
-Most of the observations were gathered from households with income ratios of 3, as shown by the skewed plot below.
-
+Income-poverty-ratio is a numeric value that indicates how close to poverty a family is living.
+It is calculated by income over the poverty income threshold and ranges from values 1 (closest to poverty) and 3 (furthest from poverty).
+The distribution of the income poverty ratio indicates a strong left skew.
+The highest frequency of observations was gathered from households with an income poverty ratio, meaning most survey participants live well above the poverty threshold.
+The family income is right skewed with the majority of the observations coming from \$75000+ income level.
+This indicates that a high frequency of participants are above the poverty income threshold, but also in a modest income class.
 ![Figure 1. Income-poverty ratio distribution](./images/income.png)
-
-Next, the distribution of surveys given per state was analyzed.
-The state variable shows a more uniformed distribution, suggesting that the observations were gathered fairly across the US states and territory.
-Although it is noted a few territories provides 0 observations.
-![Figure 2. State distribution](./images/state.distribution.png)
-
-The family income variable provides a more normal distribution with the majority of the observation coming from \$75000+ income level.
-![Figure 3. The family income distribution](./images/family.income.distribution.png)
-
+![Figure 2. The family income distribution](./images/family.income.distribution.png)
+The state variable shows a more uniform distribution, suggesting that the observations were gathered fairly across the US states and territories.
+States and territories with a significantly larger number of observations include Texas, West Virginia, Pennsyvania, Puerto Rico, New York, and Illinois.
+Although it is noted a few territories provide 0 observations.
 The race and ethnicity variables show a distribution that is similar to the U.S.
-census data, which concludes that the racial composition of the U.S.
-is 58.9% white, 13.6% African American, and 19.1% Hispanic or Latino.
-
-![Figure 4.Racial distribution](./images/race.distribution.png)
-
-![Figure 5.Ethnicity distribution](./images/ethnicity.distribution.png)
-
-The Insurance status variable shows that the majority of the observations came from households with private insurance only.
-This is followed by any medicaid and then other insurance.
-This is mirrored in the facility variable, where a majority of its observations show private facilities being the location.
-![Figure 6.
-Insurance status distribution](./images/insurance.status.distribution.png) ![Figure 7.
-Distribution of the medical facility where the survey was administerred](./images/facility.distribution.png)
-
-The vaccination variable indicates nearly 80% of the observations had an Up-to-date status for the 1+ shot HPV vaccination.
-We will follow by examining the percentage of up-to-date vaccination status by state.
-
-![Table 2. Vaccine status distribution](./images/state-vaccine-completion.png)
-
-We examine the facility that submitted the survey, which is stratified by the vaccine completion status of the survey subject.
-
-![Figure 8.Facility and vaccine completion distribution](./images/facility-vaccination.png)
-
-We examine the percentage of vaccination completion by race and ethnicity.
-Because there are different counts of each race/ethnicity for each factor level, the percentage of those with up-to-date vaccine status out of each race/ethnicity level was determined.
-This adjusts for the different counts in each racial category.
-The Hispanic ethnicity was found to have the highest vaccine completion while non-hispanic white has the lowest.
-
-![Figure 9. Distribution of vaccine completion stratified by race and ethnicity](./images/raceeth-income-stratified.png)
-
-## Model and Plot Fitting
-
-A boxplot is created with income-poverty ratio to vaccination status.
-The box plot indicates little difference between the two status.
-A statistical model is fitted to the same variables.
-A p-value of 0.1358 suggests there is little significance between the income-poverty variable and the vaccination status.
-
-![Figure 10.](./images/income-vaccination.png)
-
-A boxplot is created with income-poverty ratio to insurance status.
-The box plot indicates there are distinct difference between the two status.
-A statistical model is fitted to the same variables.
-The p-value of \< 2.26e-16 suggests there is statistical significance between the income-poverty variable and the insurance status.
-
+census data.
+The census data states that the racial composition of the U.S.
+is 58.9% white, 13.6% African American, and 19.1% Hispanic or Latino, while our survey population is composed of 58.% white, 9.15% African American, and 19.9% Hispanic or Latino participants.
+The majority of observations came from households with private insurance only The Insurance status variable shows that the majority of the observations came from households with private insurance only (58.6%), followed by Medicaid (31.6%), other insurance (7.7%), and uninsured (2.2%).
+This is mirrored in the facility variable, as survey location, where a majority of its observations show private facilities (34.4%), followed by mixed facilities (25.8%), hospitals (11.3%), and all public facilities (9.9%).
+15.2% of the facility types were unknown.
+Upon stratification of facility type by HPV vaccination completion status,  it was found that the majority of participants who were up-to-date were surveyed at private or mixed facilities.
+They likely received their HPV vaccination there as well.
+![Figure 3.Facility and vaccine completion distribution](./images/facility-vaccination.png)
+A total of 79% of participants had up-to-date status for receiving 1 or more shots of HPV vaccination.
+U.S.
+states and territories had an HPV vaccination completion rate ranging from 69% to 95%, fluctuating around ~80%.
+There are different counts of each race/ethnicity for each factor level, so the percentage of those with up-to-date vaccine status out of each race/ethnicity level was adjusted.
+The Hispanic ethnicity was found to have the highest vaccine completion while the non-Hispanic white ethnicity had the lowest.
+It was also found that for Hispanic and Non-Hispanic black ethnicities, the income poverty ratio was higher for those who were not up-to-date in HPV vaccination status, compared to those who were.
+The opposite was true for mixed ethnic groups, where vaccine completion corresponded with a higher mean income-poverty ratio.
+When examining the overall HPV vaccination status stratified by income poverty ratio, the income poverty ratio did not differ between those who are up to date and those who are not.
+This may be accounted for by other trends such as that shown with ethnicity, above.
+![Figure 9. Distribution of income poverty ratio stratified by race and HPV vaccination completion status](./images/raceeth-income-stratified.png)
+	Subsequently, income poverty ratio and insurance status were examined, and stratified by HPV vaccination status.
+It is apparent that those with private and other insurance status had the highest income-poverty ratio, while those with Medicaid or uninsured were closer to the poverty income threshold.
+For those with Medicaid, those who were up-to-date in HPV vaccinations had a slightly lower income-poverty ratio than those who were not up to date.
 ![Figure 11. Income poverty ratio and insurance status](./images/insurance-income-stratified.png)
-
-Income-poverty ratio was fit to vaccine status and insurance status in a multi-linear model.
-
-A box-plot was created for income-poverty level versus race/ethnicity and this was stratified by vaccine completion (UTD or NOT UTD).
-A lower income-poverty ratio means the group is closer to the poverty line.
-For the non-Hispanic black and Hispanic racial and ethnic groups, vaccine completion corresponded with a lower mean income-poverty ratio.
-For mixed ethnic groups, vaccine completion corresponded with a higher mean income-poverty ratio.
-
-![Figure 12. Income-poverty level versus ethnic group](./images/raceeth-income-stratified.png)
-
 ## Analysis
-
 We begin with various socioeconomic predictors that interest us.

### 2024-04-19 code/analysis-code/analysis2.5.qmd
Added results to manuscript
Tried to save the VIP plot from the rf model, but it is an environment and I'm not sure how to extract the plot.

@@
   vip::vip() #use vip for visualizing the most important variables
-
-print(rf_vars)
+rf_vars
 ```

### 2024-04-19 products/manuscript/Manuscript.qmd
Added results to manuscript
Tried to save the VIP plot from the rf model, but it is an environment and I'm not sure how to extract the plot.

@@
 ```
+
 Income-poverty-ratio is a numeric value that indicates how close to poverty a family is living.
@@
 ![Figure 1. Income-poverty ratio distribution](./images/income.png)
+
 ![Figure 2. The family income distribution](./images/family.income.distribution.png)
+
 The state variable shows a more uniform distribution, suggesting that the observations were gathered fairly across the US states and territories.
@@
 ![Figure 3.Facility and vaccine completion distribution](./images/facility-vaccination.png)
+
 A total of 79% of participants had up-to-date status for receiving 1 or more shots of HPV vaccination.
@@
 ![Figure 9. Distribution of income poverty ratio stratified by race and HPV vaccination completion status](./images/raceeth-income-stratified.png)
+
 	Subsequently, income poverty ratio and insurance status were examined, and stratified by HPV vaccination status.
@@
 ![Figure 11. Income poverty ratio and insurance status](./images/insurance-income-stratified.png)
+
 ## Analysis
-We begin with various socioeconomic predictors that interest us.
-Some of these predictors are a subset of the other, or may be used to describe an overall category.
-
-The variables used were chosen based on their interpretability and applicability to our research question “Which socioeconomic and geographic factors impact HPV vaccine completion rates among teenagers?”
-.
-There are seven potential outcome variables for HPV vaccine completion:
-
+The variables selected for analysis were chosen based on their interpretability and applicability to our research question “Which socioeconomic and geographic factors impact HPV vaccine completion rates among teenagers?” Any variables that capture repetitive data were removed and the most important or straightforward version of the variable was kept in the model.
+Seven variables represented data for the outcome of HPV vaccination completion:
 ![Table 3. Human Papillomavirus vaccine status variables and meanings.](./images/UTDHPV-variable-table.png)
@@
 It has the benefit of being able to use Age as a predictor variable and also not biasing the model towards older ages who are required to receive more than one shot to be considered up-to-date.
-Additionally, in April of 2022, the SAGE work group of the WHO officially recommended a single-dose HPV vaccine regime, because vaccine efficacy was found to be significantly high with a single dose.
+Additionally, in April of 2022, the SAGE work group of the WHO officially recommended a single-dose HPV vaccine regime, because vaccine efficacy was found to be nearly as high with a single dose as it is with the multi-dose regime.
 This recommendation was given with the goal of increasing vaccine coverage in areas with lower vaccine accessibility and availability (13).
 In September of 2023, PAHO followed suit in recommending a single-dose HPV vaccine regime for the Americas (14).
-
-The predictor variables we chose to examine are listed below:
-
+This supports our rationale for choosing the single or more dose, up-to-date, outcome variable.
+The variables that remained in the initial logistic regression are listed below.
 ![Table 4. Potential predictor variable names and meanings.](./images/Predictor-variable-table.png)
@@
 RACEETHK is ideal because it includes both race and ethnicity.
-
-To determine which predictor variables will be best suited for a general linear model, we employed a stepwise comparison is performed and then assessed with AIC and WAIC.
+### Stepwise Selection with Logistic Regression
+To determine which predictor variables will be best suited for a general logistic model, we employed a bi-directional stepwise selection method.
+The results were assessed with AIC and WAIC.
 The stepwise model is set up with `P_UTDPHPV` as the outcome and the remaining chosen variables as the predictors.
 Each predictors were added and dropped in different variations, and the resulting AIC were recorded for each of the different models.
-From this, we were able to determine that the model with the lowest AIC (and thus the greatest fit) was the model with all of the predictors that we previously picked out.
-
-Moving from that, we plotted the residuals from the logistic regression model to check for non-linearity and discovered that the predictors potentially had a high indication of collinearity.
-to further confirm this, a VIF analysis was attempted but was unable to proceed as the multicollinearity was too great for too many variables.
+From this, we were able to determine that the model with the lowest AIC (and thus the greatest fit) was the model that included all of the original factors.
+The logistic regression with the selected variables was assessed by plotting the residuals.
+This revealed that the predictors had a high indication of collinearity.
+To confirm this, a VIF analysis was attempted but was unable to proceed as the multicollinearity was too great for too many variables.
 As a result, we decided to perform a correlation matrix, which performed as expected and verified our concerns.
-
+Multi-collinearity was present between the majority of the predictor variables, thus another model type that could handle collinearity would be ideal.
 INCLUDE EITHER REGRESSION PLOT OR STEPWISE FINAL RESULT IMAGE
-
+### Principal Component Analysis and LASSO Regression
 In order to reduce the multicollinearity issue, we initially attempted to use PCA.
@@
 The PCA resulted in one principle component with a high standard deviation and high proportion of variance, showing that the Pc1 captures a large portion (97.9%) of variability in the data.
-
 INCLUDE IMAGE OF PCA RESULTS
-
 The result of having only one principle component further supports the fact that many of the predictor variables are highly collinear.
@@
 Using the same package, I used the LASSO logistical regression model and performed a co-validation test on it. 
-
 ```
@@
 ```
+### Random Forest Model
+	We used a Random Forest Model to assess which predictors are associated with HPV vaccination completion rates in U.S.
+teens.
+Decision tree models are known to be able to handle collinearity and predictors with different trends.
+A random forest model is ideal because it employs bootstrapping to produce randomly chosen iterations of decision trees to introduce variability.
+The best random forest model can then be tuned for and selected based on which tree performs best in a certain metric.
+The random forest model was produced using the tidy model framework and tuned using a cross-validated grid search for the minimum node size (min_n) and the number of variables to randomly sample at each split (mtry).
+The model with the best ROC AUC metric was selected, and predictions were produced.
+The model indicated that the most important predictors of HPV vaccination completion, in order of most to least important, are state, income level, facility, age, race/ ethnicity, insurance status, education, housing status, sex, and geographic mobility.
+[insert plot  of most important predictors from vip package; not sure how, because it is an environment and I don’t know how to extract one plot]
+However, after testing the same metrics with new data (testing data), the F1 metric was reduced from 0.86 to 0.16, which could indicate overfitting or that the model is “too big” for the data.
 
-Random Forest
-
-LASSO Regression
+#### LASSO Regression with Tidymodels
 # Discussion

### 2024-04-29 code/analysis-code/analysis4.qmd
Made elastic net model and tested
tuned elastic net model performs the same as the previous lasso model. Because of this, I am thinking that we should choose the rf model with the best performance metrics (despite being overfit).

--- 2024-04-13 code/analysis-code/analysis4.qmd
@@
 ---
-# Finding Best Model
+# Elastic Net Regression
+Because the random forest model was overfit and the lasso model did not capture all of the trends, we will create an elastic net model of our categorical variables.
+
+## Loading data
+We will begin by loading the data and opening the libraries needed for the analysis
+```{r}
+# Open the libraries needed
+library(tidymodels) # use tidymodels framework
+library(ggplot2) # producing visual displays of data
+library(dplyr) # manipulating and cleaning data
+library(here) # making relative pathways
+library(glmnet) # for LASSO regression engine
+library(doParallel) # for parallel processing
+library(rsample) # for cross validation
+library(yardstick) # for metrics
+library(mlbench)
+library(vip)
+
+# Load and preprocess data
+data_location <- here::here("data","processed-data","cleandata1.rds")
+mydata <- readRDS(data_location)
+
+# Remove rows with missing values
+mydata$SEX <- droplevels(mydata$SEX, exclude = c("DON'T KNOW", "MISSING IN ERROR", "REFUSED"))
+mydata$INS_STAT2_I <- droplevels(mydata$INS_STAT2_I, exclude = "MISSING Data")
+mydata$STATE <- droplevels(mydata$STATE, exclude = "Missing Data")
+mydata$MOBIL_1 <- droplevels(mydata$MOBIL_1, exclude = c("DON'T KNOW", "MISSING IN ERROR", "REFUSED"))
+mydata$FACILTY <- droplevels(mydata$FACILITY, exclude = "Missing Data")
+mydata$P_UTDHPV <- droplevels(mydata$P_UTDHPV, exclude = "Missing Data")
+```
+Next, we will split the data into training and testing groups so that the LASSO regression has a comparison group.
+```{r}
+# Split data into training and testing datasets
+set.seed(123) # seed for reproducibility
+split_data <- initial_split(mydata, prop = 0.8) # 80% split for training/testing data
+train_data <- training(split_data)
+test_data <- testing(split_data)
+```
+## Elastic net first fit
+Now we will proceed by fitting the elastic ent model for the first time
+```{r}
+# Split the data into training and testing sets
+set.seed(123)  # for reproducibility
+
+# Create amodel with the specifications for elatic net
+elastic_net <- logistic_reg(mode = "classification") %>%
+  set_engine("glmnet") %>%
+  set_mode("classification") %>%
+  set_args(
+    penalty = 0.005,  # use the penalty for the previous lasso regression
+    mixture = 0.5   
+  )
+
+# create a recipe, using your formula, for elastic net
+elastic_rec <- recipe(P_UTDHPV ~ AGE + SEX + STATE + INS_STAT2_I + INCQ298A + INS_BREAK_I + INCPOV1 + RACEETHK + EDUC1 + LANGUAGE + MOBIL_1 + RENT_OWN + FACILITY, data = train_data) %>%
+  step_dummy(all_nominal(), -all_outcomes())  %>% # convert categorical variables (nominal) into dummy variables
+  step_rm(all_outcomes(), -all_outcomes()) #remove the outcome variables from the dataset
+
+# Prepare the elastic net recipe
+elastic_prep <- elastic_rec %>%
+  prep()
+
+# produce elastic net workflow
+workflow <- workflow() %>%
+  add_recipe(elastic_rec) %>%
+  add_model(elastic_net)
+
+# implement elastic net workflow
+en_fit <- workflow %>% 
+  fit(data = train_data) # fit to training data
+
+en_predictions <- augment(en_fit, train_data) #make predictions on training data
+print(en_predictions)
+```
+Now that we have produced predictions with the elastic net model, we will look at the performance metrics of this model and then tune the model parameters.
+```{r}
+# Calculate ROC AUC
+roc_auc <- en_predictions %>%
+  roc_auc(truth = P_UTDHPV, .pred_UTD)
+print(roc_auc)
+```
+The ROC AUC reveals that there is no predictive power with the elastic net model, so I will tune this model to reveal if it produces better predictions.
+```{r}
+# Tune for penalty and mixture parameters
+
+# produce the model spec using the same recipe, except with the aprameters for tuning using tune()
+elastic_spec <- logistic_reg(mode = "classification") %>%
+  set_engine("glmnet") %>%
+  set_mode("classification") %>%
+  set_args(
+    penalty = tune(), #set both penalty and mixture parameters to tune  
+    mixture = tune()
+  )
+# make the tuning grid with the range for penalty and for mixture values
+tune_grid <- expand.grid(
+  penalty = seq(0.001, 0.05, length.out = 20),
+  mixture = c(0.25, 0.5, 0.75)
+)
+
+folds <- vfold_cv(train_data, v = 5, repeats = 5) # define number of folds and repeats for cross validation
+
+# Set up parallel processing for faster tuning
+registerDoParallel()
+
+set.seed(123) # set seed for predictability
+
+# produce elastic net workflow using the tuning spec
+workflow <- workflow() %>%
+  add_recipe(elastic_rec) %>%
+  add_model(elastic_spec) #use the tuned spec for the model
+
+en_tuned <- 
+  workflow %>%
+  tune_grid(
+    resamples = folds, 
+    grid = tune_grid,
+    control = control_grid(verbose = TRUE)
+  )
+
+# View tuning results
+tuned_plot <- autoplot(en_tuned)
+tuned_plot + ggtitle( "Tuned elastic Net Model")
+
+# Collect tuning results
+tuning_results <- en_tuned %>%
+  collect_metrics()
+
+# Filter for ROC AUC metric
+roc_auc <- tuning_results %>%
+  filter(.metric == "roc_auc")
+
+# View the ROC AUC values
+print(roc_auc)
+```
+For the results of the three graphs, we can see that the ideal mixture is likely 0.25.
+The amount of regularization, is best around 0.01, for a mixture of 0.25.
+Because of this, I will select the best model with a mixture of 0.25 and penalty of ~ 0.01.
+
+```{r}
+# select the model based on the best roc auc score
+best_auc <- select_best(en_tuned, metric = "roc_auc")
+
+final_en <- finalize_model( # finalize the model based on the roc auc score
+  elastic_spec,
+  best_auc
+)
+# Split the data into training and testing sets
+set.seed(123)  # for reproducibility
+
+# produce elastic net workflow
+workflow <- workflow() %>%
+  add_recipe(elastic_rec) %>%
+  add_model(final_en)
+
+# implement elastic net workflow
+en_tune_fit <- workflow %>% 
+  fit(data = train_data) # fit to training data
+
+en_predictions_tune <- augment(en_tune_fit, train_data) #make predictions on training data
+print(en_predictions)
+
+# Calculate ROC AUC
+roc_auc <- en_predictions_tune %>%
+  roc_auc(truth = P_UTDHPV, .pred_UTD)
+print(roc_auc)
+```
+The ROC AUC has slightly improved, to 0.335 rather then 0.338. I will now evaluate other performance metrics of the model, including rmse and accuracy to see if the model has improved.
+```{r}
+tuned_metrics <- metrics(truth = P_UTDHPV, estimate = .pred_class, data = en_predictions) # calculate metrics for first fit
+print(tuned_metrics)
+```
+The accuracy of the model is fairly high, at 0.79 or 79%, but we will see how these performance metrics change after using the test data to make predictions.
+```{r}
+# use test data ot make predictions
+en_predictions_test <- augment(en_tune_fit, test_data) #make predictions on testing data
+print(en_predictions_test)
+
+# evaluate roc_auc
+roc_auc <- en_predictions_test %>%
+  roc_auc(truth = P_UTDHPV, .pred_UTD)
+print(roc_auc)
+
+#evaluate accuracy and kap
+tuned_metrics <- metrics(truth = P_UTDHPV, estimate = .pred_class, data = en_predictions_test) # calculate metrics for first fit
+print(tuned_metrics)
+
+# evaluate the F1 score of the model predictions
+f1 <- en_predictions_test %>%
+  f_meas(truth = P_UTDHPV, estimate = .pred_class)
+print(f1)
+```
+The ROC_AUC stayed fairly similar, at 0.365 and the accuracy remained the same, around 0.79 and a kap value of around 0.036. The kappa value is best around 1 or -1, so a kap of 0.036 is not good.
+
+Despite this, I will see if the important predictors are in agreement with our LASSO model using the vip package.
+```{r}
+# Finding importance of variables chosen
+en_vars <- en_tune_fit %>%
+  extract_fit_parsnip() %>% # use parsnip to extract to fit
+  vip::vip() #use vip for visualizing the most important variables
+
+print(en_vars)
+```
+The elastic net model produced the same important variables as the previous lasso model.

### 2024-05-03 Reviews/Santiago-review.qmd
Updated manuscript

---
title: Project Review Template 
author: Arlyn Santiago
date: date-modified
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: HPV vaccination completion rates depending on several socioeconomic factors

Name of project author(s):Rachel Robertson and Kelly Cao

Name of project reviewer:Arlyn Santiago

# Specific project content evaluation
Evaluate the different parts of the project by filling in the sections below.

## Background, Context and Motivation
How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

I believe both authors do a job of explaining the purpose of this project and providing a solid background on HPV.

### Summary assessment (PICK ONE, DELETE THE OTHERS)

* strong contextualization and motivation

## Question description
How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments
I think the question is clear that their aim is to identify the socioeconomic and geographic factors associated with HPV vaccination completion for teenagers aged 13 to 17 in the U.S.

### Summary assessment

* question/hypotheses fully explained

## Data description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

I think the data is well explained, however, I think it would be best to include all variable names and meanings into a codebook excel sheet, and place this excel sheet into the data folder. Or, is this the README file located in raw-data folder?

### Summary assessment

* source and overall structure of data well explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

Maybe I wasn't looking in the right file, but in your EDA-code qmd there wasn't much explanation on how the data was cleaned. Most of your cleaning was done in the cleaning1.R file so it would be best to direct your reader to this file. The following is mentioned, "The following data set has been cleaned.Any observations with NAs has been removed." Maybe it would be best to add comments to your code such as using drop_na() to remove NAs. I believe your plots are helpful.

### Summary assessment

* some weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

The analysis was done properly done. However, it was a bit difficult to understand since the analysis was done over seperate files. It would be best to include all the analysis on one file. It's possible I missed it but you could also add Cross Validation to your LASSO to help the optimal lambda value and ensure the model generalizes well.

### Summary assessment

* strong and reasonable analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

The results are very well presented, however, I would rename some of the variables such as INCPOV1 your plots to the variable meaning (poverty status). For Figure it would be best to rotate your x-axis labels. This website does a good job explaining how (https://stackoverflow.com/questions/1330989/rotating-and-spacing-axis-labels-in-ggplot2)

### Summary assessment

* results are very well presented

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

I was not able to find any discussion/conclusion. I would try to interpret the findings of your random forest model as they gave you guys the best findings compared to the other models.

### Summary assessment
* major parts of discussion missing or wrong

## Further comments

_Add any other comments regarding the different aspects of the project here. Write anything you think can help your classmate improve their project._

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

I think the project is labeled in a very self-explanatory way, however, it would be best to place your data cleaning in the EDA file or mention the data cleaning within the EDA file. Also, the analysis should be placed within one file for ease of access.

### Summary assessment

* mostly clear, but some confusing parts (e.g. useless files, things in the wrong folders)

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments
I think enough information was provided to understand each step of the analysis.

### Summary assessment

* fully and well documented

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

I was able to fully run everything.

### Summary assessment

* fully reproducible without issues

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

If both of guys are able to fully address the hypothesis and explain your findings, then you should have an through overall study.

### Summary assessment

* decent level of thoroughness

## Further comments

You guys have done a great job! I would try and add more information on vaccine bias, especially within marginalized groups and how the findings of this study might combat that in your Discussion.

### 2024-05-03 Reviews/project-review-by-mutsa.qmd
Updated manuscript

--- 2024-05-03 Reviews/Santiago-review.qmd
@@
 ---
-title: Project Review Template 
-author: Arlyn Santiago
-date: date-modified
+title: Project Review Part 5 
+author: Mutsa Nyamuranga
+date: 04-25-2024
 format: 
@@
 
-Title of project: HPV vaccination completion rates depending on several socioeconomic factors
+Title of project: Teenage HPV Vaccination Coverage and Socioeconomic Factors
 
-Name of project author(s):Rachel Robertson and Kelly Cao
+Name of project author(s): Kelly Cao, Rachel Robertson
 
-Name of project reviewer:Arlyn Santiago
+Name of project reviewer: Mutsa Nyamuranga
+
+# Instructions
 
@@
 
-I believe both authors do a job of explaining the purpose of this project and providing a solid background on HPV.
+The background was very thorough with multiple citations and necessary cotectual evidence.
 
@@
 ### Feedback and Comments
-I think the question is clear that their aim is to identify the socioeconomic and geographic factors associated with HPV vaccination completion for teenagers aged 13 to 17 in the U.S.
+
+The research question was clearly stated and explained.
 
@@
 
-* question/hypotheses fully explained
+* question/hypotheses fully clear
 
@@
 
-I think the data is well explained, however, I think it would be best to include all variable names and meanings into a codebook excel sheet, and place this excel sheet into the data folder.
-Or, is this the README file located in raw-data folder?
+Data source was clearly described. There was a codebook provided.
 
@@
 
-Maybe I wasn't looking in the right file, but in your EDA-code qmd there wasn't much explanation on how the data was cleaned.
-Most of your cleaning was done in the cleaning1.R file so it would be best to direct your reader to this file.
-The following is mentioned, "The following data set has been cleaned.Any observations with NAs has been removed." Maybe it would be best to add comments to your code such as using drop_na() to remove NAs.
-I believe your plots are helpful.
+Data was well processed. Detailed comments/notes were provided for all procedures.
+There was no supplementary material.
 
@@
 
-* some weaknesses in wrangling and exploratory component
+* essentially no weaknesses in wrangling and exploratory component
 
@@
 
-The analysis was done properly done.
-However, it was a bit difficult to understand since the analysis was done over seperate files.
-It would be best to include all the analysis on one file.
-It's possible I missed it but you could also add Cross Validation to your LASSO to help the optimal lambda value and ensure the model generalizes well.
+The analysis methods were appropriate for the question to be answered.
+Cross-validation was used for performance evaluation and there were multiple models produced and tested to gather the appropriate metrics.
 
@@
 
-The results are very well presented, however, I would rename some of the variables such as INCPOV1 your plots to the variable meaning (poverty status).
-For Figure it would be best to rotate your x-axis labels.
-This website does a good job explaining how (https://stackoverflow.com/questions/1330989/rotating-and-spacing-axis-labels-in-ggplot2)
+The results were presented in a well-structured manner.
 
@@
 
-I was not able to find any discussion/conclusion. I would try to interpret the findings of your random forest model as they gave you guys the best findings compared to the other models.
+There was no discussion or conclusion posted
 
 ### Summary assessment
+
 * major parts of discussion missing or wrong
@@
 
-_Add any other comments regarding the different aspects of the project here. Write anything you think can help your classmate improve their project._
+There were some edits that I made to allow code to flow, however, overall the project is well done and very comprehensive.
 
@@
 
-I think the project is labeled in a very self-explanatory way, however, it would be best to place your data cleaning in the EDA file or mention the data cleaning within the EDA file.
-Also, the analysis should be placed within one file for ease of access.
+The project is generally well-structured. All files are in well labeled folders. It is easy to figure out the use for each file/folder.
 
@@
 
-* mostly clear, but some confusing parts (e.g. useless files, things in the wrong folders)
+* well structured
 
@@
 ### Feedback and Comments
-I think enough information was provided to understand each step of the analysis.
+
+The R scripts are well-documented with sufficient description. Each decision of analysis was justified and explained. The Rmd files(README) are well utilized.
 
@@
 
-I was able to fully run everything.
+I was able reproduce the results. Instructions were clear.
 
@@
 
-* fully reproducible without issues
+* small parts not reproducible or required manual intervention
 
@@
 
-If both of guys are able to fully address the hypothesis and explain your findings, then you should have an through overall study.
+The research questions have been fully addressed. All sections are relatively complete and thorough.
 
@@
 
-* decent level of thoroughness
+* strong level of thorougness
 
@@
 
-You guys have done a great job! I would try and add more information on vaccine bias, especially within marginalized groups and how the findings of this study might combat that in your Discussion.
+Overall, this study is in good shape. The authors may want to consider touching up code within some of the R scripts, however they do not take away from how well the project was done.

### 2024-05-03 code/analysis-code/analysis1.qmd
Updated manuscript

@@
 
+# Save the plot as an image
+file_path = here("results","figures", "residual_plot.png")
+ggsave(filename = file_path, plot = residual_plot_logit)
 ```

### 2024-05-03 code/analysis-code/analysis2.5.qmd
Updated manuscript

--- 2024-04-18 code/analysis-code/analysis2.5.qmd
@@
 
-print(rf_vars)
+# Modify the plot to set the color of the bars
+rf_vars_plot <- rf_vars+
+  aes(fill = "skyblue")+
+  labs(title = "Top Predictors")
+
+# Original variable names and their translations
+variable_translations <- data.frame(
+  Original_Variable = c("MOBIL_1", "SEX", "RENT_OWN", "INS_STAT2_I", "EDUC1", 
+                         "RACEETHK", "AGE", "FACILITY", "INCQ298A", "STATE"),
+  Translated_Variable = c("Geographic Mobility", "Sex", "Rental Status", 
+                          "Insurance Status", "Mother's Education Status", "Ethnicity", "Age", 
+                          "Facility", "Income", "State")
+)
+
+# Add translations to the plot
+rf_vars_plot_with_translations <- rf_vars_plot +
+  scale_fill_manual(values = c("skyblue")) +  # Set fill color manually
+  scale_x_discrete(labels = variable_translations$Translated_Variable) +  # Translate x-axis labels
+  labs(title = "Top Predictors",
+                x = "Variable")  # Add title and update x-axis label
+
+# Display the plot with translated variable names
+print(rf_vars_plot_with_translations)
+
+# Save the plot as a PNG file
+file_path = here("results", "figures", "variable_importance_plot.png")
+ggsave(filename = file_path, plot=rf_vars_plot_with_translations) 
 ```

### 2024-05-03 code/analysis-code/analysis2.qmd
Updated manuscript

@@
 library(rsample)
-
+library(knitr)
+library(kableExtra)
+library(webshot)
+library(here)
+library(gt)
+library(gtsummary)
 ```
@@
 
+```
+
+```{r}
+# Extract the components from the PCA model
+standard_deviation <- pca_model@model$model_summary$pc1[1]
+proportion_variance <- pca_model@model$model_summary$pc1[2]
+cumulative_proportion <- pca_model@model$model_summary$pc1[3]
+
+# Create a data frame with the PC information
+scree_data <- data.frame(
+  Component = "PC1",
+  Standard.deviation = standard_deviation,
+  Variance = standard_deviation^2,
+  Proportion.Variance = proportion_variance,
+  Cumulative.Proportion = cumulative_proportion
+)
+
+# Print the summary table
+print(scree_data)
+
+# Create a table using kableExtra
+pca_table <- 
+  scree_data %>%
+  gt() %>%
+  tab_header(
+    title = "PCA Summary Table"
+  )
+
+# Save the table to a file
+gtsave(pca_table, here("results","tables", "pca_summary.png"))
 ```
@@
 
+# Function to extract metrics from a model
+extract_metrics <- function(model, dataset) {
+  metrics <- model@model$training_metrics@metrics
+  mse <- metrics$MSE
+  r2 <- metrics$r2
+  logloss <- metrics$logloss
+  mpce <- metrics$mean_per_class_error
+  auc <- metrics$AUC
+  gini <- metrics$Gini
+  nulldev <- metrics$null_deviance
+  resdev <- metrics$residual_deviance
+  aic <- metrics$AIC
+
+  # Create a data frame for metrics
+  metrics_df <- data.frame(
+    Metric = c("MSE", "R^2", "Log Loss", "Mean Per-Class Error", "AUC", "Gini", "Null Deviance", "Residual Deviance", "AIC"),
+    Value = c(mse, r2, logloss, mpce, auc, gini, nulldev, resdev, aic),
+    Dataset = dataset
+  )
+
+  return(metrics_df)
+}
+
+# Extract metrics from the LASSO model (training data)
+train_metrics_df <- extract_metrics(lasso_model, "Training")
+
+# Extract metrics from the CV metrics (test data)
+test_metrics_df <- extract_metrics(cv_metrics, "Test")
+
+# Combine the training and test metrics data frames
+combined_metrics_df <- rbind(train_metrics_df, test_metrics_df)
+
+# Create a gt table
+gt_table <- combined_metrics_df %>%
+  gt(
+    groupname_col = "Dataset", 
+    row_group_as_column = TRUE
+  ) %>%
+  tab_header(
+    title = md("Model Evaluation Metrics"),
+    subtitle = md("Comparison of Metrics between Test and Training Datasets")
+  )
+
+# Save the gt table
+gtsave(gt_table, here("results", "tables", "LASSO_metrics.png"))
+
 ```

### 2024-05-03 code/analysis-code/analysis3.qmd
Updated manuscript

@@
 library(vip)
+library(gt)
 ```
@@
 
-print(lasso_vars)
+# Modify the plot to set the color of the bars
+lasso_vars_plot <- lasso_vars+
+  aes(fill = "skyblue")+
+  labs(title = "Top Predictors")
+
+# Create a data frame with original and translated variable names
+variable_translations <- data.frame(
+  Original_Variable = rev(c("STATE_RHODE.ISLAND", "FACILITY", 
+                        "STATE_VERMONT", "STATE_MISSISSIPPI", "STATE_NORTH.DAKOTA", "STATE_IOWA", 
+                        "STATE_HAWAII", "RENT_OWN_DON.T.KNOW", "STATE_NEBRASKA", 
+                        "INS_BREAK_I_CURRENTLY.UNINSURED.AND.NEVER.INSURED.SINCE.AGE.11")),
+  Translated_Variable = rev(c("Rhode Island", "All Std School Teen Clinics or Other Facilities", 
+                          "Vermont", "Mississippi", "North Dakota", "Iowa", "Hawaii", 
+                          "Rent/Own Status (Don't Know)", "Nebraska", 
+                          "Uninsured"))
+)
+
+# Add translations to the plot
+lasso_vars_plot_with_translations <- lasso_vars_plot +
+  scale_fill_manual(values = c("skyblue")) +  # Set fill color manually
+  scale_x_discrete(labels = variable_translations$Translated_Variable) +  # Translate x-axis labels
+  labs(title = "Top Predictors",
+                x = "Variable")+  # Add title and update x-axis label
+  scale_y_continuous(labels = function(x) strwrap(x, width = 10))
+
+# Display the plot with translated variable names
+print(lasso_vars_plot_with_translations)
+
+# Save the plot as a PNG file
+path = here("results", "figures", "top_lasso_predictors.png")
+ggsave(filename = path, plot=lasso_vars_plot_with_translations) 
 ```

### 2024-05-03 code/eda-code/EDA-code.qmd
Updated manuscript

@@
 ```{r}
-p1 <- mydata %>% ggplot(aes(x=INCPORAR_I)) + geom_histogram() 
-plot(p1)
+p1 <- mydata %>% 
+  ggplot(aes(x = INCPORAR_I)) +  
+  geom_histogram(binwidth = 1, color = "black", fill = "skyblue") +  # Adjust binwidth as needed
+  xlab("Income to Poverty Ratio") +                  
+  ylab("Count") +                                    
+  ggtitle("Income-to-Poverty Ratio Variable Distribution") +
+  theme_minimal() +  # Apply a minimal theme for cleaner appearance
+  theme(
+    plot.title = element_text(size = 14, face = "bold"),  # Customize title font size and style
+    axis.title.x = element_text(size = 12),              # Customize x-axis label font size
+    axis.title.y = element_text(size = 12),              # Customize y-axis label font size
+    axis.text = element_text(size = 10),                 # Customize axis text font size
+    legend.position = "none",                            # Remove legend if not needed
+    panel.grid.major = element_line(color = "lightgray"), # Add light gray grid lines
+    panel.grid.minor = element_blank()                   # Remove minor grid lines
+  ) +
+  labs(caption = "(Originally INCPORAR_I)")
+
+p1
 figure_file = here("results", "figures", "income.png")
@@
 p5 <- ggplot(family_income_df, aes(x = Var1, y = Freq)) +
-  geom_bar(stat = "identity", fill = "skyblue") +
-  labs(title = "Family Income Distribution Percentage",
+  geom_bar(stat = "identity", fill = "skyblue", color = "black") +  # black borders
+  labs(title = "Household Income Variable Distribution",
        x = "Income",
-       y = "Percentage")
+       y = "Percentage (%)") +
+  theme_minimal() +  # Minimalistic theme
+  labs(caption = "(Originally INCQ298A)") +
+  theme(axis.text.x = element_text(angle = 90, hjust = 1),  # Rotate x-axis labels
+        axis.line = element_line(color = "black"),           # Black axes
+        panel.grid.major = element_blank(),                  # Remove major gridlines
+        panel.grid.minor = element_blank(),                  # Remove minor gridlines
+        panel.border = element_blank())                      # Remove panel borders
 
@@
        y = "Percentage") +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
+  theme(axis.text.x = element_text(angle = 45, vjust = 0.5))
 
@@
 p18 <- ggplot(mydata, aes(x = P_UTDHPV, fill = FACILITY)) +
-  geom_bar(stat = "count") +  # Use geom_bar() with the count of each facility
-  labs(x = "Vaccination Status", y = "Count", title = "Facility Vaccination Completion") +
+  geom_bar(stat = "count") +  
+  labs(x = "Vaccination Status", y = "Count", caption = "UTD = Up-to-Date", title = "Facility Vaccination Completion") +
   theme_minimal()
@@
 p17 <- mydata %>% 
-  ggplot(aes(x=INS_STAT2_I, y=INCPORAR_I, color = P_UTDHPV)) +  
+  ggplot(aes(x = INS_STAT2_I, y = INCPORAR_I, color = P_UTDHPV)) +  
   geom_boxplot() +
-  labs(x = "Insurance status", y = "Income-poverty ratio", title = "Income-poverty ratio stratified by insurance and vaccination") +
-  scale_x_discrete(labels = c("PRIVATE INSURANCE ONLY" = "Private", "ANY MEDICAID" = "Medicaid", "OTHER INSURANCE (CHIP, IHS, MILITARY, OR OTHER, ALONE OR IN COMB.
-WITH PRIVATE INSURANCE)" = "Other", "UNINSURED" = "Uninsured")) +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))  # Rotate x-axis text by 90 degrees
-plot(p17)
+  labs(x = "Insurance status",
+       y = "Income-to-Poverty ratio",
+       title = "Income-to-Poverty Ratio Stratified by Insurance and Vaccination Status", 
+       color = "Vaccination Status", 
+       caption = "UTD refers to Up-to-Date vaccination status") +  
+  scale_x_discrete(
+    labels = c("PRIVATE INSURANCE ONLY" = "Private",
+               "ANY MEDICAID" = "Medicaid",
+               "OTHER INSURANCE (CHIP, IHS, MILITARY, OR OTHER, ALONE OR IN COMB. WITH PRIVATE INSURANCE)" = "Other",
+               "UNINSURED" = "Uninsured")) +
+  theme(axis.text.x = element_text(angle = 0, vjust = 0.5))  # Rotate x-axis text by 90 degrees
+
+print(p17)
+
 figure_file = here("results","figures", "insurance-income-stratified.png")
@@
 p19 <- mydata %>% 
-  ggplot(aes(x=INCPORAR_I, y=RACEETHK, color = P_UTDHPV)) +  geom_boxplot()
+  ggplot(aes(x=INCPORAR_I, y=RACEETHK, color = P_UTDHPV)) +
+  geom_boxplot()+
+  labs(title = "HPV Vaccination Status",
+       x = "Income-to-Poverty Ratio",
+       y = "Race/Ethnicity", 
+       caption = "A ratio equal to 1, suggests income and poverty levels are the same.", 
+       color = "Vaccination Status")
 plot(p19)

### 2024-05-03 products/manuscript/Manuscript.qmd
Updated manuscript

@@
 
+# Abstract 
+Human Papillomavirus (HPV) is thought to be the most common sexually transmitted disease in the United States [2].
+Though HPV consists of a group of over 200 related viruses, only 12 types of HPV are considered “high-risk” and associated with the incidence of six types of cancer, including oropharyngeal, cervical, penile, vulvar, vaginal, and anal cancers [4,5].
+The U.S.
+currently recommends HPV vaccination with Gardasil 9 to immunize against six common types of high-risk HPV infections and three HPV types associated with genital warts.
+It is recommended that a primary dose of the HPV vaccine be given between the ages of 11 to 12 and a secondary dose administered after 6 months to 1 year from the primary [6].
+Although Healthy People 2023 has set the goal of 80% HPV vaccination coverage in the U.S., only 62.6% of U.S.
+teenagers are up-to-date on their HPV vaccination [7].
+We are interested in examining which socioeconomic and geographic factors impact HPV vaccination completion among teenagers in the U.S.
+To address this question, we decided to examine 2022 survey data from the National Immunization Survey for U.S.
+Teenagers.
+This survey is distributed to both providers and households of teenagers and contains a questionnaire regarding vaccination completion for several common vaccines in the U.S.
+immunization schedule, including HPV vaccination status.
+Additionally, the survey gathers information on demographics, socioeconomic status, and geographic location of the teenage patient.
+Using this data, we examined how several socioeconomic and demographic factors impact the status of up-to-date HPV vaccination and determined which predictors were most important in determining the up-to-date HPV vaccination status.
+R version 4.2.3 with R Studio was used to clean, process, explore, and analyze the survey data.
+Multi-step logistic regression, principal component analysis (PCA), LASSO regression, a random forest model, and an elastic net model were employed to analyze which factors are most important for determining HPV vaccination completion.
+The best-performing model, the random forest model, was chosen to determine the most important factors for this prediction, while the LASSO and elastic net models were used to provide a direction of association between each predictor and the outcome of HPV vaccination completion status.
+It was found that state is the most important factor in determining up-to-date HPV vaccination status, followed by family income, facility, age, and race or ethnicity.
+A high family income bracket and non-Hispanic white ethnicity were negatively associated with HPV vaccination completion while older age groups, minority ethnic groups, and lower family income were positively associated with this outcome.
+The relationship between state and HPV vaccination status depends on the state, which differs by legislation, cultural norms, and funding for HPV vaccination.
+Our results highlight the need for consistent HPV vaccination state funding and school mandates in combination with provider education to teenage patients and their parents about the efficacy and safety of the HPV vaccination.
+
 # Introduction
@@
 
-Human Papillomavirus (HPV) is a double-stranded DNA virus that infects the cells of the skin and mucosal linings \[1\].
-HPV is the most common sexually transmitted infection (STI) in the United States, where over 90% of sexually active males and 80% of sexually active females will likely become infected within their lifetime \[2, 3\].
+Human Papillomavirus (HPV) is a double-stranded DNA virus that infects the cells of the skin and mucosal linings [1].
+HPV is the most common sexually transmitted infection (STI) in the United States, where over 90% of sexually active males and 80% of sexually active females will likely become infected within their lifetime [2, 3].
 Most cases of HPV are asymptomatic and cleared over time, but certain infections tend to become persistent and can cause genital warts.
-Specific types of HPV are known to be “high-risk” or increase the likelihood of developing cancer in persistently infected cells \[4\].
+Specific types of HPV are known to be “high-risk” or increase the likelihood of developing cancer in persistently infected cells [4].
 The HPV group contains over 200 related viruses, but only 12 types are considered “high-risk” due to the presence of oncogenes.
 These include HPV 16, 18, 31, 33, 35, 39, 45, 51, 52, 56, 58, and 59, though HPV-16 and HPV-18 are associated with the highest number of HPV-associated cancer cases.
-In the U.S., there are 37,300 cases of HPV-associated cancer cases every year, including cases of cervical, anal, oropharyngeal (throat and mouth), vulvar, vaginal, and penile cancers \[5,6\].
+In the U.S., there are 37,300 cases of HPV-associated cancer cases every year, including cases of cervical, anal, oropharyngeal (throat and mouth), vulvar, vaginal, and penile cancers [5,6].
 In an effort to prevent HPV-associated cancers worldwide, two HPV vaccines were developed and made available in 2006.
-These included a bivalent and quadrivalent vaccine that contained viral-like particles (VLPs) of the viral capsid protein (L1) from “high-risk” HPV types \[7\].
-In recent years, the vaccine used in the U.S., Gardasil 9, has been expanded to contain HPV types 6, 11, 16, 18, 31, 33, 45, 52, and 58 \[4\].
-Currently, the United States Centers for Disease Control and Prevention (CDC) recommends a two-dose Gardasil 9 regime to all teenagers, with a primary dose at age 11 to 12 and a second dose within 6 to 12 months of the first \[8\].
-In 2022, the CDC estimated that 76.0% of teenagers aged 13 to 17 had received at least 1 dose of the HPV vaccine while only 62.6% of teenagers were up-to-date with their vaccination regime \[9\].
+These included a bivalent and quadrivalent vaccine that contained viral-like particles (VLPs) of the viral capsid protein (L1) from “high-risk” HPV types [7].
+In recent years, the vaccine used in the U.S., Gardasil 9, has been expanded to contain HPV types 6, 11, 16, 18, 31, 33, 45, 52, and 58 [4].
+Currently, the United States Centers for Disease Control and Prevention (CDC) recommends a two-dose Gardasil 9 regime to all teenagers, with a primary dose at age 11 to 12 and a second dose within 6 to 12 months of the first [8].
+In 2022, the CDC estimated that 76.0% of teenagers aged 13 to 17 had received at least 1 dose of the HPV vaccine while only 62.6% of teenagers were up-to-date with their vaccination regime [9].
 
@@
 This serves as valuable information to determine differences in healthcare access and vaccination coverage for teenagers living in the U.S.
+
 It is generally known that those of lower income and without insurance have limited healthcare access in the U.S.
-\[10\], but it will be interesting to see if this pattern is reflected at the state level.
+[10], but it will be interesting to see if this pattern is reflected at the state level.
 If so, policy should be examined as it may be related to healthcare access disparities.
-Recent studies have examined HPV vaccination completion according to social determinants, but they have not examined geographic distribution and include data prior to 2019 \[11\].
-More recent years display differences in global HPV vaccination trends due to the COVID19 pandemic \[12\].
+Recent studies have examined HPV vaccination completion according to social determinants, but they have not examined geographic distribution and include data prior to 2019 [11].
+More recent years display differences in global HPV vaccination trends due to the COVID19 pandemic [12].
 Our study will fill in geographical gaps and provide updated vaccination trends based on socioeconomic factors.
-
-{{< pagebreak >}}
 
@@
 The data was loaded from the original DAT file, entitled 'NISTEENPUF22.DAT' by specifying the column positions within the data file.
+This is performed with the `readr` library and its `fwf_position` function.
 The columns were then read and made into a data frame.
-Each of the factor variables was defined and the corresponding levels were assigned labels.
+Each of the factor variables was defined and the corresponding levels were assigned labels, based on the codebook.
+This is performed with the `dplyr` library and its `mutate` function.
 The structure and summary statistics of all of the variables were examined.
@@
 ## Cleaning Data 
-
 The structure of the processed data was examined.
@@
 We saved the cleaned data as an rds and csv in the 'processed-data' folder.
-## Exploratory Analysis Exploratory analysis was used to determine the distribution of the predictors, including the socioeconomic variables that were chosen for analysis.
+
+## Exploratory Analysis
+
+Exploratory analysis was used to determine the distribution of the predictors, including the socioeconomic variables that were chosen for analysis.
 The distribution of single predictors including income-to-poverty ratio, income, state, race, ethnicity, vaccination status, and facility were examined using histograms and scatter plots.
 The following predictors were compared visually and fitted to a linear regression to examine preexisting associations: vaccination status and ethnicity, vaccination status and poverty status, income poverty ratio and insurance status, income poverty ratio and ethnicity.
-## Analysis \### Generalized Logistic Regression A generalized logistic regression was performed with bi-directional step-wise selection for predictor reduction.
+
+## Analysis
+
+### Generalized Logistic Regression
+
+A generalized logistic regression was performed with bi-directional step-wise selection for predictor reduction.
 Variables that were redundant of a major variable were excluded from analysis.
@@
 This is the basis for the subsequent analysis methods, as they are more resistant to collinearity.
-### Principal Component Analysis Principal component analysis (PCA) was performed to do dimensionality reduction.
+
+### Principal Component Analysis
+
+Principal component analysis (PCA) was performed to do dimensionality reduction.
 PCA would ideally transform the highly correlated predictors into uncorrelated components.
@@
 A LASSO Ridge Regression was performed with the h2o package, to mitigate the multicollinearity issue.
-### Random Forest Model A random forest model was used, as it is resistant to multicollinearity.
+
+### Random Forest Model
+
+A random forest model was used, as it is resistant to multicollinearity.
 The tidymodels package was used to produce a random forest model with the classification method.
@@
 Finally, the tuned random forest model was re-evaluated by being fit to the testing data.
+
 ### LASSO Regression
 
-{{< pagebreak >}}
+A LASSO model was produced as it is resistant to multicollinearity because it implements regularization and variable selection, by reducing unimportant coefficients to zero.
+Tidymodels in R was used to produce a logistic regression LASSO model using the glmnet engine.
+The model recipe included all of the selected predictors from the stepwise selection and was fit to the training data split from 80% of the original data.
+The penalty parameter was tuned from a grid using cross-validation and the ideal penalty was selected using the model that maximized ROC AUC.
+The tuned model was extracted and re-fit to the training data.
+The metrics accuracy, F1, and confusion matrix were compared between tuned and original fit models.
+The tuned model was fit to the testing data and the most important variables were extracted using the vip package.
+
+### Elastic Net Regression
+
+An elastic net regression was produced using the tidymodels framework in R using the glmnet engine set to classification mode.
+The initial elastic net model included all of the variables selected by the step-wise selection above.
+The model was then tuned for the parameters penalty and mixture using a cross-validation grid search.
+The ideal model was selected by maximizing ROC AUC for both of the tuning parameters.
+The final model was fit to the test data and re-evaluated with the metrics: accuracy and kappa.
+The most important variables were extracted using the vip package.
 
 # Results
+
 ## Exploratory Data Analysis
+
 Exploratory analysis is used to determine the distribution of our predictors, including the socioeconomic variables that we chose for analysis.
@@
 ```{r}
+
 #| label: Table1
+
 #| tbl-cap: "Table 1.Socioeconomic characteristics of U.S. teenagers who have completed or are up-to-date with HPV vaccination series."
+
 #| echo: FALSE
+
 Table1 <- readRDS(here("results/tables/table1.rds"))
+
 kable(Table1)
+
 ```
@@
 This indicates that a high frequency of participants are above the poverty income threshold, but also in a modest income class.
-![Figure 1. Income-poverty ratio distribution](./images/income.png)
 
-![Figure 2. The family income distribution](./images/family.income.distribution.png)
+![Figure 1. Income-poverty ratio distribution](../../results/figures/income.png)
+
+![Figure 2. The family income distribution](../../results/figures/family.income.distribution.png)
 
@@
 is 58.9% white, 13.6% African American, and 19.1% Hispanic or Latino, while our survey population is composed of 58.% white, 9.15% African American, and 19.9% Hispanic or Latino participants.
+
 The majority of observations came from households with private insurance only The Insurance status variable shows that the majority of the observations came from households with private insurance only (58.6%), followed by Medicaid (31.6%), other insurance (7.7%), and uninsured (2.2%).
@@
 They likely received their HPV vaccination there as well.
-![Figure 3.Facility and vaccine completion distribution](./images/facility-vaccination.png)
+
+![Figure 3.Facility and vaccine completion distribution](../../results/figures/facility-vaccination.png)
 
@@
 This may be accounted for by other trends such as that shown with ethnicity, above.
-![Figure 9. Distribution of income poverty ratio stratified by race and HPV vaccination completion status](./images/raceeth-income-stratified.png)
+
+![Figure 4. Distribution of income poverty ratio stratified by race and HPV vaccination completion status](../../results/figures/raceeth-income-stratified.png)
 
@@
 For those with Medicaid, those who were up-to-date in HPV vaccinations had a slightly lower income-poverty ratio than those who were not up to date.
-![Figure 11. Income poverty ratio and insurance status](./images/insurance-income-stratified.png)
+
+ ![Figure 5. Income poverty ratio and insurance status](../../results/figures/insurance-income-stratified.png)
 
 ## Analysis
+
 The variables selected for analysis were chosen based on their interpretability and applicability to our research question “Which socioeconomic and geographic factors impact HPV vaccine completion rates among teenagers?” Any variables that capture repetitive data were removed and the most important or straightforward version of the variable was kept in the model.
 Seven variables represented data for the outcome of HPV vaccination completion:
-![Table 3. Human Papillomavirus vaccine status variables and meanings.](./images/UTDHPV-variable-table.png)
+
+![Table 2. Human Papillomavirus vaccine status variables and meanings.](./images/UTDHPV-variable-table.png)
 
@@
 The variables that remained in the initial logistic regression are listed below.
-![Table 4. Potential predictor variable names and meanings.](./images/Predictor-variable-table.png)
+
+![Table 3. Potential predictor variable names and meanings.](./images/Predictor-variable-table.png)
 
@@
 RACEETHK is ideal because it includes both race and ethnicity.
+
 ### Stepwise Selection with Logistic Regression
+
 To determine which predictor variables will be best suited for a general logistic model, we employed a bi-directional stepwise selection method.
@@
 From this, we were able to determine that the model with the lowest AIC (and thus the greatest fit) was the model that included all of the original factors.
+
 The logistic regression with the selected variables was assessed by plotting the residuals.
@@
 Multi-collinearity was present between the majority of the predictor variables, thus another model type that could handle collinearity would be ideal.
-INCLUDE EITHER REGRESSION PLOT OR STEPWISE FINAL RESULT IMAGE
+
+![Figure 6. Residual Plot Results](../../results/figures/residual_plot.png)
+
 ### Principal Component Analysis and LASSO Regression
+
 In order to reduce the multicollinearity issue, we initially attempted to use PCA.
@@
 The PCA resulted in one principle component with a high standard deviation and high proportion of variance, showing that the Pc1 captures a large portion (97.9%) of variability in the data.
-INCLUDE IMAGE OF PCA RESULTS
+
+![Table 4. Summary of PCA Attempt](../../results/tables/pca_summary.png)
+
 The result of having only one principle component further supports the fact that many of the predictor variables are highly collinear.
@@
 Using the same package, I used the LASSO logistical regression model and performed a co-validation test on it.
-```
-MSE:  0.1558815
-RMSE:  0.3948183
-LogLoss:  0.4850906
-Mean Per-Class Error:  0.4906615
-AUC:  0.6596287
-AUCPR:  0.8704217
-Gini:  0.3192575
-R^2:  0.05771195
-Residual Deviance:  16070.08
-AIC:  16266.08
-```
+
+![Table 5. Summary of Initial LASSO Regression](../../results/tables/LASSO_metrics.png){height=500}
+
 ### Random Forest Model
+
 We used a Random Forest Model to assess which predictors are associated with HPV vaccination completion rates in U.S.
@@
 Decision tree models are known to be able to handle collinearity and predictors with different trends.
-A random forest model is ideal because it employs bootstrapping to produce randomly chosen iterations of decision trees to introduce variability.
+A random forest model is ideal because it employs bootstrapping to produce randomly chosen iterations of decision trees, which introduces variability into the models.
 The best random forest model can then be tuned for and selected based on which tree performs best in a certain metric.
@@
 The model with the best ROC AUC metric was selected, and predictions were produced.
-The model indicated that the most important predictors of HPV vaccination completion, in order of most to least important, are state, income level, facility, age, race/ ethnicity, insurance status, education, housing status, sex, and geographic mobility.
-[insert plot  of most important predictors from vip package; not sure how, because it is an environment and I don’t know how to extract one plot]
-However, after testing the same metrics with new data (testing data), the F1 metric was reduced from 0.86 to 0.16, which could indicate overfitting or that the model is “too big” for the data.
+This model initially indicated good performance with accuracy and F1 measure of 0.95 and 0.86 respectively.
+A confusion matrix estimated an out-of-bag (OOB) error rate of 23.0%, a specificity of 99.4%, and a sensitivity of 76.6%.
+The most important predictors for HPV vaccination completion from this model, in order of most to least important, were state, income level, facility, age, race/ ethnicity, insurance status, education, housing status, sex, and geographic mobility.
+
+![Figure 7. Top Predictors Chosen by Random Forest Model](../../results/figures/variable_importance_plot.png)
+
+However, after testing the same metrics with new data (testing data), the F1 metric was reduced from 0.86 to 0.16 and the accuracy decreased from  0.95 to 0.78, which could indicate overfitting or that the model is “too big” for the data.
+To address the concern of overfitting, a LASSO model was utilized.
 
 #### LASSO Regression with Tidymodels
+
+The LASSO regression was performed using the Tidymodels framework.
+A LASSO regression is ideal for multicollinearity as it not only regularizes the variables but also performs variable selection by reducing the coefficients of unimportant variables to zero.
+The LASSO model was tuned for the penalty parameter and selected based on the best ROC AUC performance.
+The tuned model improved in performance from the initial model, with an initial ROC AUC of 0.37 and a tuned ROC AUC of 0.34.
+The accuracy of the initial and tuned model remained the same, at 0.79.
+The F1 measure increased from 0.03 to 0.07 from the initial to the tuned model.
+The specificity and sensitivity of the tuned model were calculated from the confusion matrix, to be 98.7% and 3.4%, respectively.
+This means the model is highly likely to produce false negatives or predict that HPV vaccination was not completed when in reality it was.
+The most important predictors, extracted from the tuned model from most to least important, were state (Rhode Island), facility (STD, school, or teen clinics), state (Vermont), state (Mississippi), state (North Dakota), state (Iowa), state (Hawaii), housing status (unknown), state (Nebraska), and insurance breaks (currently uninsured or never insured since 11 years of age).
+
+![Figure 8. Top Predictors chosen by LASSO Regression Model](../../results/figures/top_lasso_predictors.png)
+
+#### Elastic Net Regression
+
+In an attempt to produce a better-performing model compared to the LASSO regression, an elastic net regression was produced using the Tidymodels framework.
+The elastic model employs methods from both the LASSO regression and Ridge regression by implementing a penalty parameter and mixture parameter.
+The model was fit initially and then tuned for both the penalty and mixture parameters using cross-validation to maximize ROC AUC.
+The initial fit elastic model and tuned model both performed with an ROC AUC of 0.34.
+The tuned model had an accuracy and kappa measure of 0.79 and 0.06, respectively.
+After fitting the tuned model to testing data, the accuracy and kappa were reduced to 0.79 and 0.4 respectively.
+The ROC AUC for this fit increased to 0.36 and the F1 measure was 0.07.
+The performance of the elastic net model was very similar to the LASSO model.
+Neither the LASSO nor elastic net model performed as well as the random forest tree model, despite the overfitting of the random forest model.
+The most important predictors chosen by the elastic net model were the same as the predictors chosen by the LASSO regression.
+The most important predictors in determining HPV vaccination completion, from most to least important, were state (Rhode Island), facility (std, school, and teen clinics), state (Vermont), state (Mississippi), state (North Dakota), state (Iowa), state (Hawaii), housing status (unknown), insurance breaks (currently uninsured or never insured since 11 years of age), and state (Nebraska).
+
 # Discussion
 
+The random forest model was chosen to predict HPV vaccination completion, as it was the best-performing model for all of the performance metrics, despite potential overfitting to the data.
+The other two machine learning models, the LASSO and elastic net models, may also provide insight into which predictors are most important for determining HPV vaccination completion.
+These models are mainly in agreement with each other and display similar variable importance rankings as the random forest model.
+The findings will be interpreted by examining which variables were chosen by all three models, as they may hold the most importance, followed by the predictors chosen by the random forest model.
+
+ For all three model types that were tested, state was chosen to be the most important factor in determining HPV vaccine completion status.
+Both the LASSO regression and elastic net models determined that Rhode Island was the state most closely associated with HPV vaccine completion, followed by Vermont, Mississippi, North Dakota, Iowa, Hawaii, and lastly, Nebraska.
+For both of these models, all of these states have positive coefficients associated with the outcome of up-to-date HPV vaccination, except for Mississippi, which has a negative coefficient.
+This may be explained by the fact that each state determines HPV vaccination legislation, policy, and funding, for teenagers aged 11 to 13.
+As of 2021, four states have a law requiring vaccination for school entry, including,  Hawaii, Rhode Island, Virginia, and Washington D.C (15).
+As of 2019, four states recommend HPV vaccination for school attendance, by law, including, Arkansas, Delaware, West Virginia, and Wyoming (16).
+As of 2019, 42 U.S.
+states and jurisdictions provide legislative funding to promote HPV vaccination, however, Rhode Island is the only state to implement both a school entry requirement and funding for HPV vaccination (17).
+Additionally, Rhode Island has the highest HPV vaccination coverage out of all U.S.
+states and jurisdictions, exceeding 80% coverage as of 2022 (18).
+The states selected as most important by the models that have legislation to fund HPV vaccination campaigns include Rhode Island, North Dakota, Mississippi, and Iowa (19).
+States chosen as important by the model with legislation mandating HPV vaccination are Rhode Island and Hawaii.
+Neither Vermont nor Nebraska have legislature funding and mandating HPV vaccination, although, both are in the top 15 states for HPV vaccination coverage in the U.S.
+(18,19).
+Conversely, Mississippi has a legislature to reimburse the cost of HPV vaccination to insurance providers but has the lowest HPV vaccination coverage in the country.
+This is thought to be due to misconceptions and misinformation about the HPV vaccine’s purpose and the way that HPV is transmitted (18,20).
+The mixture of policies and vaccination coverage rate, which is not always policy dependent, demonstrates the importance of a combination of legislation to mandate and fund HPV vaccination as well as provider and patient education to support HPV vaccination.
+
+All three models also indicated that facility and housing status are important in determining up-to-date HPV vaccination status.
+According to the LASSO and elastic net models, the facility being an STD clinic, school clinic, or teen clinic, is negatively associated with HPV vaccination completion, while all other facility categories (hospitals, private facilities, mixed facilities) are positively associated with the up-to-date HPV vaccination status.
+This may indicate that STD clinics, school clinics, and/or teen clinics neglect to recommend or educate their teenage patients on HPV vaccination.
+Although school-based HPV vaccination programs have shown promise in increasing HPV vaccine uptakes internationally, these programs in the U.S.
+face barriers such as politicization of vaccines, patient consent processes for teens, and funding for these programs (21,22).
+An unknown housing status was also negatively associated with the outcome of up-to-date HPV vaccination in the LASSO and elastic net models.
+This may reflect that housing security indirectly impacts HPV vaccine uptake in teenagers.
+A cross-sectional study and analysis of house-based socioeconomic status index in the U.S.
+in 2020, indicated that HPV vaccination initiation was related to housing and socioeconomic status, where lower quartiles corresponded to lower rates of HPV vaccine initiation (23).
+
+Along with the variables state and facility, the random forest model indicated family income, age, and race/ ethnicity to be in the top five most important predictors for HPV vaccination completion.
+The LASSO and elastic net models both indicate that family income at the highest three income brackets is negatively associated with HPV vaccination completion, while the lowest three income brackets are positively associated with HPV vaccination completion.
+In contrast to our findings, a CDC data brief found that the percentage of teenagers who received one or more HPV vaccine doses increased as family income moved to be greater than the federal poverty level.
+However, this study does not differentiate vaccination levels at the uppermost income brackets (24).
+A recent study analyzing National Immunization Teen Surveys from 2017 to 2021 found that 64.7% of parents in advantaged income brackets had no intention to begin HPV vaccination while 40.9% of parents in disadvantaged income brackets had no intention to begin the vaccine regime.
+Additionally, the advantaged group was less likely to complete HPV vaccination than the disadvantaged group.
+For the advantaged group, the main reason for not starting the regime was safety concerns and the main reason for not completing the regime was the belief that their child was up-to-date on HPV vaccinations (25).
+This highlights the importance of addressing vaccination misinformation and provider education to parents on the recommended HPV vaccination schedule.
+
+Age was selected as the fourth most important predictor of HPV vaccination completion by the random forest model, but not selected by the LASSO or elastic net models, so the direction of association for this study is unknown.
+A 2024 data brief by the CDC found that the percentage of teenagers who have received at least one dose of the HPV vaccination increases with age, despite the recommended age for beginning the HPV vaccination being 11 to 12 years old in the U.S.
+(24).
+Next, Race and ethnicity were selected as the fifth most important predictor of HPV vaccination completion by the random forest model, and selected by the LASSO and elastic net models, though it was not found to be the top most important predictor by this model.
+The LASSO and elastic net models found that being part of the non-Hispanic ethnic group is negatively associated with the up-to-date HPV vaccination status, while all other racial and ethnic groups are positively associated with the up-to-date HPV vaccination status.
+Current literature provides mixed results on the association between race and ethnicity and HPV vaccination status.
+Several studies found that Hispanic children are less likely to have received one or more HPV vaccination doses than non-Hispanic white children, while other studies indicate that vaccine initiation is lower is non-Hispanic whites than in racial and ethnic minority groups.
+However, these resources support that HPV vaccine follow-through may be lower in racial and ethnic minority groups, compared to lower HPV vaccine initiation rates in non-Hispanic white teenagers (24,26,27,28).
+Overall, larger-scale, in-depth studies are necessary to determine the association between racial and ethnic group differences in HPV vaccination initiation and completion rates in the U.S.
+
 # Conclusion 
+The analysis studying the socioeconomic factors impacting HPV vaccination completion rates among teenagers yielded many valuable insights despite some challenges that were involved.
+
+The exploratory data analysis (EDA) aimed to understand the distribution of potential predictors and their impact on vaccination status.
+The combination of descriptive statistics, visualizations, and statistical modeling provided a comprehensive understanding of the data set.
+
+Subsequent attempts to fit the data to models met with some challenges and limitation.
+The Logistic Regression models encountered multicollinearity issues, which prompted the use of subsequent techniques/tools like PCA, LASSO regression, or even elastic net regression.
+The PCA led to the reduction of only one principle component, which made it impossible for subsequent analysis.
+
+While the LASSO regression model and the elastic net regression model attempt at solving the multicollinearity issue was satisfactory, both proved itself to be a poor model with low predictive power.
+The Random Forest model was initially promising, but ultimately showed signs of overfitting, thus limiting its reliability.
+
+There are likely other factors beyond those captured in the dataset, such as cultural norms, influencing HPV vaccination completion rates among teenagers in the US.
+These factors could not be adequately captured by the trends identified in the analysis.
+
+In addressing collinearity concerns, feature selection played a pivotal role during the variable selection process, ensuring that only non-repetitive variables were chosen.
+This deliberate strategy aimed to mitigate collinearity effects and enhance the robustness of our analysis.
+Despite these efforts, however, several barriers, including model limitations and logistical challenges, persisted.
+Moving forward, refining modeling techniques and embracing collaborative approaches remain essential for advancing our understanding and overcoming these challenges.
 
@@
 
-1.
-https://www.who.int/teams/health-product-policy-and-standards/standards-and-specifications/vaccine-standardization/human-papillomavirus#:\~:text=Human%20papillomavirus%20(HPV)%20is%20a,approximately%208%2Dkb%20in%20length.
+1.	https://www.who.int/teams/health-product-policy-and-standards/standards-and-specifications/vaccine-standardization/human-papillomavirus#:~:text=Human%20papillomavirus%20(HPV)%20is%20a,approximately%208%2Dkb%20in%20length.
 
-2.
-https://www.womenshealth.gov/a-z-topics/human-papillomavirus#:\~:text=Genital%20HPV%20is%20the%20most,79%20million%20Americans%20have%20HPV.&text=It%20is%20so%20common%20that,some%20point%20in%20their%20lifetime.
+2.	https://www.womenshealth.gov/a-z-topics/human-papillomavirus#:~:text=Genital%20HPV%20is%20the%20most,79%20million%20Americans%20have%20HPV.&text=It%20is%20so%20common%20that,some%20point%20in%20their%20lifetime.
 
-3.
-https://www.health.state.mn.us/data/mcrs/data/qfhpv.html#:\~:text=Human%20Papillomavirus%20(HPV)%20is%20a,with%20HPV%20in%20their%20lifetime.&text=Around%2050%20percent%20of%20HPV,HPV%2C%20which%20can%20cause%20cancer.
+3.	https://www.health.state.mn.us/data/mcrs/data/qfhpv.html#:~:text=Human%20Papillomavirus%20(HPV)%20is%20a,with%20HPV%20in%20their%20lifetime.&text=Around%2050%20percent%20of%20HPV,HPV%2C%20which%20can%20cause%20cancer.
 
-4.  HPV and Cancer.
-    In: National Cancer Institute \[Internet\].
-    1 Mar 2019 \[cited 12 Mar 2024\].
-    Available: https://www.cancer.gov/about-cancer/causes-prevention/risk/infectious-agents/hpv-and-cancer
+4.	HPV and Cancer. In: National Cancer Institute [Internet]. 1 Mar 2019 [cited 12 Mar 2024]. Available: https://www.cancer.gov/about-cancer/causes-prevention/risk/infectious-agents/hpv-and-cancer
 
-5.  https://www.cancer.gov/about-cancer/causes-prevention/risk/infectious-agents/hpv-and-cancer
+5.	https://www.cdc.gov/vaccines/vpd/hpv/hcp/vaccines.html#:~:text=Vaccine%20Composition,45%2C%2052%2C%20and%2058.
 
-6.  https://www.cdc.gov/vaccines/vpd/hpv/hcp/vaccines.html#:\~:text=Vaccine%20Composition,45%2C%2052%2C%20and%2058.
+6.	https://www.cdc.gov/vaccines/vpd/hpv/hcp/recommendations.html#:~:text=Two%20doses%20of%20HPV%20vaccine,third%20dose%20of%20HPV%20vaccine.
 
-7.
-https://www.who.int/teams/health-product-policy-and-standards/standards-and-specifications/vaccine-standardization/human-papillomavirus#:\~:text=Human%20papillomavirus%20(HPV)%20is%20a,approximately%208%2Dkb%20in%20length.
+7.	https://www.cdc.gov/mmwr/volumes/72/wr/mm7234a3.htm#:~:text=In%202022%2C%2089.9%25%20of%20adolescents,HPV%20vaccination%20(HPV%20UTD).
 
-8.  https://www.cdc.gov/vaccines/vpd/hpv/hcp/recommendations.html#:\~:text=Two%20doses%20of%20HPV%20vaccine,third%20dose%20of%20HPV%20vaccine.
-
-9.  https://www.cdc.gov/mmwr/volumes/72/wr/mm7234a3.htm#:\~:text=In%202022%2C%2089.9%25%20of%20adolescents,HPV%20vaccination%20(HPV%20UTD).
-
-10. Sciences, N. A.
-    of, Engineering, & Medicine, and.
-    (2018, March 1).
+8.	Sciences, N.
+A.
+of, Engineering, & Medicine.
+(2018).
 Factors that affect health-care utilization.
@@
 
-11. Mansfield, L. N., Chung, R. J., Silva, S. G., Merwin, E. I., & Gonzalez-Guarda, R. M.
+9.	Mansfield, L.
+N., Chung, R.
+J., Silva, S.
+G., Merwin, E.
+I., & Gonzalez-Guarda, R.
+M.
 (2022).
-    Social determinants of human papillomavirus vaccine series completion among U.S. adolescents: A mixed-methods study.
+Social determinants of human papillomavirus vaccine series completion among U.S.
+adolescents: A mixed-methods study.
 SSM - population health, 18, 101082.
@@
 
-12. Casey, R. M., Akaba, H., Hyde, T. B, et al. (2024).
+10.	Casey, R.
+M., Akaba, H., Hyde, T.
+B, et al.
+(2024).
 Covid-19 pandemic and equity of global human papillomavirus vaccination: descriptive study of World Health Organization-Unicef vaccination coverage estimates.
@@
 
-13. https://www.who.int/news/item/20-12-2022-WHO-updates-recommendations-on-HPV-vaccination-schedule
+11.	https://www.who.int/news/item/20-12-2022-WHO-updates-recommendations-on-HPV-vaccination-schedule
 
-14. https://www.paho.org/en/news/5-9-2023-paho-technical-advisory-group-recommends-countries-americas-use-single-dose-hpv
+12.	https://www.paho.org/en/news/5-9-2023-paho-technical-advisory-group-recommends-countries-americas-use-single-dose-hpv
+
+13.	https://www.kff.org/womens-health-policy/fact-sheet/the-hpv-vaccine-access-and-use-in-the-u-s/
+
+14.	 https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6746494/
+
+15.	 https://hpvroundtable.org/wp-content/uploads/2019/09/05_School_Entry_Requirements_for_HPV_Vaccination_WEB_updated.pdf
+
+16.	 https://chlpi.org/wp-content/uploads/2022/07/CHPLI-HPV-Vax-design-v9-web.pdf
+
+17.	https://mississippitoday.org/2023/08/21/hpv-vaccine-mississippi/
+
+18.	https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10478734/
+
+19.	https://pubmed.ncbi.nlm.nih.gov/30735468/
+
+20.	https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7484398/
+
+21.	https://www.cdc.gov/nchs/products/databriefs/db495.htm#:~:text=degree%20or%20higher.-,The%20percentage%20of%20children%20who%20have%20received%20one%20or%20more,of%20the%20federal%20poverty%20level.
+
+22.	https://www.thelancet.com/journals/lanam/article/PIIS2667-193X(24)00021-8/fulltext
+
+23.	https://www.tandfonline.com/doi/full/10.1080/21645515.2024.2313249#:~:text=With%20regard%20to%20race%2Fethnicity,likely%20to%20have%20been%20vaccinated.
+
+24.	https://pubmed.ncbi.nlm.nih.gov/29347831/
+
+25.	https://www.sciencedirect.com/science/article/abs/pii/S0091743519301148

### 2024-05-03 code/analysis-code/analysis2.qmd
Updated figures

@@
 # Combine the training and test metrics data frames
-combined_metrics_df <- rbind(train_metrics_df, test_metrics_df)
+combined_metrics_df <- rbind(train_metrics_df, test_metrics_df) %>%
+  filter(Metric %in% c("MSE", "R^2", "AUC"))
 
@@
     subtitle = md("Comparison of Metrics between Test and Training Datasets")
+  )%>%
+  fmt_number(
+    decimals = 3  
   )

### 2024-05-03 code/analysis-code/analysis2.qmd
Final edits and cleaning

@@
+<<<<<<< Updated upstream
 ---
@@
 ### Conclusion
+=======
+---
+title: "Data Analysis and Modelling of Teen Vaccination Surveys"
+date: '`r format(Sys.Date())`'
+Author: Kelly Cao and Rachel Robertson
+output: html_document
+editor: 
+  markdown: 
+    wrap: sentence
+
+---
+### Introduction
+The previous analysis performed shown on analysis1.qmd found in the same directory, showed collinearity among many of the predictor variables.
+The following script is an attempt to reduce that issue to prevent problem when performing analyses further downstream.
+
+```{r setup, include=FALSE} 
+knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
+```
+
+#### Loading library and data set
+``` {R, warnings=FALSE, messages=FALSE}
+# Load required libraries
+library(MASS) # for robust regression
+library(caret) # for cross-validation
+library(h2o) #For dimension reduction
+library(dplyr) 
+library(rsample)
+library(knitr)
+library(kableExtra)
+library(webshot)
+library(here)
+library(gt)
+library(gtsummary)
+```
+
+I start with loading the same data set and performing the same functions to clean up the data by removing NAs.
+We will be looking at the variable P_UTDHPV, which reflects the up-to-date HPV vaccination status of the individual.
+```{R}
+# Load and preprocess data
+data_location <- here::here("data","processed-data","cleandata1.rds")
+mydata <- readRDS(data_location)
+
+# Remove rows with missing values for the response variable
+mydata <- mydata[complete.cases(mydata$P_UTDHPV), ]
+
+# Set Seed for reproducibility
+rngseed = 1234
+set.seed(rngseed)
+
+# Create training data by alloting 3/4 of the data for training
+data_split <- initial_split(mydata, prop = 3/4)
+
+# Create data frames for the two sets:
+train_data <- training(data_split)
+test_data  <- testing(data_split)
+
+```
+
+### PCA Analysis
+I then initialize the h2o cluster and converted my data object to a H2oFrame or an h2o object.
+I then performed PCA using the function `h20.prcomp` and using the newly converted h20 object as a training frame.
+
+This is perform in attempt t oindirectly reduce the collinearity in the predictor variables.
+
+The following lines are based on the resource shown below:
+
+[Link](https://bradleyboehmke.github.io/HOML/pca.html)
+
+>PCA indirectly addresses multicollinearity by transforming the original features into a new set of orthogonal variables, whereas LASSO/Ridge regression directly addresses multicollinearity by penalizing the coefficients of correlated predictors.
+
+```{R}
+# Initialize and connect to the H2O cluster
+# Start an H2O cluster
+h2o.init()
+
+# Convert train_data to H2OFrame
+mydata_h2o <- as.h2o(train_data)
+
+# Perform PCA using h2o.prcomp
+pca_model <- h2o.prcomp(
+  training_frame = mydata_h2o, 
+  x = c("AGE", "SEX", "STATE", 
+        "INS_STAT2_I", "INCQ298A", 
+        "INS_BREAK_I", "INCPOV1", 
+        "RACEETHK", "EDUC1", 
+        "LANGUAGE", "MOBIL_1", 
+        "RENT_OWN", "FACILITY"))
+
+# Viewing the result of the PCA
+print(pca_model)
+
+```
+
+```{r}
+# Extract the components from the PCA model
+standard_deviation <- pca_model@model$model_summary$pc1[1]
+proportion_variance <- pca_model@model$model_summary$pc1[2]
+cumulative_proportion <- pca_model@model$model_summary$pc1[3]
+
+# Create a data frame with the PC information
+scree_data <- data.frame(
+  Component = "PC1",
+  Standard.deviation = standard_deviation,
+  Variance = standard_deviation^2,
+  Proportion.Variance = proportion_variance,
+  Cumulative.Proportion = cumulative_proportion
+)
+
+# Print the summary table
+print(scree_data)
+
+# Create a table using kableExtra
+pca_table <- 
+  scree_data %>%
+  gt() %>%
+  tab_header(
+    title = "PCA Summary Table"
+  )
+
+# Save the table to a file
+gtsave(pca_table, here("results","tables", "pca_summary.png"))
+```
+
+The PCA results showed one principle component with a high standard deviation and high proportion of variance, showing that the Pc1 captures a large portion (97.9%) of variability in the data.
+
+The result of having only one principle component further supports the fact that many of the predictor variables are highly collinear.
+One PC shows that a PCA must reduce the data object to a single dimension.
+Having a single principal component could be problematic for my analysis, particularly if it leads to the loss of crucial information or structure in my data due to collinearity among the original features.
+
+Understanding why this occurs and assessing whether it aligns with my expectations and the characteristics of my dataset is essential.
+I may need to potentially review my data preprocessing methods or explore different feature selection techniques.
+
+### LASSO Regression
+I first investigate alternative approaches to address collinearity, such as employing regularization methods like LASSO regression.
+This would ensure that I'm effectively capturing the underlying patterns in my data while mitigating the challenges posed by collinearity.
+
+```{R}
+# Use LASSO regression for variable selection and regularization
+lasso_model <- h2o.glm(
+  x = c("AGE", "SEX", "STATE",
+        "INS_STAT2_I", "INCQ298A"
+        , "INS_BREAK_I", "INCPOV1",
+        "RACEETHK", "EDUC1",
+        "LANGUAGE", "MOBIL_1",
+        "RENT_OWN", "FACILITY"),
+  y = "P_UTDHPV", training_frame = mydata_h2o, family = "binomial", alpha = 1)
+
+# View Lasso Model
+print(lasso_model)
+
+```
+
+The model used Lasso to pick out 97 important predictors from 105, which likely dealt with issues of collinearity among the original features.
+This simplification helped in making the model easier to understand and potentially improved its performance.
+
+### Attempt at Cross-validation
+I then utilized ChatGPT to write me a code to perform a cross-validation test.
+
+```{R}
+# Create a fold column for cross-validation
+fold_column <- h2o.kfold_column(data = mydata_h2o, nfolds = 10, seed = 1)  # Specify the number of folds and a seed for reproducibility
+
+# Add the fold column to the H2OFrame
+mydata_h2o <- h2o.cbind(mydata_h2o, fold_column)
+
+# Specify predictor variables
+predictor_columns <- c("AGE", "SEX", "STATE", "INS_STAT2_I", "INCQ298A", "INS_BREAK_I", 
+                       "INCPOV1", "RACEETHK", "EDUC1", "LANGUAGE", "MOBIL_1", "RENT_OWN", "FACILITY")
+
+# Evaluate model performance using cross-validation
+cv_metrics <- h2o.glm(x = predictor_columns, 
+                      y = "P_UTDHPV", 
+                      training_frame = mydata_h2o, 
+                      fold_column = "C1",  # Use the name of the fold column (default name is "C1")
+                      family = "binomial")
+
+# Display results
+print(cv_metrics)
+```
+
+The cross-validation results for the logistic regression model show consistent performance across folds, with an average accuracy of approximately 79.4%.
+The model demonstrates stability with metrics such as an area under the curve (AUC) around 0.644 and a mean per-class error of about 49.8%.
+Precision and recall metrics also exhibit robustness, averaging around 79.6% and 99.6%, respectively.
+These results indicate the model's ability to balance correctly identifying positive cases while minimizing false positives.
+However, areas for model refinement include residual deviance and specificity.
+
+```{r}
+# Replicating the previous analysis with test data
+
+# Convert train_data to H2OFrame
+mydata_h2o <- as.h2o(test_data)
+
+# Create a fold column for cross-validation
+fold_column <- h2o.kfold_column(data = mydata_h2o, nfolds = 10, seed = 1)  # Specify the number of folds and a seed for reproducibility
+
+# Add the fold column to the H2OFrame
+mydata_h2o <- h2o.cbind(mydata_h2o, fold_column)
+
+# Specify predictor variables
+predictor_columns <- c("AGE", "SEX", "STATE", "INS_STAT2_I", "INCQ298A", "INS_BREAK_I", 
+                       "INCPOV1", "RACEETHK", "EDUC1", "LANGUAGE", "MOBIL_1", "RENT_OWN", "FACILITY")
+
+# Evaluate model performance using cross-validation
+cv_metrics <- h2o.glm(x = predictor_columns, 
+                      y = "P_UTDHPV", 
+                      training_frame = mydata_h2o, 
+                      fold_column = "C1",  # Use the name of the fold column (default name is "C1")
+                      family = "binomial")
+
+# Display results
+print(cv_metrics)
+
+# Function to extract metrics from a model
+extract_metrics <- function(model, dataset) {
+  metrics <- model@model$training_metrics@metrics
+  mse <- metrics$MSE
+  r2 <- metrics$r2
+  logloss <- metrics$logloss
+  mpce <- metrics$mean_per_class_error
+  auc <- metrics$AUC
+  gini <- metrics$Gini
+  nulldev <- metrics$null_deviance
+  resdev <- metrics$residual_deviance
+  aic <- metrics$AIC
+
+  # Create a data frame for metrics
+  metrics_df <- data.frame(
+    Metric = c("MSE", "R^2", "Log Loss", "Mean Per-Class Error", "AUC", "Gini", "Null Deviance", "Residual Deviance", "AIC"),
+    Value = c(mse, r2, logloss, mpce, auc, gini, nulldev, resdev, aic),
+    Dataset = dataset
+  )
+
+  return(metrics_df)
+}
+
+# Extract metrics from the LASSO model (training data)
+train_metrics_df <- extract_metrics(lasso_model, "Training")
+
+# Extract metrics from the CV metrics (test data)
+test_metrics_df <- extract_metrics(cv_metrics, "Test")
+
+# Combine the training and test metrics data frames
+combined_metrics_df <- rbind(train_metrics_df, test_metrics_df) %>%
+  filter(Metric %in% c("MSE", "R^2", "AUC"))
+
+# Create a gt table
+gt_table <- combined_metrics_df %>%
+  gt(
+    groupname_col = "Dataset", 
+    row_group_as_column = TRUE
+  ) %>%
+  tab_header(
+    title = md("Model Evaluation Metrics"),
+    subtitle = md("Comparison of Metrics between Test and Training Datasets")
+  )%>%
+  fmt_number(
+    decimals = 3  
+  )
+
+# Save the gt table
+gtsave(gt_table, here("results", "tables", "LASSO_metrics.png"))
+
+```
+
+### Conclusion
+>>>>>>> Stashed changes
 our analysis aimed to address collinearity issues in our teen vaccination surveys dataset through dimension reduction techniques like PCA and regularization methods such as LASSO regression.

### 2024-05-03 code/analysis-code/analysis4.qmd
Final edits and cleaning

@@
 
+We will examine the coefficients of the model to determine the associations of the variables with the outcome.
+```{r}
+# Extract and tidy model parameters
+en_tune_coef <- en_tune_fit %>%
+  extract_fit_parsnip() %>%
+  tidy()
+
+# Filter out predictors with non-zero coefficients
+en_nonzero_coef <- en_tune_coef %>%
+  filter(estimate != 0)
+en_nonzero_coef
+```
+
 Despite this, I will see if the important predictors are in agreement with our LASSO model using the vip package.
+
+The elastic net model produced the same important variables as the previous lasso model. 
 ```{r}
@@
 
-print(en_vars)
+# Modify the plot to set the color of the bars
+en_vars_plot <- en_vars+
+  aes(fill = "skyblue")+
+  labs(title = "Top Elastic Net Predictors")
+
+# Create a data frame with original and translated variable names
+variable_translations <- data.frame(
+  Original_Variable = rev(c("STATE_RHODE.ISLAND", "FACILITY", 
+                        "STATE_VERMONT", "STATE_MISSISSIPPI", "STATE_NORTH.DAKOTA", "STATE_IOWA", 
+                        "STATE_HAWAII", "RENT_OWN_DON.T.KNOW", "STATE_NEBRASKA", 
+                        "INS_BREAK_I_CURRENTLY.UNINSURED.AND.NEVER.INSURED.SINCE.AGE.11")),
+  Translated_Variable = rev(c("Rhode Island", "All Std School Teen Clinics or Other Facilities", 
+                          "Vermont", "Mississippi", "North Dakota", "Iowa", "Hawaii", 
+                          "Rent/Own Status (Don't Know)", "Nebraska", 
+                          "Uninsured"))
+)
+
+# Add translations to the plot
+en_plot_with_translations <- en_vars_plot +
+  scale_fill_manual(values = c("skyblue")) +  # Set fill color manually
+  scale_x_discrete(labels = variable_translations$Translated_Variable) +  # Translate x-axis labels
+  labs(title = "Top Predictors",
+                x = "Variable")+  # Add title and update x-axis label
+  scale_y_continuous(labels = function(x) strwrap(x, width = 10))
+
+# Display the plot with translated variable names
+print(en_plot_with_translations)
+
+# Save the plot as a PNG file
+path = here("results", "figures", "top_en_predictors.png")
+ggsave(filename = path, plot=en_plot_with_translations) 
 ```
-The elastic net model produced the same important variables as the previous lasso model.

### 2024-05-03 code/eda-code/EDA-code.qmd
Final edits and cleaning

@@
 library(knitr)
+library(gtsummary)
 ```
@@
 # save to file
+
 summarytable_file = here("results","tables", "summarytable.rds")
 saveRDS(summary_df, file = summarytable_file)
+```
+I will create a table with gt summary to display in the manuscript.
+```{r}
+table1 <- mydata %>%
+  select(AGE, SEX, RACEETHK, INS_STAT2_I, INCPOV1, P_UTDHPV) %>% # select the variables to include in table
+  mutate(SEX = factor(SEX, # change the factor labels using dpylr
+                      levels = c("MALE", "FEMALE"),
+                      labels = c("Male", "Female")),
+         RACEETHK = factor(RACEETHK, 
+                           levels = c("NON-HISPANIC WHITE ONLY", 
+                                      "NON-HISPANIC BLACK ONLY", 
+                                      "NON-HISPANIC ASIAN ONLY",
+                                      "HISPANIC",
+                                      "NON-HISPANIC OTHER + MULTIPLE RACE"),
+                           labels = c("White", "Black", "Asian", "Hispanic", "Multiracial")),
+         INS_STAT2_I = factor(INS_STAT2_I,
+                              levels = c("PRIVATE INSURANCE ONLY", 
+                                         "ANY MEDICAID", 
+                                         "UNINSURED",
+                                         "OTHER INSURANCE (CHIP, IHS, MILITARY, OR OTHER, ALONE OR IN COMB. WITH PRIVATE INSURANCE)"),
+                              labels = c("Private", "Medicaid", "Uninsured", "Other insurance")),
+         INCPOV1 = factor(INCPOV1,
+                          levels = c("BELOW POVERTY", 
+                                     "ABOVE POVERTY > $75K", 
+                                     "ABOVE POVERTY <= $75K", 
+                                     "UNKNOWN"),
+                          labels = c("Below Poverty", "Above Poverty > $75K", "Above poverty <= $75K", "Unknown")),
+         P_UTDHPV = factor(P_UTDHPV,
+                           levels = c("UTD", "NOT UTD"),
+                           labels = c("Up-to-date", "Not Up-to-date")))%>%
+  tbl_summary(by = P_UTDHPV, # group by HPV vaccination status
+      statistic = list( 
+      all_continuous() ~ "{mean} ({sd})", # select the mean and standard deviation of continuous variables
+      all_categorical() ~ "{n} ({p}%)"),# select the number and percentage of categorical variables
+    label = list(
+      AGE ~ "Age (years)", # change the names of the variables to be more readable
+      SEX ~ "Sex",   
+      RACEETHK ~ "Ethnicity", 
+      INS_STAT2_I ~ "Insurance Status",      
+      INCPOV1 ~ "Poverty Status"), missing = "no") %>%
+    bold_labels() %>%
+  modify_caption("Table 1. Study sample of Teenagers aged 13-17 who received the NIS Teen survey in the U.S.")
+
+table1
+
+# Save the summary table
+saveRDS(table1, file = here("results/tables/summary_table.rds"))
 ```

### 2024-05-03 products/manuscript/Manuscript.qmd
Updated figures
Final edits and cleaning

@@
 ---
-title: "Robertson and Cao Project"
-subtitle: "Part 1:Teenage HPV Vaccination Coverage and Socioeconomic Factors"
+title: "Kelly Cao and Rachel Robertson Project"
+subtitle: "Socioeconomic and geographic determinants of HPV vaccination completion for US teenagers in 2022"
 author: Kelly Cao and Rachel Robertson
@@
   markdown: 
-    wrap: sentence
+    wrap: 72
 ---
@@
 # Abstract
-Human Papillomavirus (HPV) is thought to be the most common sexually transmitted disease in the United States [2].
-Though HPV consists of a group of over 200 related viruses, only 12 types of HPV are considered “high-risk” and associated with the incidence of six types of cancer, including oropharyngeal, cervical, penile, vulvar, vaginal, and anal cancers [4,5].
-The U.S.
-currently recommends HPV vaccination with Gardasil 9 to immunize against six common types of high-risk HPV infections and three HPV types associated with genital warts.
-It is recommended that a primary dose of the HPV vaccine be given between the ages of 11 to 12 and a secondary dose administered after 6 months to 1 year from the primary [6].
-Although Healthy People 2023 has set the goal of 80% HPV vaccination coverage in the U.S., only 62.6% of U.S.
-teenagers are up-to-date on their HPV vaccination [7].
+
+Human Papillomavirus (HPV) is thought to be the most common sexually transmitted disease in the United States \[2\].
+Though HPV consists of a group of over 200 related viruses, only 12 types of HPV are considered “high-risk” and associated with the incidence of six types of cancer, including oropharyngeal, cervical, penile, vulvar, vaginal, and anal cancers \[4,5\].
+The U.S. currently recommends HPV vaccination with Gardasil 9 to immunize against six common types of high-risk HPV infections and three HPV types associated with genital warts.
+It is recommended that a primary dose of the HPV vaccine be given between the ages of 11 to 12 and a secondary dose administered after 6 months to 1 year from the primary \[6\].
+Although Healthy People 2023 has set the goal of 80% HPV vaccination coverage in the U.S., only 62.6% of U.S. teenagers are up-to-date on their HPV vaccination \[7\].
 We are interested in examining which socioeconomic and geographic factors impact HPV vaccination completion among teenagers in the U.S.
@@
 
-Human Papillomavirus (HPV) is a double-stranded DNA virus that infects the cells of the skin and mucosal linings [1].
-HPV is the most common sexually transmitted infection (STI) in the United States, where over 90% of sexually active males and 80% of sexually active females will likely become infected within their lifetime [2, 3].
+Human Papillomavirus (HPV) is a double-stranded DNA virus that infects the cells of the skin and mucosal linings \[1\].
+HPV is the most common sexually transmitted infection (STI) in the United States, where over 90% of sexually active males and 80% of sexually active females will likely become infected within their lifetime \[2, 3\].
 Most cases of HPV are asymptomatic and cleared over time, but certain infections tend to become persistent and can cause genital warts.
-Specific types of HPV are known to be “high-risk” or increase the likelihood of developing cancer in persistently infected cells [4].
+Specific types of HPV are known to be “high-risk” or increase the likelihood of developing cancer in persistently infected cells \[4\].
 The HPV group contains over 200 related viruses, but only 12 types are considered “high-risk” due to the presence of oncogenes.
 These include HPV 16, 18, 31, 33, 35, 39, 45, 51, 52, 56, 58, and 59, though HPV-16 and HPV-18 are associated with the highest number of HPV-associated cancer cases.
-In the U.S., there are 37,300 cases of HPV-associated cancer cases every year, including cases of cervical, anal, oropharyngeal (throat and mouth), vulvar, vaginal, and penile cancers [5,6].
+In the U.S., there are 37,300 cases of HPV-associated cancer cases every year, including cases of cervical, anal, oropharyngeal (throat and mouth), vulvar, vaginal, and penile cancers \[4,5\].
 In an effort to prevent HPV-associated cancers worldwide, two HPV vaccines were developed and made available in 2006.
-These included a bivalent and quadrivalent vaccine that contained viral-like particles (VLPs) of the viral capsid protein (L1) from “high-risk” HPV types [7].
-In recent years, the vaccine used in the U.S., Gardasil 9, has been expanded to contain HPV types 6, 11, 16, 18, 31, 33, 45, 52, and 58 [4].
-Currently, the United States Centers for Disease Control and Prevention (CDC) recommends a two-dose Gardasil 9 regime to all teenagers, with a primary dose at age 11 to 12 and a second dose within 6 to 12 months of the first [8].
-In 2022, the CDC estimated that 76.0% of teenagers aged 13 to 17 had received at least 1 dose of the HPV vaccine while only 62.6% of teenagers were up-to-date with their vaccination regime [9].
+These included a bivalent and quadrivalent vaccine that contained viral-like particles (VLPs) of the viral capsid protein (L1) from “high-risk” HPV types \[1\].
+In recent years, the vaccine used in the U.S., Gardasil 9, has been expanded to contain HPV types 6, 11, 16, 18, 31, 33, 45, 52, and 58 \[4\].
+Currently, the United States Centers for Disease Control and Prevention (CDC) recommends a two-dose Gardasil 9 regime to all teenagers, with a primary dose at age 11 to 12 and a second dose within 6 to 12 months of the first \[6\].
+In 2022, the CDC estimated that 76.0% of teenagers aged 13 to 17 had received at least 1 dose of the HPV vaccine while only 62.6% of teenagers were up-to-date with their vaccination regime \[7\].
 
@@
 It is generally known that those of lower income and without insurance have limited healthcare access in the U.S.
-[10], but it will be interesting to see if this pattern is reflected at the state level.
+\[8\], but it will be interesting to see if this pattern is reflected at the state level.
 If so, policy should be examined as it may be related to healthcare access disparities.
-Recent studies have examined HPV vaccination completion according to social determinants, but they have not examined geographic distribution and include data prior to 2019 [11].
-More recent years display differences in global HPV vaccination trends due to the COVID19 pandemic [12].
+Recent studies have examined HPV vaccination completion according to social determinants, but they have not examined geographic distribution and include data prior to 2019 \[9\].
+More recent years display differences in global HPV vaccination trends due to the COVID19 pandemic \[10\].
 Our study will fill in geographical gaps and provide updated vaccination trends based on socioeconomic factors.
@@
 
-```{r}
-
+```{r, warning= FALSE, message = FALSE, echo = FALSE}
 #| label: Table1
@@
 
-Table1 <- readRDS(here("results/tables/table1.rds"))
+Table1 <- readRDS(here("results/tables/summary_table.rds"))
 
 kable(Table1) 
-
 ```
@@
 
-![Figure 1. Income-poverty ratio distribution](../../results/figures/income.png)
+![Figure 1. Income-poverty ratio distribution for study sample of U.S. teens.](../../results/figures/income.png){width="80%"}
 
-![Figure 2. The family income distribution](../../results/figures/family.income.distribution.png)
+![Figure 2. The family income distribution for the study sample of U.S. teens.](../../results/figures/family.income.distribution.png){width="80%"}
 
@@
 Although it is noted a few territories provide 0 observations.
-The race and ethnicity variables show a distribution that is similar to the U.S.
-census data.
+The race and ethnicity variables show a distribution that is similar to the U.S. census data.
 The census data states that the racial composition of the U.S.
@@
 
-![Figure 3.Facility and vaccine completion distribution](../../results/figures/facility-vaccination.png)
+![Figure 3.Facility at which survey was conducted versus vaccine completion distribution of U.S. teens.](../../results/figures/facility-vaccination.png){width="80%"}
 
 A total of 79% of participants had up-to-date status for receiving 1 or more shots of HPV vaccination.
-U.S.
-states and territories had an HPV vaccination completion rate ranging from 69% to 95%, fluctuating around ~80%.
+U.S. states and territories had an HPV vaccination completion rate ranging from 69% to 95%, fluctuating around \~80%.
 There are different counts of each race/ethnicity for each factor level, so the percentage of those with up-to-date vaccine status out of each race/ethnicity level was adjusted.
@@
 
-![Figure 4. Distribution of income poverty ratio stratified by race and HPV vaccination completion status](../../results/figures/raceeth-income-stratified.png)
+![Figure 4. Distribution of income poverty ratio stratified by race and HPV vaccination completion status.](../../results/figures/raceeth-income-stratified.png){width="80%"}
 
@@
 
- ![Figure 5. Income poverty ratio and insurance status](../../results/figures/insurance-income-stratified.png)
+![Figure 5. Income poverty ratio and insurance status stratified by HPV vaccination status.](../../results/figures/insurance-income-stratified.png){width="80%"}
 
@@
 
-The variables selected for analysis were chosen based on their interpretability and applicability to our research question “Which socioeconomic and geographic factors impact HPV vaccine completion rates among teenagers?” Any variables that capture repetitive data were removed and the most important or straightforward version of the variable was kept in the model.
+The variables selected for analysis were chosen based on their interpretability and applicability to our research question “Which socioeconomic and geographic factors impact HPV vaccine completion rates among teenagers?”
+Any variables that capture repetitive data were removed and the most important or straightforward version of the variable was kept in the model.
 Seven variables represented data for the outcome of HPV vaccination completion:
 
-![Table 2. Human Papillomavirus vaccine status variables and meanings.](./images/UTDHPV-variable-table.png)
+![Table 2. Human Papillomavirus vaccine status variables and meanings.](../../assets/tables/UTDHPV-variable-table.png){width="80%"}
 
@@
 Additionally, in April of 2022, the SAGE work group of the WHO officially recommended a single-dose HPV vaccine regime, because vaccine efficacy was found to be nearly as high with a single dose as it is with the multi-dose regime.
-This recommendation was given with the goal of increasing vaccine coverage in areas with lower vaccine accessibility and availability (13).
-In September of 2023, PAHO followed suit in recommending a single-dose HPV vaccine regime for the Americas (14).
+This recommendation was given with the goal of increasing vaccine coverage in areas with lower vaccine accessibility and availability \[11\].
+In September of 2023, PAHO followed suit in recommending a single-dose HPV vaccine regime for the Americas \[12\].
 This supports our rationale for choosing the single or more dose, up-to-date, outcome variable.
@@
 
-![Table 3. Potential predictor variable names and meanings.](./images/Predictor-variable-table.png)
+![Table 3. Potential predictor variable names and meanings.](../../assets/tables/Predictor-variable-table.png){width="80%"}
 
@@
 
-![Figure 6. Residual Plot Results](../../results/figures/residual_plot.png)
+![Figure 6. Residual Plot Results](../../results/figures/residual_plot.png){width="80%"}
 
@@
 
-![Table 4. Summary of PCA Attempt](../../results/tables/pca_summary.png)
+![Table 4. Summary of PCA Attempt](../../results/tables/pca_summary.png){width="80%"}
 
@@
 
-![Table 5. Summary of Initial LASSO Regression](../../results/tables/LASSO_metrics.png){height=500}
+![Table 5. Summary of Initial LASSO Regression](../../results/tables/LASSO_metrics.png){width="80%"}
 
@@
 
-We used a Random Forest Model to assess which predictors are associated with HPV vaccination completion rates in U.S.
-teens.
+We used a Random Forest Model to assess which predictors are associated with HPV vaccination completion rates in U.S. teens.
 Decision tree models are known to be able to handle collinearity and predictors with different trends.
@@
 
-![Figure 7. Top Predictors Chosen by Random Forest Model](../../results/figures/variable_importance_plot.png)
+![Figure 7. Top Predictors Chosen by Random Forest Model](../../results/figures/variable_importance_plot.png){width="80%"}
 
@@
 
-![Figure 8. Top Predictors chosen by LASSO Regression Model](../../results/figures/top_lasso_predictors.png)
+![Figure 8. Top Predictors chosen by LASSO Regression Model](../../results/figures/top_lasso_predictors.png){width="80%"}
 
@@
 The most important predictors in determining HPV vaccination completion, from most to least important, were state (Rhode Island), facility (std, school, and teen clinics), state (Vermont), state (Mississippi), state (North Dakota), state (Iowa), state (Hawaii), housing status (unknown), insurance breaks (currently uninsured or never insured since 11 years of age), and state (Nebraska).
+![Figure 8. Most Important Predictors chosen by Elastic Net Model](../../results/figures/top_en_predictors.png){width="80%"}
 
@@
 The findings will be interpreted by examining which variables were chosen by all three models, as they may hold the most importance, followed by the predictors chosen by the random forest model.
-
 For all three model types that were tested, state was chosen to be the most important factor in determining HPV vaccine completion status.
@@
 This may be explained by the fact that each state determines HPV vaccination legislation, policy, and funding, for teenagers aged 11 to 13.
-As of 2021, four states have a law requiring vaccination for school entry, including,  Hawaii, Rhode Island, Virginia, and Washington D.C (15).
-As of 2019, four states recommend HPV vaccination for school attendance, by law, including, Arkansas, Delaware, West Virginia, and Wyoming (16).
+As of 2021, four states have a law requiring vaccination for school entry, including, Hawaii, Rhode Island, Virginia, and Washington D.C \[13\].
+As of 2019, four states recommend HPV vaccination for school attendance, by law, including, Arkansas, Delaware, West Virginia, and Wyoming \[14\].
 As of 2019, 42 U.S.
-states and jurisdictions provide legislative funding to promote HPV vaccination, however, Rhode Island is the only state to implement both a school entry requirement and funding for HPV vaccination (17).
-Additionally, Rhode Island has the highest HPV vaccination coverage out of all U.S.
-states and jurisdictions, exceeding 80% coverage as of 2022 (18).
-The states selected as most important by the models that have legislation to fund HPV vaccination campaigns include Rhode Island, North Dakota, Mississippi, and Iowa (19).
+states and jurisdictions provide legislative funding to promote HPV vaccination, however, Rhode Island is the only state to implement both a school entry requirement and funding for HPV vaccination \[15\].
+Additionally, Rhode Island has the highest HPV vaccination coverage out of all U.S. states and jurisdictions, exceeding 80% coverage as of 2022 \[16\].
+The states selected as most important by the models that have legislation to fund HPV vaccination campaigns include Rhode Island, North Dakota, Mississippi, and Iowa \[14\].
 States chosen as important by the model with legislation mandating HPV vaccination are Rhode Island and Hawaii.
-Neither Vermont nor Nebraska have legislature funding and mandating HPV vaccination, although, both are in the top 15 states for HPV vaccination coverage in the U.S.
-(18,19).
+Neither Vermont nor Nebraska have legislature funding and mandating HPV vaccination, although, both are in the top 15 states for HPV vaccination coverage in the U.S. \[17,18\].
 Conversely, Mississippi has a legislature to reimburse the cost of HPV vaccination to insurance providers but has the lowest HPV vaccination coverage in the country.
-This is thought to be due to misconceptions and misinformation about the HPV vaccine’s purpose and the way that HPV is transmitted (18,20).
+This is thought to be due to misconceptions and misinformation about the HPV vaccine’s purpose and the way that HPV is transmitted \[16,18\].
 The mixture of policies and vaccination coverage rate, which is not always policy dependent, demonstrates the importance of a combination of legislation to mandate and fund HPV vaccination as well as provider and patient education to support HPV vaccination.
-
 All three models also indicated that facility and housing status are important in determining up-to-date HPV vaccination status.
@@
 Although school-based HPV vaccination programs have shown promise in increasing HPV vaccine uptakes internationally, these programs in the U.S.
-face barriers such as politicization of vaccines, patient consent processes for teens, and funding for these programs (21,22).
+face barriers such as politicization of vaccines, patient consent processes for teens, and funding for these programs \[18,19\].
 An unknown housing status was also negatively associated with the outcome of up-to-date HPV vaccination in the LASSO and elastic net models.
@@
 A cross-sectional study and analysis of house-based socioeconomic status index in the U.S.
-in 2020, indicated that HPV vaccination initiation was related to housing and socioeconomic status, where lower quartiles corresponded to lower rates of HPV vaccine initiation (23).
-
+in 2020, indicated that HPV vaccination initiation was related to housing and socioeconomic status, where lower quartiles corresponded to lower rates of HPV vaccine initiation \[20\].
 Along with the variables state and facility, the random forest model indicated family income, age, and race/ ethnicity to be in the top five most important predictors for HPV vaccination completion.
@@
 In contrast to our findings, a CDC data brief found that the percentage of teenagers who received one or more HPV vaccine doses increased as family income moved to be greater than the federal poverty level.
-However, this study does not differentiate vaccination levels at the uppermost income brackets (24).
+However, this study does not differentiate vaccination levels at the uppermost income brackets \[21\].
 A recent study analyzing National Immunization Teen Surveys from 2017 to 2021 found that 64.7% of parents in advantaged income brackets had no intention to begin HPV vaccination while 40.9% of parents in disadvantaged income brackets had no intention to begin the vaccine regime.
 Additionally, the advantaged group was less likely to complete HPV vaccination than the disadvantaged group.
-For the advantaged group, the main reason for not starting the regime was safety concerns and the main reason for not completing the regime was the belief that their child was up-to-date on HPV vaccinations (25).
+For the advantaged group, the main reason for not starting the regime was safety concerns and the main reason for not completing the regime was the belief that their child was up-to-date on HPV vaccinations \[22\].
 This highlights the importance of addressing vaccination misinformation and provider education to parents on the recommended HPV vaccination schedule.
-
 Age was selected as the fourth most important predictor of HPV vaccination completion by the random forest model, but not selected by the LASSO or elastic net models, so the direction of association for this study is unknown.
 A 2024 data brief by the CDC found that the percentage of teenagers who have received at least one dose of the HPV vaccination increases with age, despite the recommended age for beginning the HPV vaccination being 11 to 12 years old in the U.S.
-(24).
+\[21\].
 Next, Race and ethnicity were selected as the fifth most important predictor of HPV vaccination completion by the random forest model, and selected by the LASSO and elastic net models, though it was not found to be the top most important predictor by this model.
@@
 Several studies found that Hispanic children are less likely to have received one or more HPV vaccination doses than non-Hispanic white children, while other studies indicate that vaccine initiation is lower is non-Hispanic whites than in racial and ethnic minority groups.
-However, these resources support that HPV vaccine follow-through may be lower in racial and ethnic minority groups, compared to lower HPV vaccine initiation rates in non-Hispanic white teenagers (24,26,27,28).
+However, these resources support that HPV vaccine follow-through may be lower in racial and ethnic minority groups, compared to lower HPV vaccine initiation rates in non-Hispanic white teenagers \[21,23,24,25\].
 Overall, larger-scale, in-depth studies are necessary to determine the association between racial and ethnic group differences in HPV vaccination initiation and completion rates in the U.S.
@@
 # Conclusion
+
 The analysis studying the socioeconomic factors impacting HPV vaccination completion rates among teenagers yielded many valuable insights despite some challenges that were involved.
@@
 
-1.	https://www.who.int/teams/health-product-policy-and-standards/standards-and-specifications/vaccine-standardization/human-papillomavirus#:~:text=Human%20papillomavirus%20(HPV)%20is%20a,approximately%208%2Dkb%20in%20length.
+1.
+WHO updates recommendations on HPV vaccination schedule.
+(2022, December 20).
+World Health Organization.
+https://www.who.int/news/item/20-12-2022-WHO-updates-recommendations-on-HPV-vaccination-schedule
 
-2.	https://www.womenshealth.gov/a-z-topics/human-papillomavirus#:~:text=Genital%20HPV%20is%20the%20most,79%20million%20Americans%20have%20HPV.&text=It%20is%20so%20common%20that,some%20point%20in%20their%20lifetime.
+2.  Human papillomavirus. (2022, December 29). Office on Women’s Health. https://www.womenshealth.gov/a-z-topics/human-papillomavirus
 
-3.	https://www.health.state.mn.us/data/mcrs/data/qfhpv.html#:~:text=Human%20Papillomavirus%20(HPV)%20is%20a,with%20HPV%20in%20their%20lifetime.&text=Around%2050%20percent%20of%20HPV,HPV%2C%20which%20can%20cause%20cancer.
+3.
+Quick facts: HPV-associated cancer.
+Quick Facts: HPV-Associated Cancer - MN Dept.
+of Health.
+(n.d.).
+https://www.health.state.mn.us/data/mcrs/data/qfhpv.html#:\~:text=Human%20Papillomavirus%20(HPV)%20is%20a,with%20HPV%20in%20their%20lifetime.&text=Around%2050%20percent%20of%20HPV,HPV%2C%20which%20can%20cause%20cancer.
 
-4.	HPV and Cancer. In: National Cancer Institute [Internet]. 1 Mar 2019 [cited 12 Mar 2024]. Available: https://www.cancer.gov/about-cancer/causes-prevention/risk/infectious-agents/hpv-and-cancer
+4.  HPV and Cancer—NCI (nciglobal,ncienterprise). (2019, March 1). [cgvArticle]. National Cancer Institute. https://www.cancer.gov/about-cancer/causes-prevention/risk/infectious-agents/hpv-and-cancer
 
-5.	https://www.cdc.gov/vaccines/vpd/hpv/hcp/vaccines.html#:~:text=Vaccine%20Composition,45%2C%2052%2C%20and%2058.
+5.  HPV Vaccine | CDC. (2023, May 22). Center for Disease Control and Prevention. https://www.cdc.gov/vaccines/vpd/hpv/hcp/vaccines.html
 
-6.	https://www.cdc.gov/vaccines/vpd/hpv/hcp/recommendations.html#:~:text=Two%20doses%20of%20HPV%20vaccine,third%20dose%20of%20HPV%20vaccine.
+6.  HPV Vaccination Recommendations | CDC. (2023, May 22). Center for Disease Control and Prevention. https://www.cdc.gov/vaccines/vpd/hpv/hcp/recommendations.html
 
-7.	https://www.cdc.gov/mmwr/volumes/72/wr/mm7234a3.htm#:~:text=In%202022%2C%2089.9%25%20of%20adolescents,HPV%20vaccination%20(HPV%20UTD).
+7.
+Pingali, C., Yankey, D., Elam-Evans, L.
+D., Markowitz, L.
+E., Valier, M.
+R., Fredua, B., Crowe, S.
+J., DeSisto, C.
+L., Stokley, S., & Singleton, J.
+A.
+(2023).
+Vaccination Coverage Among Adolescents Aged 13–17 Years—National Immunization Survey–Teen, United States, 2022.
+MMWR.
+Morbidity and Mortality Weekly Report, 72.
+https://doi.org/10.15585/mmwr.mm7234a3
 
-8.	Sciences, N.
-A.
-of, Engineering, & Medicine.
+8.
+National Academies of Sciences, E., Division, H.
+and M., Services, B.
+on H.
+C., & Disabilities, C.
+on H.
+C.
+U.
+and A.
+with.
 (2018).
-Factors that affect health-care utilization.
-Health-Care Utilization as a Proxy in Disability Determination.
+Factors That Affect Health-Care Utilization.
+In Health-Care Utilization as a Proxy in Disability Determination.
+National Academies Press (US).
 https://www.ncbi.nlm.nih.gov/books/NBK500097/
 
-9.	Mansfield, L.
-N., Chung, R.
-J., Silva, S.
-G., Merwin, E.
-I., & Gonzalez-Guarda, R.
-M.
+9.  Mansfield, L. N., Chung, R. J., Silva, S. G., Merwin, E. I., & Gonzalez-Guarda, R. M.
     (2022).
-Social determinants of human papillomavirus vaccine series completion among U.S.
-adolescents: A mixed-methods study.
+    Social determinants of human papillomavirus vaccine series completion among U.S. adolescents: A mixed-methods study.
     SSM - population health, 18, 101082.
@@
 
-10.	Casey, R.
-M., Akaba, H., Hyde, T.
-B, et al.
-(2024).
+10. Casey, R. M., Akaba, H., Hyde, T. B, et al. (2024).
     Covid-19 pandemic and equity of global human papillomavirus vaccination: descriptive study of World Health Organization-Unicef vaccination coverage estimates.
@@
 
-11.	https://www.who.int/news/item/20-12-2022-WHO-updates-recommendations-on-HPV-vaccination-schedule
+11.
+WHO updates recommendations on HPV vaccination schedule.
+(2022, December 20).
+World Health Organization.
+https://www.who.int/news/item/20-12-2022-WHO-updates-recommendations-on-HPV-vaccination-schedule
 
-12.	https://www.paho.org/en/news/5-9-2023-paho-technical-advisory-group-recommends-countries-americas-use-single-dose-hpv
+12.
+PAHO Technical Advisory Group recommends countries of the Americas to use single-dose HPV vaccine schedule—PAHO/WHO | Pan American Health Organization.
+(2023, September 5).
+PAHO News.
+https://www.paho.org/en/news/5-9-2023-paho-technical-advisory-group-recommends-countries-americas-use-single-dose-hpv
 
-13.	https://www.kff.org/womens-health-policy/fact-sheet/the-hpv-vaccine-access-and-use-in-the-u-s/
+13. Published: (2021, July 12). The HPV Vaccine: Access and Use in the U.S. KFF. https://www.kff.org/womens-health-policy/fact-sheet/the-hpv-vaccine-access-and-use-in-the-u-s/
 
-14.	 https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6746494/
+14.
+Hoss, A., Meyerson, B.
+E., & Zimet, G.
+D.
+(2019).
+State statutes and regulations related to human papillomavirus vaccination.
+Human Vaccines & Immunotherapeutics, 15(7–8), 1519–1526.
+https://doi.org/10.1080/21645515.2019.1627817
 
-15.	 https://hpvroundtable.org/wp-content/uploads/2019/09/05_School_Entry_Requirements_for_HPV_Vaccination_WEB_updated.pdf
+15. HPV Roundtable. (n.d.-a). School entry requirements for HPV vaccine. http://hpvroundtable.org/wp-content/uploads/2019/09/05_School_Entry_Requirements_for_HPV_Vaccination_WEB_updated.pdf
 
-16.	 https://chlpi.org/wp-content/uploads/2022/07/CHPLI-HPV-Vax-design-v9-web.pdf
+16. Harvard Law School. (n.d.). HPV vaccine policy landscape. Center for Health Law and Policy Innovation. https://chlpi.org/wp-content/uploads/2022/07/CHPLI-HPV-Vax-design-v9-web.pdf
 
-17.	https://mississippitoday.org/2023/08/21/hpv-vaccine-mississippi/
+17.
+Paffenroth, S.
+(2023, August 21).
+Mississippi has the worst HPV vaccine uptake in the nation, and women are dying of preventable cancer.
+Mississippi Today.
+http://mississippitoday.org/2023/08/21/hpv-vaccine-mississippi/
 
-18.	https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10478734/
+18.
+Shin, M.
+B., Sloan, K., Baezconde-Garbanati, L., Dang, E., Garcia, S., Palinkas, L.
+A., Unger, J.
+B., Willgerodt, M., Crabtree, B.
+F., & Tsui, J.
+(2023).
+Multilevel perspectives on school-based opportunities to improve HPV vaccination among medically underserved adolescents: Beyond school entry mandates.
+Human Vaccines & Immunotherapeutics, 19(2), 2251815.
+https://doi.org/10.1080/21645515.2023.2251815
 
-19.	https://pubmed.ncbi.nlm.nih.gov/30735468/
+19.
+Oliver, K., McCorkell, C., Pister, I., Majid, N., Benkel, D.
+H., & Zucker, J.
+R.
+(2019).
+Improving HPV vaccine delivery at school-based health centers.
+Human Vaccines & Immunotherapeutics, 15(7–8), 1870–1877.
+https://doi.org/10.1080/21645515.2019.1578596
 
-20.	https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7484398/
+20.
+MacLaughlin, K.
+L., Jacobson, R.
+M., St.
+Sauver, J.
+L., Jacobson, D.
+J., Fan, C., Wi, C.-I., & Finney Rutten, L.
+J.
+(2020).
+An Innovative Housing-Related Measure for Individual Socioeconomic Status and Human Papillomavirus Vaccination Coverage: A Population-Based Cross-Sectional Study.
+Vaccine, 38(39), 6112–6119.
+https://doi.org/10.1016/j.vaccine.2020.07.026
 
-21.	https://www.cdc.gov/nchs/products/databriefs/db495.htm#:~:text=degree%20or%20higher.-,The%20percentage%20of%20children%20who%20have%20received%20one%20or%20more,of%20the%20federal%20poverty%20level.
+21.
+Villarroel, M.
+A., Galinsky, A.
+M., Lu, P.-J., & Pingali, C.
+(2024, February 13).
+Products—Data Briefs—Number 495—February 2024.
+Center for Disease Control and Prevention.
+https://doi.org/10.15620/cdc:145593
 
-22.	https://www.thelancet.com/journals/lanam/article/PIIS2667-193X(24)00021-8/fulltext
+22.
+Sonawane, K., Zhu, Y., Damgacioglu, H., Garg, A., Graboyes, E.
+M., Montealegre, J.
+R., Brownstein, N.
+C., Ford, M.
+E., Roberts, J.
+R., Sterba, K.
+R., Giuliano, A.
+R., & Deshmukh, A.
+A.
+(2024).
+Factors associated with parental human papillomavirus vaccination intentions among adolescents from socioeconomically advantaged versus deprived households: A nationwide, cross-sectional survey.
+The Lancet Regional Health – Americas, 31.
+https://doi.org/10.1016/j.lana.2024.100694
 
-23.	https://www.tandfonline.com/doi/full/10.1080/21645515.2024.2313249#:~:text=With%20regard%20to%20race%2Fethnicity,likely%20to%20have%20been%20vaccinated.
+23.
+Rincon, N.
+L., McDowell, K.
+R., Weatherspoon, D., Ritchwood, T.
+D., Rocke, D.
+J., Adjei Boakye, E., & Osazuwa-Peters, N.
+(2024).
+Racial and ethnic disparities in human papillomavirus (HPV) vaccine uptake among United States adults, aged 27–45 years.
+Human Vaccines & Immunotherapeutics, 20(1), 2313249.
+https://doi.org/10.1080/21645515.2024.2313249.
 
-24.	https://pubmed.ncbi.nlm.nih.gov/29347831/
+24.
+Agénor, M., Pérez, A.
+E., Peitzmeier, S.
+M., & Borrero, S.
+(2020).
+Racial/ethnic disparities in human papillomavirus vaccination initiation and completion among U.S.
+women in the post-Affordable Care Act era.
+Ethnicity & Health, 25(3), 393–407.
+https://doi.org/10.1080/13557858.2018.1427703
 
-25.	https://www.sciencedirect.com/science/article/abs/pii/S0091743519301148
+25.
+Spencer, J.
+C., Calo, W.
+A., & Brewer, N.
+T.
+(2019).
+Disparities and reverse disparities in HPV vaccination: A systematic review and meta-analysis.
+Preventive Medicine, 123, 197–203.
+https://doi.org/10.1016/j.ypmed.2019.03.037

### 2024-05-03 products/manuscript/supplement/Supplementary-Material.qmd
Final edits and cleaning

---
title: "Supplementary Material"

author: rachel Robertson and Kelly Cao
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
---
```{r, echo=FALSE, message=FALSE}
# load packages
library(here)
library(dplyr)
library(knitr)
library(kableExtra)
```
The supplementary materials include tables and figures from the exploratory analysis and data analysis that were unused in the final manuscript.

![S.1. Distribution of ethnicity and race in sample populaiton of U.S. teens.](../../../results/figures/ethnicity.distribution.png)

![S.2. The distribution of facilities in which the survey was given to study sample of U.S. teens.](../../../results/figures/facility.distribution.png)
![S.3. Distribution of income poverty ratio compared to HPV vaccination status (up-to-date or not up-to-date).](../../../results/figures/income-vaccination.png)

![S.4. distribution of insurance status in study population of U.S. teenagers.](../../../results/figures/insurance.status.distribution.png)
![S.5. The distribution of race/ ethnicity in U.S. teens in the study population, stratified by HPV vaccination status.](../../../results/figures/race-ethnicity-vaccination.png)
![S.6. The distribution of race/ ethnicity in U.S. teens in the study population.](../../../results/figures/race.distribution.png)

![S.7. The percentage of surveys submitted for each state in the U.S. study.](../../../results/figures/state_distribution.png)

![S.8. The distribution of HPV vaccination status in the overall study population of U.S. teens.](../../../results/figures/vaccination.status.distribution.png)

![S.9. The distribution of race/ ethnicity in U.S. teens in the study population, stratified by HPV vaccination status.](../../../results/figures/race-ethnicity-vaccination.png)

```{r, warning= FALSE, message = FALSE, echo = FALSE}
#read the table
table1 <- here("results/tables/state-vaccine-completion.txt")
S10 <- read.table(table1, header = TRUE)

# Render the table using kable
kable(S10, caption = "S.10. Percentage of up-to-date HPV vaccination status for each state from study sample.")
```

```{r, warning= FALSE, message = FALSE, echo = FALSE}
#read the table
table2 <- here("results/tables/vaccine_percent_by_race.txt")
S11 <- read.table(table2, header = TRUE)

# Render the table using kable
kable(S11, caption = "S.11. Percentage and Count of up-to-date HPV vaccination status according to racial and ethnic group.")
```

### 2024-05-03 code/analysis-code/rf-tree.R

@@ -1,180 +1,0 @@
-
-## ---- packages --------
-
-library(tidymodels) # use tidymodels framework.
-library(ggplot2) # producing visual displays of data
-library(dplyr) # manipulating and cleaning data
-library(here) # making relative pathways
-library(randomForest) # making random forest model
-library(doParallel) # for parallel processing
-library(rsample) # for cross validation
-library(yardstick)
-
-## ---- load-data --------
-
-# Load and preprocess data
-data_location <- here::here("data","processed-data","cleandata1.rds")
-mydata <- readRDS(data_location)
-
-# Remove rows with missing values
-mydata$SEX <- droplevels(mydata$SEX, exclude = c("DON'T KNOW", "MISSING IN ERROR", "REFUSED"))
-mydata$INS_STAT2_I <- droplevels(mydata$INS_STAT2_I, exclude = "MISSING Data")
-mydata$STATE <- droplevels(mydata$STATE, exclude = "Missing Data")
-mydata$MOBIL_1 <- droplevels(mydata$MOBIL_1, exclude = c("DON'T KNOW", "MISSING IN ERROR", "REFUSED"))
-mydata$FACILTY <- droplevels(mydata$FACILITY, exclude = "Missing Data")
-mydata$P_UTDHPV <- droplevels(mydata$P_UTDHPV, exclude = "Missing Data")
-
-## ---- split-data --------
-# Split data into training and testing datasets
-set.seed(123) # seed for reproducibility
-split_data <- initial_split(mydata, prop = 0.8) # 80% split for training/testing data
-train_data <- training(split_data)
-test_data <- testing(split_data)
-
-## ---- first-fit --------
-rf_rec <- recipe(P_UTDHPV ~ AGE + SEX + STATE + INS_STAT2_I + INCQ298A + INS_BREAK_I + INCPOV1 + RACEETHK + EDUC1 + LANGUAGE + MOBIL_1 + RENT_OWN + FACILITY, data = train_data) # use full model as recipe for the random forest model
-
-rf_model <- rand_forest()%>% # use rand_forest() to make a random forest model
-  set_engine("randomForest", seed = 123)%>%
-  set_mode("classification")
-
-rf_workflow <- workflow() %>% # create workflow for rf model
-  add_recipe(rf_rec)%>% # apply recipe
-  add_model(rf_model) # apply model
-
-rf_fit <- rf_workflow%>%
-  fit(data = train_data)%>% # use the workflow to fit the rf model to the data
[…]
